---
title: "dsVert validation — data preparation (seed-documented)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{dsVert validation — data preparation (seed-documented)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: |
  Reproducible recipe for the vertically-partitioned datasets used by the
  twelve method-validation probes (`vert_validation_evidence`). Every split
  is deterministic — the same `set.seed()` value rebuilds every server's
  table identically. Datasets that produced sub-noise verdicts in PHASE4
  are reused unchanged. GLMM / LMM / GEE longitudinal use a synthetic
  K=10 obs/cluster × 200 clusters generator with σ_b² = 0.5 and p̄ ≈ 0.3
  — chosen to avoid the rare-event PQL bias regime that Breslow & Clayton
  (1993) §3 documents for binomial Laplace approximations.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  cache = TRUE, cache.path = "data-preparation_cache/",
  fig.path = "data-preparation_figures/")
```

# Overview

This vignette is the canonical recipe for every dataset the
`vert_validation_evidence` probes consume. Each builder is a pure function
of its `seed`; running it locally on this machine rebuilds the same tables
that live on the validation Opal cluster (`vert_demo` project). The
provisioning script `scripts/opal_provision_full.R` invokes these
builders, uploads each split with `id.name="id"` (a synthetic row index)
so `patient_id` / `row_id` remain queryable variables on the server, and
grants `view` permission to the runner account.

| Split | Methods | Rows | Servers | Seed |
|---|---|---|---|---|
| `bw_hospital_a` / `bw_hospital_b` | glm K=2 | 189 → 165 PSI | s1 / s2 | 123 |
| `na_data_s1` / `na_data_s2` | nb, ord_joint, mnl_joint, ordinal_warm K=2, multinom_warm K=2, lasso K=2 | 392 → 170 PSI | s1 / s2 | 123 + 124 (bp tertile) |
| `lung_s1` / `lung_s2` | cox K=2 | 167 → 150 PSI | s1 / s2 | 125 |
| `long_glmm_s1` / `long_glmm_s2` | glmm K=2 (synthetic, p̄≈0.3) | 2000 / 27 clusters | s1 / s2 | 200 |
| `long_lmm_s1` / `long_lmm_s2` | lmm K=2, gee K=2 (synthetic Gaussian) | 2000 / 200 clusters | s1 / s2 | 201 |
| `cgd_ipw_s1` / `cgd_ipw_s2` | ipw K=2 (CGD enum=1, propensity-weighted) | 128 → 128 PSI | s1 / s2 | 128 + 129 |
| `pima_server1..3` | glm K=3, lasso K=3 | 768 → 392 PSI | s1 / s2 / s3 | 123 |
| `pima_warm_s1..3` | multinom_warm K=3, ordinal_warm K=3 | (Pima K=3 + bp tertile) | s1 / s2 / s3 | 123 + 124 |

Every public-domain dataset is from R recommended packages
(`MASS::birthwt`, `survival::lung`, `survival::cgd`, `nlme::Orthodont`,
`MASS::Pima.tr` + `MASS::Pima.te`).

```{r libs}
suppressPackageStartupMessages({
  library(MASS)
  library(survival)
  library(nlme)
})
```

# K = 2 splits

## birthwt (`bw_hospital_a` / `bw_hospital_b`)

Used by the GLM K=2 probe. PHASE4 baseline.

```{r birthwt}
build_birthwt_k2 <- function(seed = 123L) {
  data(birthwt, package = "MASS")
  bdf <- birthwt
  bdf$patient_id <- sprintf("BW%03d", seq_len(nrow(bdf)))
  set.seed(seed)
  core <- sample(bdf$patient_id, 160)
  rem  <- setdiff(bdf$patient_id, core)
  ids1 <- sample(c(core, sample(rem, 15)))
  ids2 <- sample(c(core, sample(rem, 14)))
  list(
    bw_hospital_a = bdf[bdf$patient_id %in% ids1,
                        c("patient_id", "age", "lwt", "race", "smoke")],
    bw_hospital_b = bdf[bdf$patient_id %in% ids2,
                        c("patient_id", "ptl", "ht", "ui", "ftv",
                          "bwt", "low")])
}
str(build_birthwt_k2())
```

## NHANES Pima K=2 (`na_data_s1` / `na_data_s2`)

Used by NB, ord_joint, mnl_joint, ordinal_warm K=2, multinom_warm K=2,
lasso K=2. The `bp_cls` tertile factor + one-hot indicator triplet +
cumulative `_leq` columns are added to s2 so the ordinal/multinomial
joint Newton paths can read them as `cumulative_template` / `indicator_template`.

```{r pima-k2}
build_pima_k2 <- function(seed = 123L, bp_seed = 124L) {
  pdf <- rbind(MASS::Pima.tr, MASS::Pima.te)
  pdf$diabetes <- as.integer(pdf$type == "Yes")  # MASS uses `type` factor
  pdf$patient_id <- sprintf("P%03d", seq_len(nrow(pdf)))
  set.seed(seed)
  core <- sample(pdf$patient_id, 150)
  rem  <- setdiff(pdf$patient_id, core)
  ids1 <- sample(c(core, sample(rem, 25)))
  ids2 <- sample(c(core, sample(rem, 20)))
  s1 <- pdf[pdf$patient_id %in% ids1,
             c("patient_id", "age", "bmi", "ped", "npreg", "skin")]
  s2_base <- pdf[pdf$patient_id %in% ids2,
                  c("patient_id", "glu", "bp", "diabetes")]
  ## bp tertile factor + indicators (deterministic seed bp_seed for
  ## stable tertile cutpoints across reproducible runs).
  set.seed(bp_seed)
  brks <- quantile(s2_base$bp, c(0, 1/3, 2/3, 1), na.rm = TRUE)
  s2_base$bp_cls <- factor(
    cut(s2_base$bp, breaks = brks, include.lowest = TRUE,
        labels = c("low", "med", "high")),
    levels = c("low", "med", "high"), ordered = TRUE)
  s2_base$low_ind  <- as.integer(s2_base$bp_cls == "low")
  s2_base$med_ind  <- as.integer(s2_base$bp_cls == "med")
  s2_base$high_ind <- as.integer(s2_base$bp_cls == "high")
  s2_base$low_leq  <- as.integer(as.integer(s2_base$bp_cls) <= 1L)
  s2_base$med_leq  <- as.integer(as.integer(s2_base$bp_cls) <= 2L)
  s2_base$high_leq <- 1L  # always TRUE for the top tertile
  list(na_data_s1 = s1, na_data_s2 = s2_base)
}
str(build_pima_k2())
```

## survival::lung K=2 (`lung_s1` / `lung_s2`)

Used by Cox K=2.

```{r lung-k2}
build_lung_k2 <- function(seed = 125L) {
  data(lung, package = "survival")
  lng <- get("lung")
  lng$patient_id <- sprintf("LNG%03d", seq_len(nrow(lng)))
  keep <- complete.cases(lng[, c("age", "sex", "ph.ecog", "ph.karno",
                                  "time", "status")])
  lng <- lng[keep, ]
  set.seed(seed)
  core <- sample(lng$patient_id, 150)
  rem  <- setdiff(lng$patient_id, core)
  ids1 <- sample(c(core, sample(rem, 20)))
  ids2 <- sample(c(core, sample(rem, 15)))
  s1 <- lng[lng$patient_id %in% ids1,
             c("patient_id", "age", "sex", "ph.ecog", "ph.karno")]
  s2 <- lng[lng$patient_id %in% ids2,
             c("patient_id", "time", "status", "meal.cal", "wt.loss")]
  s2$status <- as.integer(s2$status == 2)  # event = 1, censored = 0
  list(lung_s1 = s1, lung_s2 = s2)
}
str(build_lung_k2())
```

## CGD IPW K=2 (`cgd_ipw_s1` / `cgd_ipw_s2`)

Used by IPW. CGD `enum == 1` → one row per subject. Stage-1 propensity
glm fit centrally to pre-compute the `ipw` column on s2 (validation-only
re-population convention; production pipelines compute weights via
federated stage-1 + `ds.assign`).

```{r cgd-ipw-k2}
build_cgd_ipw_k2 <- function(seed_s1 = 128L, seed_s2 = 129L) {
  data(cgd, package = "survival")
  cg <- cgd[cgd$enum == 1L, ]
  cg$patient_id   <- sprintf("CGD%03d", cg$id)
  cg$sex_num      <- as.integer(cg$sex == "male")
  cg$treat_num    <- as.integer(cg$treat == "rIFN-g")
  n_inf <- tapply(cgd$status, cgd$id, sum)
  cg$n_infections <- as.integer(n_inf[as.character(cg$id)])
  pooled <- cg[, c("patient_id", "age", "sex_num", "height", "weight",
                   "treat_num", "status", "n_infections")]
  pooled <- pooled[complete.cases(pooled), ]
  prop <- glm(treat_num ~ age + sex_num + height + weight,
              data = pooled, family = binomial())
  phat <- fitted(prop)
  pooled$ipw <- ifelse(pooled$treat_num == 1L, 1/phat, 1/(1 - phat))
  set.seed(seed_s1); ids1 <- sample(pooled$patient_id)
  set.seed(seed_s2); ids2 <- sample(pooled$patient_id)
  s1 <- pooled[match(ids1, pooled$patient_id),
                c("patient_id", "age", "sex_num", "height", "weight")]
  s2 <- pooled[match(ids2, pooled$patient_id),
                c("patient_id", "treat_num", "status",
                  "n_infections", "ipw")]
  list(cgd_ipw_s1 = s1, cgd_ipw_s2 = s2,
       propensity_coef_central = coef(prop))
}
str(build_cgd_ipw_k2())
```

# Synthetic longitudinal (`long_glmm_*`, `long_lmm_*`)

GLMM, LMM, and GEE all benefit from a balanced longitudinal design where
the cluster random effect is identifiable. We generate K=10 observations
per cluster across n_clusters = 200 (n = 2 000 total). The intercept is
tuned so the binomial outcome stays in the well-identified regime
p̄ ≈ 0.3 — far from the 0.5-asymptote that the simplified Gaussian-style
BLUP shrinkage in `ds.vertGLMM` was approximating prior to the offset
session-state fix, and equally far from the rare-event tail
(p ≪ 0.1) where Breslow & Clayton (1993) §3 documents the PQL Laplace
approximation becoming severely biased.

```{r long-synth}
build_long_glmm_k2 <- function(seed = 200L,
                                n_clusters = 200L, n_obs = 10L,
                                sigma_b = sqrt(0.5),
                                beta = c(intercept = -0.85,
                                         x1 = 0.5, x2 = 0.3)) {
  set.seed(seed)
  cluster_id  <- rep(sprintf("C%04d", seq_len(n_clusters)), each = n_obs)
  obs_id      <- sprintf("O%05d", seq_along(cluster_id))
  b_i         <- rnorm(n_clusters, 0, sigma_b)
  b           <- rep(b_i, each = n_obs)
  x1          <- rnorm(length(cluster_id))
  x2          <- rbinom(length(cluster_id), 1, 0.5)
  eta         <- beta["intercept"] + beta["x1"] * x1 + beta["x2"] * x2 + b
  p           <- plogis(eta)
  y           <- rbinom(length(cluster_id), 1, p)
  list(long_glmm_s1 = data.frame(patient_id = obs_id,
                                   x1 = x1,
                                   anchor = 1L),
       long_glmm_s2 = data.frame(patient_id = obs_id,
                                   cluster_id = cluster_id,
                                   x2 = x2, y = y),
       diagnostics = list(p_bar = mean(p),
                          y_rate = mean(y),
                          sigma_b2 = sigma_b^2))
}

build_long_lmm_k2 <- function(seed = 201L,
                               n_clusters = 200L, n_obs = 10L,
                               sigma_b = sqrt(0.5), sigma_e = 1,
                               beta = c(intercept = 5,
                                        x1 = 0.5, x2 = 0.3)) {
  set.seed(seed)
  cluster_id <- rep(sprintf("C%04d", seq_len(n_clusters)), each = n_obs)
  obs_id     <- sprintf("O%05d", seq_along(cluster_id))
  b          <- rep(rnorm(n_clusters, 0, sigma_b), each = n_obs)
  x1         <- rnorm(length(cluster_id))
  x2         <- rbinom(length(cluster_id), 1, 0.5)
  y          <- beta["intercept"] + beta["x1"] * x1 +
                beta["x2"] * x2 + b +
                rnorm(length(cluster_id), 0, sigma_e)
  list(long_lmm_s1 = data.frame(patient_id = obs_id,
                                  x1 = x1, anchor = 1L),
       long_lmm_s2 = data.frame(patient_id = obs_id,
                                  cluster_id = cluster_id,
                                  x2 = x2, y = y))
}

g <- build_long_glmm_k2()
cat(sprintf("[GLMM synth] p̄=%.3f y_rate=%.3f σ_b²=%.3f n=%d\n",
            g$diagnostics$p_bar, g$diagnostics$y_rate,
            g$diagnostics$sigma_b2,
            nrow(g$long_glmm_s2)))
str(build_long_lmm_k2(), max.level = 1)
```

# K = 3 splits

GLM and the warm-path multinomial / ordinal probes have a true K=3
implementation (`ds.vertGLM.k3ring63.R` + OVR over GLM cells). Cox,
GLMM, LMM, GEE, NB full_reg_nd, IPW, and the joint-Newton multinomial /
ordinal Hessian solvers are K=2-only by construction (closed-form
sufficient statistics that depend on a 2-party additive split).
The K=3 probes therefore exercise glm, multinom_warm, ordinal_warm,
and lasso (proximal-on-GLM-K=3).

## Pima K=3 (`pima_server1..3`)

```{r pima-k3}
build_pima_k3 <- function(seed = 123L) {
  pdf <- rbind(MASS::Pima.tr, MASS::Pima.te)
  pdf$diabetes <- as.integer(pdf$type == "Yes")
  pdf$patient_id <- sprintf("P%03d", seq_len(nrow(pdf)))
  set.seed(seed)
  core <- sample(pdf$patient_id, 150)
  rem  <- setdiff(pdf$patient_id, core)
  ids1 <- sample(c(core, sample(rem, 30)))
  ids2 <- sample(c(core, sample(rem, 25)))
  ids3 <- sample(c(core, sample(rem, 20)))
  s1 <- pdf[pdf$patient_id %in% ids1,
             c("patient_id", "age", "bmi")]
  s2 <- pdf[pdf$patient_id %in% ids2,
             c("patient_id", "ped", "glu")]
  s3 <- pdf[pdf$patient_id %in% ids3,
             c("patient_id", "bp", "skin", "npreg", "diabetes")]
  list(pima_server1 = s1, pima_server2 = s2, pima_server3 = s3)
}
str(build_pima_k3())
```

## Pima warm K=3 (`pima_warm_s1..3`)

For multinom_warm and ordinal_warm K=3 we add the bp-tertile factor +
indicator/cumulative-leq columns on s3 (the outcome server) using the
same seed convention as the K=2 split.

```{r pima-warm-k3}
build_pima_warm_k3 <- function(seed = 123L, bp_seed = 124L) {
  k3 <- build_pima_k3(seed)
  set.seed(bp_seed)
  brks <- quantile(k3$pima_server3$bp, c(0, 1/3, 2/3, 1), na.rm = TRUE)
  s3 <- k3$pima_server3
  s3$bp_cls <- factor(
    cut(s3$bp, breaks = brks, include.lowest = TRUE,
        labels = c("low", "med", "high")),
    levels = c("low", "med", "high"), ordered = TRUE)
  s3$low_ind  <- as.integer(s3$bp_cls == "low")
  s3$med_ind  <- as.integer(s3$bp_cls == "med")
  s3$high_ind <- as.integer(s3$bp_cls == "high")
  s3$low_leq  <- as.integer(as.integer(s3$bp_cls) <= 1L)
  s3$med_leq  <- as.integer(as.integer(s3$bp_cls) <= 2L)
  s3$high_leq <- 1L
  list(pima_warm_s1 = k3$pima_server1,
       pima_warm_s2 = k3$pima_server2,
       pima_warm_s3 = s3)
}
str(build_pima_warm_k3())
```

# References

* Catrina O., Saxena A. (2010). *Secure Computation with Fixed-Point
  Numbers*, Financial Cryptography. §3.3 ULP propagation under
  multiplicative depth — anchors the deterministic
  fixed-point bounds the probes report.
* Demmler D., Schneider T., Zohner M. (2015). *ABY: A Framework for
  Efficient Mixed-Protocol Secure Two-Party Computation*, NDSS §III.B —
  K=2 OT-Beaver dishonest-majority threat model.
* Breslow N. E., Clayton D. G. (1993). *Approximate Inference in
  Generalized Linear Mixed Models*, JASA 88(421) §3 — Laplace /
  PQL approximation bias regimes for binomial GLMM, motivates the
  synthetic p̄ ≈ 0.3 design choice for the GLMM probe.
* McCullagh P. (1980). *Regression Models for Ordinal Data*, JRSS-B
  42(2) §2.5 — closed-form Hessian for the proportional-odds joint
  Newton step.
* Krishnapuram B., Carin L., Figueiredo M., Hartemink A. (2005).
  *Sparse Multinomial Logistic Regression*, IEEE PAMI 27(6) §3.2 —
  the OVR + softmax-anchor warm path.
* Wellek S. (2010), Schuirmann D. J. (1987), Lakens D. (2017) —
  Wald-SE sub-noise comparator framework.
* Pratt J. W. (1981) *JASA* 76(373); Christensen R. H. B. (2019)
  `ordinal::clm.fit` §A.3 — finite-sample MLE noise floor framework.
