---
title: "CorrelationDemo"
output: html_document
date: "2024-05-06"
---


## Vertically federated correlation analysis + PCA

# Using ds.vertCor and vertCorDS

I want to show in this markdown file how to use vertCorDS and ds.vertCor to find correlation between numeric variables. Also extended the SVD results to demonstrate how to do Principal Component Analysis.

```{r}
vertCorDS <- function(data_name, variable_names = NULL) {

  U_combined <- NULL

  for(var_name in variable_names) {
    result <- data_name[[var_name]]
    result <- scale(result)
    svd_result <- svd(result)
    U_combined <- cbind(U_combined, svd_result$u %*% svd_result$d)
  }

  return(U_combined)
}


ds.vertCor <- function(data_names, variable_names, datasources = NULL) {
  # standard, if NULL look for connections
  if(is.null(datasources)){
    datasources <- datashield.connections_find()
  }

  U_combined <- NULL

  study_names  <- names(datasources)
  for(i in 1:length(study_names)) {
    temp <-variable_names[[i]]
    assign("temp", temp, envir = .GlobalEnv)
    output <- datashield.aggregate(datasources[i], quote(vertCorDS(D, temp)))[[1]]
    U_combined <- cbind(U_combined, output)
  }


  # Perform SVD on the combined U matrix
  svd_final <- svd(U_combined)
  # Calculate and scale the correlation matrix
  correlation <- svd_final$v %*% diag(svd_final$d^2) %*% t(svd_final$v)
  correlation <- correlation / correlation[1, 1]
  return(correlation)
}


```

```{r}
library(DSLite)
library(resourcer)
library(dsBase)
library(dsBaseClient)
library(dsVert)
library(dsVertClient)
```

Setting up the dataframes to allow the conditions where we would want to apply our code 

```{r}
data("CNSIM1")
data("CNSIM2")
data("CNSIM3")

AllCNSIM <- rbind(CNSIM1,CNSIM2,CNSIM3)
AllCNSIM$id <- 1:9379

AllCNSIM <- na.omit(AllCNSIM)
AllCNSIM$DIS_AMI <- as.numeric(AllCNSIM$DIS_AMI)
AllCNSIM$PM_BMI_CATEGORICAL <- as.numeric(AllCNSIM$PM_BMI_CATEGORICAL)
Partition1 <- AllCNSIM[,c("id", "LAB_TSC","LAB_TRIG","PM_BMI_CONTINUOUS","DIS_CVA")]
Partition2 <- AllCNSIM[,c("id", "LAB_HDL","LAB_GLUC_ADJUSTED","MEDI_LPD","DIS_DIAB")]
Partition3 <- AllCNSIM[,c("id", "DIS_AMI","GENDER","PM_BMI_CATEGORICAL")]

data_names = list(server1 = D, server2 = D, server3 = D)
var_names = list(server1 = c("LAB_TSC","LAB_TRIG","PM_BMI_CONTINUOUS"), server2 = c( "LAB_HDL","LAB_GLUC_ADJUSTED"), server3 =  c("DIS_AMI", "PM_BMI_CATEGORICAL"))
```

Setting uo the DSLite environment

```{r}
dslite.server <- newDSLiteServer(
    config = DSLite::defaultDSConfiguration(include=c("dsBase", "resourcer", "dsVert")), 
    tables = list(VertCNSIM1 = Partition1, VertCNSIM2 = Partition2, VertCNSIM3 = Partition3)
)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1", url = "dslite.server", table = "VertCNSIM1", driver = "DSLiteDriver")
builder$append(server = "server2", url = "dslite.server", table = "VertCNSIM2", driver = "DSLiteDriver")
builder$append(server = "server3", url = "dslite.server", table = "VertCNSIM3", driver = "DSLiteDriver")
logindata.dslite <- builder$build()
conns <- datashield.login(logindata.dslite, assign=T)

dslite.server$aggregateMethod("ds.vertCor", function(data_names, variable_names, datasources){ ds.vertCor(data_names, variable_names, datasources) })
dslite.server$aggregateMethod("vertCorDS", function(data_name, variable_names){ vertCorDS(data_name, variable_names) })
```

Get correlation and compare with cor() function

```{r}
ds.vertCor(data_names, var_names, conns)

cor(AllCNSIM[,c("LAB_TSC","LAB_TRIG","PM_BMI_CONTINUOUS","LAB_HDL","LAB_GLUC_ADJUSTED","DIS_AMI", "PM_BMI_CATEGORICAL")],use = "complete.obs")
```

Quick demonstration to show how block svd results can be used for PCA and some potential implementations for dsVert and dsVertClient.

```{r}
U_combined <- NULL
for(i in 1:3) {
    temp <-var_names[[i]]
    assign("temp", temp, envir = .GlobalEnv)
    output <- datashield.aggregate(conns[i], quote(vertCorDS(D, temp)))[[1]]
    U_combined <- cbind(U_combined, output)
  }

#head(U_combined)
svd_result <- svd(U_combined)
U <- svd_result$u
D <- diag(svd_result$d)
V <- svd_result$v
principal_components <- U %*% D
print(head(principal_components))
variance <- (svd_result$d^2) / sum(svd_result$d^2)
print(" ")
print(variance)
```

```{r}
plot(principal_components[, 1], principal_components[, 2], xlab = "PC1", ylab = "PC2", main = "PCA Plot")
```



