---
title: "Data Preparation: Creating the Validation Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation: Creating the Validation Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This document shows how the validation datasets on `opal-demo.obiba.org` were
created. Both datasets come from standard R packages (MASS, survival), so any
researcher can reproduce the exact same partitions using the seed below.

## birthwt (K=2)

The birthwt dataset (189 observations) is split vertically into two hospitals
with deliberately non-identical patient populations — simulating a realistic
scenario where institutions have overlapping but not identical patient registries.

```{r birthwt}
library(MASS)
data(birthwt)

df <- birthwt
df$patient_id <- sprintf("BW%03d", seq_len(nrow(df)))

set.seed(123)
core <- sample(df$patient_id, 160)
remaining <- setdiff(df$patient_id, core)
ids1 <- sample(c(core, sample(remaining, 15)))  # 175 patients
ids2 <- sample(c(core, sample(remaining, 10)))  # 170 patients

# Hospital A: maternal demographics
hospital_a <- df[df$patient_id %in% ids1, c("patient_id", "age", "lwt", "race", "smoke")]

# Hospital B: clinical history + outcomes
hospital_b <- df[df$patient_id %in% ids2, c("patient_id", "ptl", "ht", "ui", "ftv", "bwt", "low")]

# Shuffle row order independently (PSI must handle this)
hospital_a <- hospital_a[sample(nrow(hospital_a)), ]
hospital_b <- hospital_b[sample(nrow(hospital_b)), ]
```

## CGD (K=3)

The CGD dataset (128 patients from a clinical trial of interferon gamma for
chronic granulomatous disease) is split vertically across three institutions.
All 128 patients are present at all three sites, but row order is shuffled
independently.

```{r cgd}
library(survival)
data(cgd)

d <- cgd[cgd$enum == 1, ]
d$n_infections <- as.integer(tapply(cgd$status, cgd$id, sum)[as.character(d$id)])
d$patient_id <- sprintf("CGD%03d", seq_len(nrow(d)))
d$treat_num <- as.integer(d$treat == "rIFN-g")
d$sex_num <- as.integer(d$sex == "male")

# Demographics
demographics <- d[, c("patient_id", "age", "sex_num", "height", "weight")]

# Treatment assignment
treatment <- d[, c("patient_id", "treat_num", "steroids")]

# Outcomes
outcomes <- d[, c("patient_id", "status", "n_infections")]

set.seed(42); demographics <- demographics[sample(nrow(demographics)), ]
set.seed(43); treatment    <- treatment[sample(nrow(treatment)), ]
set.seed(44); outcomes     <- outcomes[sample(nrow(outcomes)), ]
```

## Upload to Opal

```{r upload}
library(opalr)

o <- opal.login("administrator", "password", url = "https://opal-demo.obiba.org")

# birthwt
hospital_a$id <- seq_len(nrow(hospital_a))
hospital_b$id <- seq_len(nrow(hospital_b))
opal.table_save(o, hospital_a, "vert_demo", "bw_hospital_a", overwrite = TRUE, force = TRUE)
opal.table_save(o, hospital_b, "vert_demo", "bw_hospital_b", overwrite = TRUE, force = TRUE)

# CGD
demographics$id <- seq_len(nrow(demographics))
treatment$id    <- seq_len(nrow(treatment))
outcomes$id     <- seq_len(nrow(outcomes))
opal.table_save(o, demographics, "vert_demo", "cgd_demographics", overwrite = TRUE, force = TRUE)
opal.table_save(o, treatment,    "vert_demo", "cgd_treatment",    overwrite = TRUE, force = TRUE)
opal.table_save(o, outcomes,     "vert_demo", "cgd_outcomes",     overwrite = TRUE, force = TRUE)

opal.logout(o)
```
