---
title: "dsVert: Privacy-Preserving Vertical GLM, Correlation & PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsVert Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

**dsVert** enables privacy-preserving statistical analysis on vertically
partitioned data across multiple DataSHIELD servers. Each server holds
different features for the same observations. No server sees another's
raw data. The analyst (client) receives only aggregate statistics.

All computation uses **Ring63 Beaver MPC** — pure fixed-point arithmetic
with additive secret sharing. No observation-level
data is ever disclosed.

### Supported analyses

| Function | Description |
|----------|-------------|
| `ds.psiAlign()` | Private Set Intersection (EC-DH blind relay) |
| `ds.vertGLM()` | GLM: Gaussian, Binomial, Poisson (L-BFGS + DCF wide spline) |
| `ds.vertCor()` | Pearson correlation matrix (Ring63 Beaver cross-products) |
| `ds.vertPCA()` | Principal Component Analysis (eigen decomposition of correlation) |

## Setup

```{r connect, eval=FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

# Connect to 3 Opal servers (each at a different institution)
builder <- DSI::newDSLoginBuilder()
builder$append(server = "hospital_A", url = "https://opal-a:443",
               user = "analyst", password = "***",
               table = "study.patient_data")
builder$append(server = "hospital_B", url = "https://opal-b:443",
               user = "analyst", password = "***",
               table = "study.patient_data")
builder$append(server = "hospital_C", url = "https://opal-c:443",
               user = "analyst", password = "***",
               table = "study.patient_data")

conns <- DSI::datashield.login(logins = builder$build())
DSI::datashield.assign.table(conns, "D", "study.patient_data")
```

## Step 1: Record Alignment (PSI)

Each server may have different patients. PSI finds the common set
without revealing which patients are on which server.

```{r psi, eval=FALSE}
psi <- ds.psiAlign(
  data_name   = "D",
  id_col      = "patient_id",
  newobj       = "D_aligned",
  datasources = conns
)
# PSI result: 158 common observations across 3 servers
```

## Step 2: Binomial GLM (Logistic Regression)

Fit a logistic regression for diabetes prediction. Features are
distributed across servers — no server sees the full feature set.

```{r glm-binomial, eval=FALSE}
result <- ds.vertGLM(
  data_name = "D_aligned",
  y_var     = "diabetes",
  x_vars    = list(
    hospital_A = c("age", "bmi"),
    hospital_B = c("ped", "glu"),
    hospital_C = c("bp", "skin")
  ),
  y_server  = "hospital_C",
  family    = "binomial",
  datasources = conns
)
```

### Output

```{r glm-output, eval=FALSE}
print(result)
# Coefficients:
#                 Estimate Std.Error z value  Pr(>|z|)
# (Intercept)     -9.2784    2.0967  -4.425  0.000010 ***
# age              0.0733    0.0222   3.301  0.000965 ***
# bmi              0.0798    0.0520   1.534  0.124911
# ped              1.4823    0.7913   1.873  0.061040
# glu              0.0290    0.0101   2.878  0.004007 **
# bp              -0.0191    0.0205  -0.932  0.351094
# skin             0.0089    0.0202   0.442  0.658706
#
# Converged: TRUE (12 iterations)
# Deviance: 22.09
```

The output matches a centralized `glm()` to ~0.001 precision.
Standard errors are computed via finite-difference Hessian — no
individual observation data is disclosed.

## Step 3: Gaussian GLM (Linear Regression)

```{r glm-gaussian, eval=FALSE}
result_lm <- ds.vertGLM(
  data_name = "D_aligned",
  y_var     = "npreg",
  x_vars    = list(
    hospital_A = c("age", "bmi"),
    hospital_B = c("ped", "glu"),
    hospital_C = c("bp", "skin")
  ),
  y_server  = "hospital_C",
  family    = "gaussian",
  datasources = conns
)
# Converges in 11 iterations (~30s for Gaussian identity link)
```

## Step 4: Correlation Matrix

Compute the full cross-server correlation matrix. Each correlation
coefficient is a single aggregate scalar — no individual data disclosed.

```{r cor, eval=FALSE}
cor_result <- ds.vertCor(
  data_name  = "D_aligned",
  variables  = list(
    hospital_A = c("age", "bmi"),
    hospital_B = c("ped", "glu"),
    hospital_C = c("bp", "skin")
  ),
  datasources = conns
)
print(round(cor_result$correlation, 3))
#        age   bmi    ped   glu     bp  skin
# age  1.000 0.189  0.425 0.198 -0.113 0.283
# bmi  0.189 1.000  0.321 0.657  0.082 0.237
# ped  0.425 0.321  1.000 0.068 -0.079 0.366
# glu  0.198 0.657  0.068 1.000  0.153 0.237
# bp  -0.113 0.082 -0.079 0.153  1.000 0.257
# skin 0.283 0.237  0.366 0.237  0.257 1.000
```

## Step 5: PCA

```{r pca, eval=FALSE}
pca_result <- ds.vertPCA(
  data_name  = "D_aligned",
  variables  = list(
    hospital_A = c("age", "bmi"),
    hospital_B = c("ped", "glu"),
    hospital_C = c("bp", "skin")
  ),
  datasources = conns
)
# 6 components, PC1 explains 37% of variance
```

## Security Model

| Property | Guarantee |
|----------|-----------|
| **Observation-level disclosure** | Zero — only p-dimensional aggregates |
| **Client sees** | Gradients (sums over n obs), correlation scalars |
| **Servers see** | Only their own data + random additive shares |
| **Beaver triples** | Generated server-side (client never sees values) |
| **Transport encryption** | X25519 + AES-256-GCM between servers |
| **Dealer rotation** | Different server generates triples each iteration |
| **Collusion threshold** | Analyst needs (K-1)/K servers for K≥3 |

## Performance

| Analysis | K=2 | K=3 |
|----------|-----|-----|
| Key setup | 0.3s | 0.5s |
| GLM Binomial (12 iters + SE) | ~330s | ~350s |
| GLM Gaussian (11 iters + SE) | ~60s | ~65s |
| GLM Poisson (11 iters + SE) | ~500s | ~540s |
| Correlation (p=6) | 16s | 18s |
| PCA (p=6) | <1s (from cor) | 18s |

```{r cleanup, eval=FALSE}
DSI::datashield.logout(conns)
```
