---
title: "K=2 Validation: Two-Party Gauss-Seidel BCD-IRLS"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{K=2 Validation: Two-Party Gauss-Seidel BCD-IRLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

cache_dir <- file.path(getwd(), "k2-mpc-validation_cache")
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)

.cache_path <- function(name) file.path(cache_dir, paste0(name, ".rds"))

.cached <- function(name, expr) {
  path <- .cache_path(name)
  if (file.exists(path)) {
    message(sprintf("[cache] Loading %s from %s", name, basename(path)))
    return(readRDS(path))
  }
  result <- expr
  saveRDS(result, path)
  message(sprintf("[cache] Saved %s to %s", name, basename(path)))
  result
}
```

## Introduction

This vignette validates the K=2 **Gauss-Seidel BCD-IRLS** (`k2_mpc`) mode
of dsVert for binomial and Poisson GLMs. This mode is an alternative to the
HE-Link (CKKS polynomial) pathway, offering near-exact coefficient recovery
at the cost of revealing the aggregated linear predictor to the coordinator.

The protocol mirrors the K$\geq$3 secure routing pattern
(`glmCoordinatorStepDS` + `glmSecureBlockSolveDS`), using X25519+AES-256-GCM
transport encryption instead of CKKS:

1. **Coordinator** (label server) updates first: decrypts non-label's
   $\boldsymbol{\eta}_\text{nl}$, runs full IRLS, recomputes $\mu$/$w$/residual
   with new $\beta$ (**Gauss-Seidel key step**), encrypts $(w, r)$ for non-label.
2. **Non-label** decrypts $(w, r)$, computes $g = X^T r$, solves IRLS block,
   encrypts new $\eta$ for coordinator.

Per iteration: 2 server calls + 2 blob relays ($\sim$2s on Opal).

### Comparison with HE-Link

| | HE-Link (`he_link`) | GS-IRLS (`k2_mpc`) |
|---|---|---|
| Accuracy | $\leq 5 \times 10^{-2}$ (polynomial approx) | $\leq 10^{-4}$ (exact link) |
| Speed | Hours (CKKS at `log_n`=14) | Seconds (~7 iters) |
| Privacy | $\eta$ stays encrypted | Coordinator sees $\eta_\text{total}$ |

### Data

We use the Pima diabetes dataset (`MASS::Pima.tr`, $n = 200$) split across two
servers: server 1 (`age`, `bmi`, `ped`) and server 3 (`npreg`, `diabetes`).

## Data preparation

```{r data-prep}
library(MASS)
data("Pima.tr", package = "MASS")
df <- as.data.frame(Pima.tr)
df$patient_id <- sprintf("P%03d", seq_len(nrow(df)))
df$diabetes   <- ifelse(df$type == "Yes", 1L, 0L)

set.seed(123)

# Same partitioning as K=3 validation (reproducible with set.seed(123))
core_ids <- sample(df$patient_id, 150)
remaining <- setdiff(df$patient_id, core_ids)

ids1 <- sample(c(core_ids, sample(remaining, 30)))  # 180 patients
ids3 <- sample(c(core_ids, sample(remaining, 20)))  # 170 patients

server1_data <- df[df$patient_id %in% ids1, c("patient_id", "age", "bmi", "ped")]
server3_data <- df[df$patient_id %in% ids3, c("patient_id", "npreg", "diabetes")]

server1_data <- server1_data[sample(nrow(server1_data)), ]
server3_data <- server3_data[sample(nrow(server3_data)), ]

cat(sprintf("Server 1: %d patients (age, bmi, ped)\n", nrow(server1_data)))
cat(sprintf("Server 3: %d patients (npreg, diabetes)\n", nrow(server3_data)))
cat(sprintf("Common patients: %d\n", length(intersect(ids1, ids3))))
```

## Connect and align

```{r connect, results = "hide", message = FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server1")
builder$append(server = "server3",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server3")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

### PSI Record Alignment

```{r psi}
aligned <- .cached("psi_result", {
  ds.psiAlign("D", id_col = "patient_id", datasources = conns)
})
cat(sprintf("Common cohort: %d patients\n", aligned$n_common))
```

## Centralized baseline

```{r baseline}
common <- intersect(ids1, ids3)
pooled <- df[df$patient_id %in% common, ]
cat(sprintf("Pooled dataset: %d observations\n", nrow(pooled)))
```

## K=2 Binomial GLM (Gauss-Seidel IRLS)

```{r binomial, message = FALSE}
binom_k2 <- .cached("binom_k2_mpc", {
  ds.vertGLM(
    data_name   = "D_aligned",
    y_var       = "diabetes",
    x_vars      = list(server1 = c("age", "ped"),
                       server3 = c("npreg")),
    y_server    = "server3",
    family      = "binomial",
    lambda      = 1e-4,
    max_iter    = 100,
    tol         = 1e-4,
    verbose     = FALSE,
    datasources = conns,
    eta_privacy = "k2_mpc"
  )
})

binom_central <- glm(diabetes ~ age + ped + npreg,
                      data = pooled, family = binomial)

cat("K=2 Binomial (Gauss-Seidel BCD-IRLS):\n")
cat(sprintf("  Converged: %s in %d iterations\n",
            binom_k2$converged, binom_k2$iterations))

coefs_k2_b <- binom_k2$coefficients
coefs_central_b <- coef(binom_central)[names(coefs_k2_b)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_b - coefs_central_b))))
```

```{r binomial-table, echo = FALSE}
comparison_b <- data.frame(
  Variable    = names(coefs_k2_b),
  dsVert_K2   = round(coefs_k2_b, 6),
  Centralized = round(coefs_central_b, 6),
  Difference  = sprintf("%.2e", coefs_k2_b - coefs_central_b),
  row.names   = NULL
)
knitr::kable(comparison_b, caption = "Binomial GLM: K=2 GS-IRLS vs centralized")
```

## K=2 Poisson GLM (Gauss-Seidel IRLS)

```{r poisson, message = FALSE}
poisson_k2 <- .cached("poisson_k2_mpc", {
  ds.vertGLM(
    data_name   = "D_aligned",
    y_var       = "npreg",
    x_vars      = list(server1 = c("age", "bmi", "ped"),
                       server3 = c("diabetes")),
    y_server    = "server3",
    family      = "poisson",
    lambda      = 1e-4,
    max_iter    = 100,
    tol         = 1e-4,
    verbose     = FALSE,
    datasources = conns,
    eta_privacy = "k2_mpc"
  )
})

poisson_central <- glm(npreg ~ age + bmi + ped + diabetes,
                        data = pooled, family = poisson)

cat("K=2 Poisson (Gauss-Seidel BCD-IRLS):\n")
cat(sprintf("  Converged: %s in %d iterations\n",
            poisson_k2$converged, poisson_k2$iterations))

coefs_k2_p <- poisson_k2$coefficients
coefs_central_p <- coef(poisson_central)[names(coefs_k2_p)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_p - coefs_central_p))))
```

```{r poisson-table, echo = FALSE}
comparison_p <- data.frame(
  Variable    = names(coefs_k2_p),
  dsVert_K2   = round(coefs_k2_p, 6),
  Centralized = round(coefs_central_p, 6),
  Difference  = sprintf("%.2e", coefs_k2_p - coefs_central_p),
  row.names   = NULL
)
knitr::kable(comparison_p, caption = "Poisson GLM: K=2 GS-IRLS vs centralized")
```

## K=2 Gaussian GLM (HE-Link, unchanged)

For reference, the Gaussian family uses the existing HE-Link (identity link)
mode, which does not require the Gauss-Seidel pathway.

```{r gaussian, message = FALSE}
gauss_k2 <- .cached("gauss_k2_helink", {
  ds.vertGLM(
    data_name   = "D_aligned",
    y_var       = "npreg",
    x_vars      = list(server1 = c("age", "bmi", "ped"),
                       server3 = c("diabetes")),
    y_server    = "server3",
    family      = "gaussian",
    verbose     = FALSE,
    datasources = conns
  )
})

gauss_central <- glm(npreg ~ age + bmi + ped + diabetes,
                      data = pooled, family = gaussian)

cat("K=2 Gaussian (identity link, HE-Link):\n")
cat(sprintf("  Converged: %s in %d iterations\n",
            gauss_k2$converged, gauss_k2$iterations))

coefs_k2_g <- gauss_k2$coefficients
coefs_central_g <- coef(gauss_central)[names(coefs_k2_g)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_g - coefs_central_g))))
```

```{r gaussian-table, echo = FALSE}
comparison_g <- data.frame(
  Variable    = names(coefs_k2_g),
  dsVert_K2   = round(coefs_k2_g, 6),
  Centralized = round(coefs_central_g, 6),
  Difference  = sprintf("%.2e", coefs_k2_g - coefs_central_g),
  row.names   = NULL
)
knitr::kable(comparison_g, caption = "Gaussian GLM: K=2 HE-Link vs centralized")
```

## Cleanup

```{r cleanup}
DSI::datashield.logout(conns)
```

## Summary

| Family | Mode | Iters | Max Coef Error | Privacy |
|--------|------|-------|----------------|---------|
| Binomial | `k2_mpc` (GS-IRLS) | ~7 | $\leq 10^{-4}$ | Coordinator sees $\eta_\text{total}$ |
| Poisson | `k2_mpc` (GS-IRLS) | ~12 | $\leq 10^{-5}$ | Coordinator sees $\eta_\text{total}$ |
| Gaussian | `he_link` (identity) | ~16 | $\leq 10^{-3}$ | $\eta$ encrypted throughout |

The Gauss-Seidel BCD-IRLS mode achieves near-exact coefficient recovery
(matching centralized `glm()` to 4--5 significant figures) in under 15
iterations. The privacy tradeoff is that the coordinator observes
$\boldsymbol{\eta}_\text{total}$, which for K=2 reveals the non-label's
contribution — the same information visible to the coordinator in the K$\geq$3
secure aggregation regime.

Users who require stronger K=2 privacy guarantees can explicitly set
`eta_privacy = "he_link"` to use the CKKS polynomial pathway at the cost of
higher approximation error ($\sim 5 \times 10^{-2}$) and longer runtime.
