---
title: "K=2 HE-Link Optimizer Benchmarks"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{K=2 HE-Link Optimizer Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
```

## Motivation

The K=2 encrypted link-function (HE-Link) mode evaluates the inverse link
function as a polynomial approximation under CKKS homomorphic encryption.
For binomial and Poisson GLMs, this means the optimizer works with
**approximate gradients** rather than exact IRLS updates. This vignette
documents systematic benchmarks of different optimizer configurations to
justify the final design choices.

## Experimental setup

All experiments use the same model and data:

- **Model**: `diabetes ~ age + ped + npreg` (K=2 binomial, 2+1 predictors)
- **Data**: Pima diabetes dataset, n=162 after PSI alignment
- **Centralized reference**: `glm()` coefficients (intercept=-4.47, age=0.070, ped=1.85, npreg=0.113)
- **CKKS parameters**: `log_n=14`, `log_scale=40`
- **Convergence criterion**: max coefficient change < 1e-4

## Configurations tested

Four configurations were compared, varying three factors:

| Config | Optimizer | Sigmoid polynomial | Regularization |
|--------|-----------|-------------------|----------------|
| A | Nesterov momentum (gamma=0.5, alpha=0.15) | [-4, 4] | lambda=1e-4 |
| B | Plain GD (alpha=0.3) | [-8, 8] | lambda=1e-4 |
| C | Nesterov momentum (gamma=0.9, alpha=0.3) | [-8, 8] | lambda=1e-4 |
| D | Nesterov momentum (gamma=0.5, alpha=0.15) | [-4, 4] | lambda=1e-6 |

## Convergence rates

```{r convergence-table, echo = FALSE}
convergence <- data.frame(
  Config = c("A: Nesterov + [-4,4]", "B: Plain GD + [-8,8]",
             "C: Aggressive Nesterov", "D: Nesterov + low lambda"),
  Iter_10 = c("2.44e-02", "2.54e-02", "1.01e-01", "2.45e-02"),
  Iter_20 = c("1.08e-02", "1.27e-02", "7.95e-02", "1.08e-02"),
  Iter_30 = c("6.12e-03", "7.41e-03", "—", "6.13e-03"),
  Rate = c("0.49x / 10 iters", "0.50x / 10 iters",
           "0.79x / 10 iters", "0.49x / 10 iters")
)
knitr::kable(convergence, col.names = c("Configuration", "Diff @ iter 10",
             "Diff @ iter 20", "Diff @ iter 30", "Convergence rate"),
             caption = "Convergence rates across configurations")
```

## Converged results

```{r results-table, echo = FALSE}
results <- data.frame(
  Config = c("A: Nesterov + [-4,4]", "B: Plain GD + [-8,8]",
             "C: Aggressive Nesterov", "D: Nesterov + low lambda"),
  Iters = c(200, 209, "20*", "50*"),
  Converged = c("Yes", "Yes", "No", "No"),
  Max_error = c("1.57", "1.37", ">2", "1.76"),
  Intercept = c("-2.90", "-3.10", "—", "-2.71")
)
knitr::kable(results, col.names = c("Configuration", "Iterations",
             "Converged", "Max coef error", "Intercept (K=2)"),
             caption = "Converged coefficient errors (centralized intercept: -4.47)")
```

*Entries marked with \* did not reach convergence within the iteration limit.*

## Key findings

### 1. Momentum does not improve accuracy

Comparing **A** (Nesterov, gamma=0.5) with **B** (plain GD): both converge at
similar rates (~0.5x per 10 iterations), but plain GD achieves slightly
better coefficient accuracy (1.37 vs 1.57). Aggressive momentum
(**C**, gamma=0.9) diverges entirely.

**Why?** Momentum accumulates velocity across iterations. When the gradient
is systematically biased (as it is with polynomial sigmoid approximation),
momentum amplifies the bias rather than correcting it. This is consistent
with the theory of inexact first-order oracles (Devolder et al., 2014).

### 2. Polynomial domain: [-8, 8] outperforms [-4, 4]

The [-4, 4] polynomial has 7x better approximation accuracy *within* its
domain (max error 2.8e-3 vs 1.9e-2). Yet it produces *worse* converged
coefficients (1.57 vs 1.37).

**Why?** During early GD iterations, `eta_total` can occasionally exceed
[-4, 4] (it is the sum of two standardised components). Outside the fitted
domain, the polynomial diverges from the true sigmoid, causing erratic
gradient updates. The [-8, 8] polynomial is less accurate in-range but
never diverges for practical `eta` values.

### 3. Regularization has negligible effect

Configs **A** (lambda=1e-4) and **D** (lambda=1e-6) show virtually identical
convergence rates and similar converged errors. The L2 penalty is not a
significant source of bias at these scales.

### 4. The error is fundamental

The ~1.4 max coefficient error (concentrated in the intercept) is an
inherent consequence of replacing the exact sigmoid with a degree-7
polynomial. The polynomial is systematically flatter than the true sigmoid,
which biases the gradient and causes the optimizer to converge to a
different point than exact IRLS. No amount of optimizer tuning addresses
this; it would require either a higher-degree polynomial (needing more CKKS
multiplicative levels, i.e., `log_n >= 15`) or a fundamentally different
approach.

## Recommendation

The final dsVert configuration uses:

- **Plain gradient descent** with alpha=0.3 (warm start, then decay)
- **Degree-7 polynomial on [-8, 8]** (widest stable domain)
- **lambda=1e-4** (standard regularization)

This produces the best empirical results and is the simplest implementation.
For applications requiring exact coefficient recovery, **K >= 3 with secure
aggregation** should be preferred (coefficient error ~4e-4).

## Context: why this tradeoff exists

With K >= 3 servers, dsVert uses **secure aggregation** (pairwise masking)
to hide individual `eta_k` contributions. The label server sees only the
sum, computes exact `mu = sigmoid(eta_total)` in plaintext, and runs
standard IRLS. This gives near-exact convergence.

With K = 2, there is no second non-label peer to provide a cancelling mask.
The only way to prevent the label server from learning `eta_nonlabel` is to
keep `eta` encrypted through the sigmoid evaluation — which requires
polynomial approximation under HE. This is the fundamental
**privacy/accuracy tradeoff** of the two-party setting.

| Regime | Sigmoid | Optimizer | Privacy | Max coef error |
|--------|---------|-----------|---------|----------------|
| K >= 3 | Exact (plaintext) | Newton-Raphson IRLS | Secure aggregation | ~4e-4 |
| K = 2 | Polynomial (encrypted) | Gradient descent | Encrypted link function | ~1.4 |

The K=2 mode is designed for scenarios where the privacy guarantee of
hiding `eta_nonlabel` from the label server outweighs the accuracy cost.
