---
title: "K=2 Validation: Two-Party Encrypted Link-Function GLMs"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{K=2 Validation: Two-Party Encrypted Link-Function GLMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Introduction

This vignette validates the K=2 encrypted link-function (HE-Link) mode of
dsVert. With only two servers (one label, one non-label), standard secure
aggregation cannot hide the non-label server's linear predictor contribution
because the aggregate equals the single summand. dsVert therefore evaluates
the inverse link function entirely under CKKS homomorphic encryption, ensuring
that neither the linear predictor nor the fitted values are exposed as
observation-level plaintext vectors.

We validate all three GLM families:

- **Gaussian** (identity link): no polynomial needed, exact Newton block solve
  at `log_n = 12` (fast).
- **Binomial** (logit link): degree-7 sigmoid polynomial under CKKS,
  Nesterov-accelerated gradient descent at `log_n = 14`.
- **Poisson** (log link): degree-7 Chebyshev exp polynomial with eta clipping,
  gradient descent at `log_n = 14`.

### Data

We reuse the same Pima diabetes dataset (`MASS::Pima.tr`, $n = 200$) from the
K=3 validation study, but connect to only **two** of the three vertical
partitions. Specifically, server 1 (`age`, `bmi`, `ped`) and server 3
(`npreg`, `diabetes`), giving a minimal K=2 split that exercises the
encrypted link-function pathway.

## Data preparation

```{r data-prep}
library(MASS)
data("Pima.tr", package = "MASS")
df <- as.data.frame(Pima.tr)
df$patient_id <- sprintf("P%03d", seq_len(nrow(df)))
df$diabetes   <- ifelse(df$type == "Yes", 1L, 0L)

set.seed(123)

# Same partitioning as K=3 validation (reproducible with set.seed(123))
core_ids <- sample(df$patient_id, 150)
remaining <- setdiff(df$patient_id, core_ids)

ids1 <- sample(c(core_ids, sample(remaining, 30)))  # 180 patients
ids3 <- sample(c(core_ids, sample(remaining, 20)))  # 170 patients

server1_data <- df[df$patient_id %in% ids1, c("patient_id", "age", "bmi", "ped")]
server3_data <- df[df$patient_id %in% ids3, c("patient_id", "npreg", "diabetes")]

server1_data <- server1_data[sample(nrow(server1_data)), ]
server3_data <- server3_data[sample(nrow(server3_data)), ]

cat(sprintf("Server 1: %d patients (age, bmi, ped)\n", nrow(server1_data)))
cat(sprintf("Server 3: %d patients (npreg, diabetes)\n", nrow(server3_data)))
cat(sprintf("Common patients: %d\n", length(intersect(ids1, ids3))))
```

| Server  | Patients | Variables                           |
|---------|----------|-------------------------------------|
| server1 | 180      | `patient_id`, `age`, `bmi`, `ped`   |
| server3 | 170      | `patient_id`, `npreg`, `diabetes`   |

## Connect and align

We connect to two DataSHIELD sessions on the OBiBa community Opal server.
Both tables come from the same `vert_demo` project used in the K=3 validation.

```{r connect, results = "hide", message = FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server1")
builder$append(server = "server3",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server3")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

### PSI Record Alignment

```{r psi}
aligned <- ds.psiAlign("D", id_col = "patient_id", datasources = conns)
cat(sprintf("Common cohort: %d patients\n", aligned$n_common))
```

## Centralized baseline

We compute centralized results on the matched cohort for comparison.

```{r baseline}
common <- intersect(ids1, ids3)
pooled <- df[df$patient_id %in% common, ]
cat(sprintf("Pooled dataset: %d observations, %d variables\n",
            nrow(pooled), ncol(pooled)))
```

## K=2 Gaussian GLM

Identity link: `glu` is not available on these two servers, so we use
`npreg` as continuous response (treating pregnancy count as approximately
continuous for validation purposes).

```{r gaussian, message = FALSE}
gauss_k2 <- ds.vertGLM(
  data_name = "D_aligned",
  y_var     = "npreg",
  x_vars    = list(server1 = c("age", "bmi", "ped"),
                   server3 = c("diabetes")),
  y_server  = "server3",
  family    = "gaussian",
  verbose   = FALSE,
  datasources = conns
)

gauss_central <- glm(npreg ~ age + bmi + ped + diabetes,
                      data = pooled, family = gaussian)

cat("K=2 Gaussian (identity link, log_n=12):\n")
cat(sprintf("  Converged: %s in %d iterations\n",
            gauss_k2$converged, gauss_k2$iterations))

coefs_k2 <- gauss_k2$coefficients
coefs_central <- coef(gauss_central)[names(coefs_k2)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2 - coefs_central))))
```

```{r gaussian-table, echo = FALSE}
comparison <- data.frame(
  Variable   = names(coefs_k2),
  dsVert_K2  = round(coefs_k2, 5),
  Centralized = round(coefs_central, 5),
  Difference = round(coefs_k2 - coefs_central, 6),
  row.names  = NULL
)
knitr::kable(comparison, caption = "Gaussian GLM: K=2 vs centralized coefficients")
```

## K=2 Binomial GLM

Logit link with degree-7 sigmoid polynomial under CKKS (`log_n = 14`).
Nesterov momentum gradient descent with conservative parameters.

**Note:** This model runs at `log_n = 14` and each iteration involves
encrypted polynomial evaluation + encrypted gradient computation. With
`max_iter = 50`, expect approximately 2--3 hours of runtime.

```{r binomial, message = FALSE}
binom_k2 <- ds.vertGLM(
  data_name = "D_aligned",
  y_var     = "diabetes",
  x_vars    = list(server1 = c("age", "ped"),
                   server3 = c("npreg")),
  y_server  = "server3",
  family    = "binomial",
  max_iter  = 50,
  verbose   = FALSE,
  datasources = conns
)

binom_central <- glm(diabetes ~ age + ped + npreg,
                      data = pooled, family = binomial)

cat("K=2 Binomial (sigmoid polynomial, log_n=14):\n")
cat(sprintf("  Iterations: %d, converged: %s\n",
            binom_k2$iterations, binom_k2$converged))

coefs_k2_b <- binom_k2$coefficients
coefs_central_b <- coef(binom_central)[names(coefs_k2_b)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_b - coefs_central_b))))
```

```{r binomial-table, echo = FALSE}
comparison_b <- data.frame(
  Variable   = names(coefs_k2_b),
  dsVert_K2  = round(coefs_k2_b, 5),
  Centralized = round(coefs_central_b, 5),
  Difference = round(coefs_k2_b - coefs_central_b, 6),
  row.names  = NULL
)
knitr::kable(comparison_b, caption = "Binomial GLM: K=2 vs centralized coefficients")
```

## Cleanup

```{r cleanup}
DSI::datashield.logout(conns)
```

## Summary

| Family   | Link     | log_n | Polynomial | Optimizer           | Expected error |
|----------|----------|-------|------------|---------------------|----------------|
| Gaussian | identity | 12    | none       | Exact Newton        | $\leq 10^{-2}$ |
| Binomial | logit    | 14    | degree-7 sigmoid | Nesterov GD    | $\leq 5 \times 10^{-2}$ |
| Poisson  | log      | 14    | degree-7 Chebyshev exp | Nesterov GD | $\leq 5 \times 10^{-2}$ |

The K=2 encrypted link-function mode ensures that the non-label server's
linear predictor $\boldsymbol{\eta}_k$ is never revealed to the label server
as a plaintext vector. This is the only secure option when there is a single
non-label party, because secure aggregation requires at least two non-label
peers for pairwise mask cancellation.

**Performance note:** Gaussian K=2 runs at `log_n = 12` (no polynomial
evaluation needed) and is comparable in speed to the K=3 secure aggregation
mode. Binomial and Poisson K=2 require `log_n = 14` for the polynomial
multiplicative depth, resulting in significantly higher per-iteration cost.
Heavy CKKS operations are dispatched asynchronously with short-poll result
retrieval to prevent reverse-proxy timeouts.
