---
title: "K=2 Validation: Two-Party Encrypted Link-Function GLMs"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{K=2 Validation: Two-Party Encrypted Link-Function GLMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# Result caching: expensive dsVert computations are saved as .rds files.
# If cached results exist, they are loaded instead of re-running.
# Delete the cache directory to force re-execution.
cache_dir <- file.path(getwd(), "k2-validation_cache")
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)

.cache_path <- function(name) file.path(cache_dir, paste0(name, ".rds"))

.cached <- function(name, expr) {
  path <- .cache_path(name)
  if (file.exists(path)) {
    message(sprintf("[cache] Loading %s from %s", name, basename(path)))
    return(readRDS(path))
  }
  result <- expr
  saveRDS(result, path)
  message(sprintf("[cache] Saved %s to %s", name, basename(path)))
  result
}
```

## Introduction

This vignette validates the K=2 encrypted link-function (HE-Link) mode of
dsVert. With only two servers (one label, one non-label), standard secure
aggregation cannot hide the non-label server's linear predictor contribution
because the aggregate equals the single summand. dsVert therefore evaluates
the inverse link function entirely under Ring63 MPC, ensuring
that neither the linear predictor nor the fitted values are exposed as
observation-level plaintext vectors.

We validate all three GLM families:

- **Gaussian** (identity link): no polynomial needed, exact Newton block solve
  at `log_n = 12` (fast).
- **Binomial** (logit link): degree-7 sigmoid polynomial under Ring63,
  gradient descent at `log_n = 14`.
- **Poisson** (log link): degree-7 Chebyshev exp polynomial with eta clipping,
  gradient descent at `log_n = 14`.

**Result caching:** Expensive server-side computations are cached as `.rds`
files in `k2-validation_cache/`. Re-rendering the vignette loads cached
results instantly. Delete the cache directory to force re-execution.

### Data

We reuse the same Pima diabetes dataset (`MASS::Pima.tr`, $n = 200$) from the
K=3 validation study, but connect to only **two** of the three vertical
partitions. Specifically, server 1 (`age`, `bmi`, `ped`) and server 3
(`npreg`, `diabetes`), giving a minimal K=2 split that exercises the
encrypted link-function pathway.

## Data preparation

```{r data-prep}
library(MASS)
data("Pima.tr", package = "MASS")
df <- as.data.frame(Pima.tr)
df$patient_id <- sprintf("P%03d", seq_len(nrow(df)))
df$diabetes   <- ifelse(df$type == "Yes", 1L, 0L)

set.seed(123)

# Same partitioning as K=3 validation (reproducible with set.seed(123))
core_ids <- sample(df$patient_id, 150)
remaining <- setdiff(df$patient_id, core_ids)

ids1 <- sample(c(core_ids, sample(remaining, 30)))  # 180 patients
ids3 <- sample(c(core_ids, sample(remaining, 20)))  # 170 patients

server1_data <- df[df$patient_id %in% ids1, c("patient_id", "age", "bmi", "ped")]
server3_data <- df[df$patient_id %in% ids3, c("patient_id", "npreg", "diabetes")]

server1_data <- server1_data[sample(nrow(server1_data)), ]
server3_data <- server3_data[sample(nrow(server3_data)), ]

cat(sprintf("Server 1: %d patients (age, bmi, ped)\n", nrow(server1_data)))
cat(sprintf("Server 3: %d patients (npreg, diabetes)\n", nrow(server3_data)))
cat(sprintf("Common patients: %d\n", length(intersect(ids1, ids3))))
```

| Server  | Patients | Variables                           |
|---------|----------|-------------------------------------|
| server1 | 180      | `patient_id`, `age`, `bmi`, `ped`   |
| server3 | 170      | `patient_id`, `npreg`, `diabetes`   |

## Connect and align

We connect to two DataSHIELD sessions on the OBiBa community Opal server.
Both tables come from the same `vert_demo` project used in the K=3 validation.

```{r connect, results = "hide", message = FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server1")
builder$append(server = "server3",
               url = "https://opal-demo.obiba.org",
               user = "administrator", password = "password",
               driver = "OpalDriver",
               table = "vert_demo.pima_server3")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

### PSI Record Alignment

```{r psi}
aligned <- .cached("psi_result", {
  ds.psiAlign("D", id_col = "patient_id", datasources = conns)
})
cat(sprintf("Common cohort: %d patients\n", aligned$n_common))
```

## Centralized baseline

We compute centralized results on the matched cohort for comparison.

```{r baseline}
common <- intersect(ids1, ids3)
pooled <- df[df$patient_id %in% common, ]
cat(sprintf("Pooled dataset: %d observations, %d variables\n",
            nrow(pooled), ncol(pooled)))
```

## K=2 Gaussian GLM

Identity link: `npreg` as continuous response, `diabetes` as a predictor on
server 3 alongside `age`, `bmi`, `ped` on server 1.

```{r gaussian, message = FALSE}
gauss_k2 <- .cached("gauss_k2", {
  ds.vertGLM(
    data_name = "D_aligned",
    y_var     = "npreg",
    x_vars    = list(server1 = c("age", "bmi", "ped"),
                     server3 = c("diabetes")),
    y_server  = "server3",
    family    = "gaussian",
    verbose   = FALSE,
    datasources = conns
  )
})

gauss_central <- glm(npreg ~ age + bmi + ped + diabetes,
                      data = pooled, family = gaussian)

cat("K=2 Gaussian (identity link, log_n=12):\n")
cat(sprintf("  Converged: %s in %d iterations\n",
            gauss_k2$converged, gauss_k2$iterations))

coefs_k2 <- gauss_k2$coefficients
coefs_central <- coef(gauss_central)[names(coefs_k2)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2 - coefs_central))))
```

```{r gaussian-table, echo = FALSE}
comparison <- data.frame(
  Variable   = names(coefs_k2),
  dsVert_K2  = round(coefs_k2, 5),
  Centralized = round(coefs_central, 5),
  Difference = round(coefs_k2 - coefs_central, 6),
  row.names  = NULL
)
knitr::kable(comparison, caption = "Gaussian GLM: K=2 vs centralized coefficients")
```

## K=2 Binomial GLM

Logit link with degree-7 sigmoid polynomial under Ring63 (`log_n = 14`).
Gradient descent with warm-start step-size schedule.

**Note:** This model runs at `log_n = 14` and each iteration involves
encrypted polynomial evaluation + encrypted gradient computation. With the
default `max_iter`, expect several hours of runtime on first execution.
Subsequent renders load the cached result instantly.

```{r binomial, message = FALSE}
binom_k2 <- .cached("binom_k2", {
  ds.vertGLM(
    data_name = "D_aligned",
    y_var     = "diabetes",
    x_vars    = list(server1 = c("age", "ped"),
                     server3 = c("npreg")),
    y_server  = "server3",
    family    = "binomial",
    verbose   = FALSE,
    datasources = conns
  )
})

binom_central <- glm(diabetes ~ age + ped + npreg,
                      data = pooled, family = binomial)

cat("K=2 Binomial (sigmoid polynomial, log_n=14):\n")
cat(sprintf("  Iterations: %d, converged: %s\n",
            binom_k2$iterations, binom_k2$converged))

coefs_k2_b <- binom_k2$coefficients
coefs_central_b <- coef(binom_central)[names(coefs_k2_b)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_b - coefs_central_b))))
```

```{r binomial-table, echo = FALSE}
comparison_b <- data.frame(
  Variable   = names(coefs_k2_b),
  dsVert_K2  = round(coefs_k2_b, 5),
  Centralized = round(coefs_central_b, 5),
  Difference = round(coefs_k2_b - coefs_central_b, 6),
  row.names  = NULL
)
knitr::kable(comparison_b, caption = "Binomial GLM: K=2 vs centralized coefficients")
```

## K=2 Poisson GLM

Log link with degree-7 Chebyshev exp polynomial on [-3, 3] under Ring63
(`log_n = 14`). Each server clips its eta contribution to [-1.5, 1.5]
before encryption.

```{r poisson, message = FALSE}
poisson_k2 <- .cached("poisson_k2", {
  ds.vertGLM(
    data_name = "D_aligned",
    y_var     = "npreg",
    x_vars    = list(server1 = c("age", "bmi", "ped"),
                     server3 = c("diabetes")),
    y_server  = "server3",
    family    = "poisson",
    verbose   = FALSE,
    datasources = conns
  )
})

poisson_central <- glm(npreg ~ age + bmi + ped + diabetes,
                        data = pooled, family = poisson)

cat("K=2 Poisson (exp polynomial, log_n=14):\n")
cat(sprintf("  Iterations: %d, converged: %s\n",
            poisson_k2$iterations, poisson_k2$converged))

coefs_k2_p <- poisson_k2$coefficients
coefs_central_p <- coef(poisson_central)[names(coefs_k2_p)]
cat(sprintf("  Max abs coefficient error: %.4e\n",
            max(abs(coefs_k2_p - coefs_central_p))))
```

```{r poisson-table, echo = FALSE}
comparison_p <- data.frame(
  Variable   = names(coefs_k2_p),
  dsVert_K2  = round(coefs_k2_p, 5),
  Centralized = round(coefs_central_p, 5),
  Difference = round(coefs_k2_p - coefs_central_p, 6),
  row.names  = NULL
)
knitr::kable(comparison_p, caption = "Poisson GLM: K=2 vs centralized coefficients")
```

## Cleanup

```{r cleanup}
DSI::datashield.logout(conns)
```

## Summary

| Family   | Link     | log_n | Polynomial | Optimizer           | Expected error |
|----------|----------|-------|------------|---------------------|----------------|
| Gaussian | identity | 12    | none       | Exact Newton        | $\leq 10^{-2}$ |
| Binomial | logit    | 14    | degree-7 sigmoid | GD    | $\leq 5 \times 10^{-2}$ |
| Poisson  | log      | 14    | degree-7 Chebyshev exp | GD | $\leq 5 \times 10^{-2}$ |

The K=2 encrypted link-function mode ensures that the non-label server's
linear predictor $\boldsymbol{\eta}_k$ is never revealed to the label server
as a plaintext vector. This is the only secure option when there is a single
non-label party, because secure aggregation requires at least two non-label
peers for pairwise mask cancellation.

**Performance note:** Gaussian K=2 runs at `log_n = 12` (no polynomial
evaluation needed) and is comparable in speed to the K=3 secure aggregation
mode. Binomial and Poisson K=2 require `log_n = 14` for the polynomial
multiplicative depth, resulting in significantly higher per-iteration cost.
Encrypted operations are dispatched asynchronously with short-poll result
retrieval to prevent reverse-proxy timeouts.
