---
title: "Multi-Instance Deployment: Three Independent Opal Servers"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Instance Deployment: Three Independent Opal Servers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

col_palette <- colorRampPalette(c("#2166AC", "#67A9CF", "#D1E5F0",
                                   "#FDDBC7", "#EF8A62", "#B2182B"))(100)
diff_palette <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
col_local  <- "#4393C3"
col_dsvert <- "#D6604D"

# Result caching: expensive dsVert computations are saved as .rds files.
# If cached results exist, they are loaded instead of re-running.
# Delete the cache directory to force re-execution.
cache_dir <- file.path(getwd(), "local-deployment_cache")
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)

.cache_path <- function(name) file.path(cache_dir, paste0(name, ".rds"))

.cached <- function(name, expr) {
  path <- .cache_path(name)
  if (file.exists(path)) {
    message(sprintf("[cache] Loading %s from %s", name, basename(path)))
    return(readRDS(path))
  }
  result <- expr
  saveRDS(result, path)
  message(sprintf("[cache] Saved %s to %s", name, basename(path)))
  result
}
```

## Overview

This vignette demonstrates dsVert running across **three independent Opal/Rock
instances**, each on a separate host and port. Unlike the
[public validation vignette](validation.html) (which uses a single Opal server
with three tables), this setup mirrors a real multi-institution deployment where
each data controller operates its own infrastructure.

The analysis is identical to the public validation: PSI alignment, MHE-CKKS
correlation, PCA, and three GLM families (Gaussian, binomial, Poisson),
compared element-wise against centralized R outputs.

### Data preparation and vertical partitioning

```{r data-prep}
library(MASS)
data("Pima.tr", package = "MASS")
df <- as.data.frame(Pima.tr)
df$patient_id <- sprintf("P%03d", seq_len(nrow(df)))
df$diabetes   <- ifelse(df$type == "Yes", 1L, 0L)

set.seed(123)

core_ids <- sample(df$patient_id, 150)
remaining <- setdiff(df$patient_id, core_ids)

ids1 <- sample(c(core_ids, sample(remaining, 30)))
ids2 <- sample(c(core_ids, sample(remaining, 25)))
ids3 <- sample(c(core_ids, sample(remaining, 20)))

server1_data <- df[df$patient_id %in% ids1, c("patient_id", "age", "bmi", "ped")]
server2_data <- df[df$patient_id %in% ids2, c("patient_id", "glu", "bp", "skin")]
server3_data <- df[df$patient_id %in% ids3, c("patient_id", "npreg", "diabetes")]

server1_data <- server1_data[sample(nrow(server1_data)), ]
server2_data <- server2_data[sample(nrow(server2_data)), ]
server3_data <- server3_data[sample(nrow(server3_data)), ]

cat(sprintf("Server 1: %d patients (age, bmi, ped)\n", nrow(server1_data)))
cat(sprintf("Server 2: %d patients (glu, bp, skin)\n", nrow(server2_data)))
cat(sprintf("Server 3: %d patients (npreg, diabetes)\n", nrow(server3_data)))
cat(sprintf("True intersection: %d patients\n",
    length(Reduce(intersect, list(ids1, ids2, ids3)))))
```

Each server runs on an independent Opal/Rock instance with its own Docker
stack, R session, and data store:

| Server  | Host | Patients | Variables                         |
|---------|------|----------|-----------------------------------|
| server1 | localhost:8443 | 180 | `patient_id`, `age`, `bmi`, `ped` |
| server2 | localhost:8444 | 175 | `patient_id`, `glu`, `bp`, `skin` |
| server3 | localhost:8445 | 170 | `patient_id`, `npreg`, `diabetes` |

---

## Connect to three independent servers

Each connection points to a **different host**. This is the key difference
from the public demo where all three point to the same Opal.

```{r connect, results = "hide", message = FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

options(opal.verifyssl = FALSE)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1", url = "https://localhost:8443",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server2", url = "https://localhost:8444",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server3", url = "https://localhost:8445",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

## PSI Record Alignment

```{r psi, message = FALSE}
psi_result <- .cached("psi_result", {
  ds.psiAlign(
    data_name   = "D",
    id_col      = "patient_id",
    newobj      = "D_aligned",
    datasources = conns
  )
})
```

```{r psi-summary, echo = FALSE}
n_matched <- psi_result$server1$n_matched
```

The PSI protocol identified **`r n_matched`** patients present
in all three servers.

---

## Centralized ground truth

```{r ground-truth}
full <- merge(merge(server1_data, server2_data, by = "patient_id"),
              server3_data, by = "patient_id")
full <- full[order(full$patient_id), ]
num_vars <- c("age", "bmi", "ped", "glu", "bp", "skin", "npreg")
local_cor <- cor(full[, num_vars])
local_pca <- eigen(local_cor)
local_gauss <- glm(glu ~ age + bmi + ped + bp + skin + npreg + diabetes,
                   data = full, family = gaussian)
local_binom <- glm(diabetes ~ age + bmi + ped + glu + bp + skin + npreg,
                   data = full, family = binomial)
local_pois  <- glm(npreg ~ age + bmi + ped + glu + bp + skin + diabetes,
                   data = full, family = poisson)
cat("Ground truth:", nrow(full), "rows\n")
```

---

## Federated analysis (dsVert across 3 independent servers)

### Correlation (MHE-CKKS)

```{r dsvert-cor, message = FALSE}
variables <- list(
  server1 = c("age", "bmi", "ped"),
  server2 = c("glu", "bp", "skin"),
  server3 = c("npreg")
)

cor_result <- .cached("cor_result", {
  ds.vertCor(
    data_name   = "D_aligned",
    variables   = variables,
    log_n       = 12,
    log_scale   = 40,
    datasources = conns
  )
})
dsvert_cor <- cor_result$correlation
```

### PCA

```{r dsvert-pca}
dsvert_pca <- eigen(dsvert_cor)
dsvert_eigenvalues <- dsvert_pca$values
local_eigenvalues  <- local_pca$values
```

### GLM --- Gaussian (y = `glu`)

```{r dsvert-gauss, message = FALSE}
glm_gauss <- .cached("glm_gauss", {
  ds.vertGLM(
    data_name = "D_aligned", y_var = "glu",
    x_vars    = list(server1 = c("age", "bmi", "ped"),
                     server2 = c("bp", "skin"),
                     server3 = c("npreg", "diabetes")),
    y_server  = "server2", family = "gaussian",
    max_iter = 100, tol = 1e-4, lambda = 1e-4,
    log_n = 12, log_scale = 40,
    datasources = conns
  )
})
```

### GLM --- Binomial (y = `diabetes`)

```{r dsvert-binom, message = FALSE}
glm_binom <- .cached("glm_binom", {
  ds.vertGLM(
    data_name = "D_aligned", y_var = "diabetes",
    x_vars    = list(server1 = c("age", "bmi", "ped"),
                     server2 = c("glu", "bp", "skin"),
                     server3 = c("npreg")),
    y_server  = "server3", family = "binomial",
    max_iter = 100, tol = 1e-4, lambda = 1e-4,
    log_n = 12, log_scale = 40,
    datasources = conns
  )
})
```

### GLM --- Poisson (y = `npreg`)

```{r dsvert-pois, message = FALSE}
glm_pois <- .cached("glm_pois", {
  ds.vertGLM(
    data_name = "D_aligned", y_var = "npreg",
    x_vars    = list(server1 = c("age", "bmi", "ped"),
                     server2 = c("glu", "bp", "skin"),
                     server3 = c("diabetes")),
    y_server  = "server3", family = "poisson",
    max_iter = 100, tol = 1e-4, lambda = 1e-4,
    log_n = 12, log_scale = 40,
    datasources = conns
  )
})
```

---

## Comparative assessment

```{r comparison, echo = FALSE}
cor_diff   <- dsvert_cor - local_cor
eig_diff   <- dsvert_eigenvalues - local_eigenvalues
gauss_diff <- coef(glm_gauss) - coef(local_gauss)
binom_diff <- coef(glm_binom) - coef(local_binom)
pois_diff  <- coef(glm_pois)  - coef(local_pois)

knitr::kable(
  data.frame(
    Protocol = c("PSI Alignment", "Correlation", "PCA",
                 "GLM Gaussian", "GLM Binomial", "GLM Poisson"),
    Metric = c("Matched cohort",
               "Max |r| error",
               "Max eigenvalue error",
               "Max |coef| error",
               "Max |coef| error",
               "Max |coef| error"),
    Value = c(paste0(n_matched, " / ", nrow(df)),
              format(max(abs(cor_diff)), digits = 3, scientific = TRUE),
              format(max(abs(eig_diff)), digits = 3, scientific = TRUE),
              format(max(abs(gauss_diff)), digits = 3, scientific = TRUE),
              format(max(abs(binom_diff)), digits = 3, scientific = TRUE),
              format(max(abs(pois_diff)), digits = 3, scientific = TRUE))
  ),
  align = c("l", "l", "r"),
  caption = "Multi-instance validation (3 independent Opals, dsVert vs. centralized R)"
)
```

### Correlation difference heatmap

```{r cor-diff-heatmap, echo = FALSE, fig.height = 5.5}
max_abs <- max(abs(cor_diff))
image(seq_along(num_vars), seq_along(num_vars), cor_diff,
      col = diff_palette, zlim = c(-max_abs, max_abs),
      xlab = "", ylab = "", axes = FALSE,
      main = paste0("Correlation Difference (3-server dsVert - Centralized)\nmax |diff| = ",
                     format(max_abs, digits = 3, scientific = TRUE)))
axis(1, at = seq_along(num_vars), labels = num_vars, las = 2)
axis(2, at = seq_along(num_vars), labels = num_vars, las = 1)
for (i in seq_along(num_vars)) for (j in seq_along(num_vars))
  text(i, j, format(cor_diff[i, j], digits = 2, scientific = TRUE), cex = 0.6)
```

### GLM coefficient comparison

```{r coeff-plot, echo = FALSE, fig.width = 8, fig.height = 5}
gauss_local  <- coef(local_gauss)
gauss_dsvert <- coef(glm_gauss)
idx <- seq_along(gauss_local)
par(mar = c(5, 4, 3, 8))
plot(idx - 0.1, gauss_local, pch = 16, col = col_local, cex = 1.5,
     xlab = "", ylab = "Coefficient", main = "Gaussian GLM (3 independent Opals)",
     xaxt = "n", xlim = c(0.5, length(idx) + 0.5),
     ylim = range(c(gauss_local, gauss_dsvert)) * 1.1)
points(idx + 0.1, gauss_dsvert, pch = 17, col = col_dsvert, cex = 1.5)
segments(idx - 0.1, gauss_local, idx + 0.1, gauss_dsvert,
         col = "grey60", lty = 2)
axis(1, at = idx, labels = names(gauss_local), las = 1)
abline(h = 0, col = "grey80")
legend(par("usr")[2] + 0.4, par("usr")[4],
       legend = c("Centralized", "dsVert (3 Opals)"),
       pch = c(16, 17), col = c(col_local, col_dsvert),
       bty = "o", bg = "white", box.col = "black", xpd = TRUE)
```

## Comparison with public demo

Results obtained on three independent Opal instances are numerically
identical to those from the [public validation](validation.html) on
`opal-demo.obiba.org`. This confirms that the dsVert protocol creates fully
isolated sessions regardless of whether connections point to the same or
different hosts.

| Aspect | Public demo (opal-demo.obiba.org) | This vignette (3 Opals) |
|--------|-----------------------------------|--------------------------|
| Servers | 1 Opal, 3 tables | 3 Opals, 3 hosts |
| Network | Client → single server | Client → 3 separate servers |
| R sessions | 3 sessions on same Rock | 3 sessions on 3 independent Rocks |
| CPU/memory | Shared | Independent per instance |
| Reproducibility | Anyone can run it | Requires local Docker setup |
| Numerical results | Identical | Identical |

```{r disconnect, include = FALSE}
DSI::datashield.logout(conns)
```
