---
title: "Multi-Instance Deployment: Three Independent Opal Servers"
author: "David Sarrat González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Instance Deployment: Three Independent Opal Servers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates dsVert running across **three independent Opal/Rock
instances**, each on a separate host and port. Unlike the
[public validation vignette](validation.html) (which uses a single Opal server
with three tables), this setup mirrors a real multi-institution deployment where
each data controller operates its own infrastructure.

## Prerequisites

You need three Opal instances running. The easiest way is
[easy-opal](https://github.com/davidsarratgonzalez/easy-opal):

```bash
# Create and start 3 independent Opal stacks
easy-opal instance create opal1
easy-opal instance create opal2
easy-opal instance create opal3

easy-opal -i opal1 setup --stack-name opal1 --port 8443 --password admin123 --ssl-strategy self-signed --yes
easy-opal -i opal2 setup --stack-name opal2 --port 8444 --password admin123 --ssl-strategy self-signed --yes
easy-opal -i opal3 setup --stack-name opal3 --port 8445 --password admin123 --ssl-strategy self-signed --yes
```

Each instance runs its own Opal, Rock (R server), MongoDB, and Nginx in
isolated Docker containers. They share no state.

## Install dsVert on each server

```{r install}
library(opalr)

# Enable R package management and install dsVert on each instance
for (port in c(8443, 8444, 8445)) {
  o <- opal.login("administrator", "admin123",
                   url = paste0("https://localhost:", port),
                   opts = list(ssl_verifyhost = 0, ssl_verifypeer = 0))

  # Enable package management (one-time)
  cfg <- opal.get(o, "system", "conf", "general")
  cfg$allowRPackageManagement <- TRUE
  opal.put(o, "system", "conf", "general",
           body = jsonlite::toJSON(cfg, auto_unbox = TRUE),
           contentType = "application/json")

  # Install dsVert from GitHub
  dsadmin.install_github_package(o, "dsVert", username = "isglobal-brge")
  opal.logout(o)
}
```

## Prepare the dataset

We generate the same vertically partitioned Pima dataset used in the main
validation vignette, with deliberately non-identical patient populations
across servers.

```{r data-prep}
library(MASS)
data("Pima.tr", package = "MASS")
df <- as.data.frame(Pima.tr)
df$patient_id <- sprintf("P%03d", seq_len(nrow(df)))
df$diabetes <- ifelse(df$type == "Yes", 1L, 0L)

set.seed(123)
core_ids <- sample(df$patient_id, 150)
remaining <- setdiff(df$patient_id, core_ids)
ids1 <- sample(c(core_ids, sample(remaining, 30)))
ids2 <- sample(c(core_ids, sample(remaining, 25)))
ids3 <- sample(c(core_ids, sample(remaining, 20)))

server1_data <- df[df$patient_id %in% ids1, c("patient_id", "age", "bmi", "ped")]
server2_data <- df[df$patient_id %in% ids2, c("patient_id", "glu", "bp", "skin")]
server3_data <- df[df$patient_id %in% ids3, c("patient_id", "npreg", "diabetes")]

server1_data <- server1_data[sample(nrow(server1_data)), ]
server2_data <- server2_data[sample(nrow(server2_data)), ]
server3_data <- server3_data[sample(nrow(server3_data)), ]
```

## Upload data to each server

Each server receives only its own vertical partition. No server sees the
other servers' variables.

```{r upload}
library(opalr)

servers <- list(
  list(url = "https://localhost:8443", data = server1_data),
  list(url = "https://localhost:8444", data = server2_data),
  list(url = "https://localhost:8445", data = server3_data)
)

for (s in servers) {
  o <- opal.login("administrator", "admin123", url = s$url,
                   opts = list(ssl_verifyhost = 0, ssl_verifypeer = 0))
  tryCatch(opal.project_delete(o, "pima_project"), error = function(e) NULL)
  opal.project_create(o, "pima_project", database = TRUE)
  tbl <- s$data
  tbl$id <- seq_len(nrow(tbl))
  opal.table_save(o, tibble::as_tibble(tbl), "pima_project", "pima_data",
                   overwrite = TRUE, force = TRUE, id.name = "id")
  cat("Uploaded", nrow(tbl), "rows to", s$url, "\n")
  opal.logout(o)
}
```

## Connect and run the full pipeline

This is the key difference from the public demo: each server connection
points to a **different host**.

```{r connect}
library(DSI)
library(DSOpal)
library(dsVertClient)

options(opal.verifyssl = FALSE)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1", url = "https://localhost:8443",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server2", url = "https://localhost:8444",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server3", url = "https://localhost:8445",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

### PSI Record Alignment

```{r psi}
psi_result <- ds.psiAlign(
  data_name   = "D",
  id_col      = "patient_id",
  newobj      = "D_aligned",
  datasources = conns
)
```

### Correlation (MHE-CKKS)

```{r cor}
variables <- list(
  server1 = c("age", "bmi", "ped"),
  server2 = c("glu", "bp", "skin"),
  server3 = c("npreg")
)

cor_result <- ds.vertCor(
  data_name   = "D_aligned",
  variables   = variables,
  log_n       = 12,
  log_scale   = 40,
  datasources = conns
)
```

### GLM (all three families)

```{r glm}
# Gaussian
glm_gauss <- ds.vertGLM(
  data_name = "D_aligned", y_var = "glu",
  x_vars = list(server1 = c("age", "bmi", "ped"),
                server2 = c("bp", "skin"),
                server3 = c("npreg", "diabetes")),
  y_server = "server2", family = "gaussian",
  datasources = conns
)

# Binomial
glm_binom <- ds.vertGLM(
  data_name = "D_aligned", y_var = "diabetes",
  x_vars = list(server1 = c("age", "bmi", "ped"),
                server2 = c("glu", "bp", "skin"),
                server3 = c("npreg")),
  y_server = "server3", family = "binomial",
  datasources = conns
)

# Poisson
glm_pois <- ds.vertGLM(
  data_name = "D_aligned", y_var = "npreg",
  x_vars = list(server1 = c("age", "bmi", "ped"),
                server2 = c("glu", "bp", "skin"),
                server3 = c("diabetes")),
  y_server = "server3", family = "poisson",
  datasources = conns
)
```

### Disconnect

```{r disconnect}
DSI::datashield.logout(conns)
```

## How this differs from the public demo

| Aspect | Public demo (opal-demo.obiba.org) | Multi-instance (this vignette) |
|--------|-----------------------------------|-------------------------------|
| Servers | 1 Opal, 3 tables | 3 Opals, 3 hosts |
| Network | Same host (loopback) | Separate containers (can be separate machines) |
| R sessions | 3 sessions on same Rock | 3 sessions on 3 independent Rocks |
| CPU/memory | Shared | Independent per instance |
| Reproducibility | Anyone can run it | Requires local Docker setup |

Both produce **identical numerical results** because the dsVert protocol
creates fully isolated DataSHIELD sessions regardless of host topology.
The cryptographic keys, ciphertexts, and protocol state are scoped per
session and never shared between connections.
