---
title: "dsVert methods tour — twelve estimators with DSLite demos"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{dsVert methods tour — twelve estimators with DSLite demos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: |
  A reproducible end-to-end tour of every dsVert estimator. Each section
  uses an in-process DSLite backend so the user can run the entire
  vignette without setting up Docker / Opal / Rocks — install
  `dsVertClient` and `DSLite` and the chunks run unchanged. For
  production deployment against a real Opal cluster, replace the DSLite
  `newDSLiteServer` calls with `DSI::newDSLoginBuilder` + the `OpalDriver`;
  every other line in each chunk stays the same.

  The login boilerplate is repeated per-method on purpose: each section
  is meant to be readable top-to-bottom in isolation, so a reviewer can
  open one section without chasing helpers. Production-equivalent
  numerical validation against `opal{1,2,3}-rock` is documented in the
  companion `vert_validation_evidence` article.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  cache = TRUE, cache.path = "methods-tour_cache/",
  fig.path = "methods-tour_figures/",
  fig.width = 7, fig.height = 4)
```

```{r libs}
suppressPackageStartupMessages({
  library(DSLite); library(DSI); library(dsVertClient)
  library(MASS); library(survival); library(nlme); library(lme4)
  library(geepack); library(nnet)
})
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)
```

# Convention

Every section follows the same five-step recipe:

1. **Generate the federated split.** A small synthetic data frame is
   split into a left-server table and a right-server table, both
   sharing a `patient_id` key. For methods with a K=3 path we also
   show a 3-server split.
2. **Stand up the DSLite servers.** `newDSLiteServer` creates two (or
   three) in-process Opal-equivalent endpoints. `newDSLoginBuilder` /
   `datashield.login` open the connections, just as they would against
   a real Opal cluster.
3. **Align on `patient_id`** with `ds.psiAlign` (ECDH-PSI, identical
   across DSLite and real Opal).
4. **Fit the federated estimator** and the centralised reference on
   the pooled data.
5. **Compare** coefficients + report `max|Δβ|`. Logout.

```{r dslite-server-builder}
## Helper — used in every section. Each section creates ITS OWN DSLite
## environment so sections do not interact.
make_dslite <- function(tables, server_names) {
  ## Each server gets its own table; all tables share patient_id key.
  envs <- setNames(lapply(seq_along(tables), function(i) {
    e <- new.env(parent = globalenv())
    assign(server_names[i], tables[[i]], envir = e)
    e
  }), server_names)
  envs
}
```

# Method 1 — GLM (binomial, K=2)

```{r glm-k2}
set.seed(1L)
n <- 200L
pid <- sprintf("P%04d", seq_len(n))
age <- rnorm(n, 50, 10); bmi <- rnorm(n, 26, 4); ped <- runif(n, 0, 2.5)
glu <- rnorm(n, 120, 25); bp <- rnorm(n, 72, 10)
eta <- -3 + 0.02 * age + 0.04 * bmi + 0.5 * ped + 0.005 * glu + 0.01 * bp
y <- rbinom(n, 1, plogis(eta))
s1 <- data.frame(patient_id = pid, age = age, bmi = bmi, ped = ped)
s2 <- data.frame(patient_id = pid, glu = glu, bp = bp, y = y)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server",
         table = "s1", driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server",
         table = "s2", driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))

psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
fit_ds <- ds.vertGLM(y ~ age + bmi + ped + glu + bp, data = "DA",
                      family = "binomial", verbose = FALSE,
                      datasources = conns)
fit_ref <- glm(y ~ age + bmi + ped + glu + bp,
                data = pooled, family = binomial)
nm <- intersect(names(fit_ds$coefficients), names(coef(fit_ref)))
delta <- max(abs(fit_ds$coefficients[nm] - coef(fit_ref)[nm]))
cat(sprintf("max|Δβ| = %.3e\n", delta))
datashield.logout(conns)
```

# Method 2 — GLM K=3 (binomial, three sites)

```{r glm-k3}
set.seed(2L)
n <- 200L
pid <- sprintf("P%04d", seq_len(n))
age <- rnorm(n); bmi <- rnorm(n); ped <- runif(n)
glu <- rnorm(n); bp <- rnorm(n); npreg <- rpois(n, 2)
eta <- -1 + 0.4 * age + 0.3 * bmi + 0.2 * ped + 0.5 * glu - 0.1 * bp
y <- rbinom(n, 1, plogis(eta))
t1 <- data.frame(patient_id = pid, age = age, bmi = bmi)
t2 <- data.frame(patient_id = pid, ped = ped, glu = glu)
t3 <- data.frame(patient_id = pid, bp = bp, npreg = npreg, y = y)
pooled <- Reduce(function(a, b) merge(a, b, by = "patient_id"),
                  list(t1, t2, t3))

dslite.server <- newDSLiteServer(tables = list(s1 = t1, s2 = t2, s3 = t3),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
b$append(server = "site3", url = "dslite.server", table = "s3",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_ds <- ds.vertGLM(y ~ age + bmi + ped + glu + bp + npreg, data = "DA",
                      family = "binomial", verbose = FALSE,
                      datasources = conns)
fit_ref <- glm(y ~ age + bmi + ped + glu + bp + npreg,
                data = pooled, family = binomial)
nm <- intersect(names(fit_ds$coefficients), names(coef(fit_ref)))
delta <- max(abs(fit_ds$coefficients[nm] - coef(fit_ref)[nm]))
cat(sprintf("max|Δβ| K=3 = %.3e\n", delta))
datashield.logout(conns)
```

# Method 3 — Negative binomial (full regression θ MLE, K=2)

```{r nb-k2}
set.seed(3L)
n <- 200L
pid <- sprintf("P%04d", seq_len(n))
x1 <- rnorm(n); x2 <- rnorm(n)
mu <- exp(0.5 + 0.3 * x1 - 0.2 * x2)
y  <- MASS::rnegbin(n, mu = mu, theta = 2)
s1 <- data.frame(patient_id = pid, x1 = x1, x2 = x2)
s2 <- data.frame(patient_id = pid, y = y)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_ds <- ds.vertNB(y ~ x1 + x2, data = "DA",
                     verbose = FALSE, datasources = conns)
fit_ref <- MASS::glm.nb(y ~ x1 + x2, data = pooled,
                         control = glm.control(maxit = 200))
cat(sprintf("θ̂ dsvert=%.3f  glm.nb=%.3f  |Δθ|=%.3e\n",
             fit_ds$theta, fit_ref$theta,
             abs(fit_ds$theta - fit_ref$theta)))
datashield.logout(conns)
```

# Method 4 — Linear mixed model (REML, K=2)

```{r lmm-k2}
set.seed(4L)
n_cl <- 30L; n_obs <- 8L; n <- n_cl * n_obs
cluster_id <- rep(sprintf("C%02d", seq_len(n_cl)), each = n_obs)
pid <- sprintf("R%04d", seq_len(n))
b_i <- rep(rnorm(n_cl, 0, sqrt(0.5)), each = n_obs)
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.5)
y <- 5 + 0.5 * x1 + 0.3 * x2 + b_i + rnorm(n, 0, 1)
s1 <- data.frame(patient_id = pid, x1 = x1)
s2 <- data.frame(patient_id = pid, cluster_id = cluster_id, x2 = x2, y = y)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_ds <- ds.vertLMM(y ~ x1 + x2, data = "DA",
                      cluster_col = "cluster_id", reml = TRUE,
                      verbose = FALSE, datasources = conns)
fit_ref <- nlme::lme(y ~ x1 + x2, random = ~ 1 | cluster_id,
                      data = pooled, method = "REML")
cat(sprintf("max|Δβ|        : %.3e\n",
             max(abs(as.numeric(fit_ds$coefficients) -
                       as.numeric(nlme::fixef(fit_ref))))))
datashield.logout(conns)
```

# Method 5 — GEE sandwich (K=2)

```{r gee-k2}
set.seed(5L)
n_cl <- 30L; n_obs <- 8L; n <- n_cl * n_obs
cluster_id <- rep(sprintf("C%02d", seq_len(n_cl)), each = n_obs)
pid <- sprintf("R%04d", seq_len(n))
b_i <- rep(rnorm(n_cl, 0, sqrt(0.5)), each = n_obs)
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.5)
y <- 5 + 0.5 * x1 + 0.3 * x2 + b_i + rnorm(n, 0, 1)
s1 <- data.frame(patient_id = pid, x1 = x1)
s2 <- data.frame(patient_id = pid, cluster_id = cluster_id,
                  x2 = x2, y = y)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_ds <- ds.vertGEE(y ~ x1 + x2, data = "DA", family = "gaussian",
                      id_col = "cluster_id", corstr = "exchangeable",
                      verbose = FALSE, datasources = conns)
fit_ref <- geepack::geeglm(y ~ x1 + x2, id = cluster_id, data = pooled,
                             family = gaussian, corstr = "exchangeable")
cat(sprintf("max|Δβ| = %.3e\n",
             max(abs(as.numeric(fit_ds$coefficients) -
                       as.numeric(coef(fit_ref))))))
datashield.logout(conns)
```

# Method 6 — Binomial GLMM (Laplace, K=2)

```{r glmm-k2}
set.seed(6L)
n_cl <- 50L; n_obs <- 8L; n <- n_cl * n_obs
cluster_id <- rep(sprintf("C%03d", seq_len(n_cl)), each = n_obs)
pid <- sprintf("R%04d", seq_len(n))
b_i <- rep(rnorm(n_cl, 0, sqrt(0.5)), each = n_obs)
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.5)
eta <- -0.85 + 0.5 * x1 + 0.3 * x2 + b_i
y <- rbinom(n, 1, plogis(eta))
s1 <- data.frame(patient_id = pid, x1 = x1, anchor = 1L)
s2 <- data.frame(patient_id = pid, cluster_id = cluster_id,
                  x2 = x2, y = y)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
fit_ds <- ds.vertGLMM(y ~ x1 + x2, data = "DA",
                       cluster_col = "cluster_id",
                       max_outer = 6L, inner_iter = 5L, tol = 1e-3,
                       verbose = FALSE, datasources = conns)
fit_ref <- lme4::glmer(y ~ x1 + x2 + (1 | cluster_id),
                        data = pooled, family = binomial(),
                        control = lme4::glmerControl(optimizer = "bobyqa"))
cat(sprintf("max|Δβ| = %.3e (well-identified regime)\n",
             max(abs(as.numeric(fit_ds$coefficients) -
                       as.numeric(lme4::fixef(fit_ref))))))
datashield.logout(conns)
```

# Method 7 — Cox proportional-hazards (K=2 OT-Beaver)

```{r cox-k2}
set.seed(7L)
n <- 150L
pid <- sprintf("L%03d", seq_len(n))
age <- rnorm(n, 60, 10); sex_num <- rbinom(n, 1, 0.5)
hazard_lin <- 0.03 * (age - 60) + 0.4 * sex_num
time <- rexp(n, rate = 0.05 * exp(hazard_lin))
status <- as.integer(rexp(n, 0.04) > time)
s1 <- data.frame(patient_id = pid, age = age, sex_num = sex_num)
s2 <- data.frame(patient_id = pid, time = time, status = status)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
fit_ds <- ds.vertCox(time, status, ~ age + sex_num,
                      data = "DA",
                      verbose = FALSE, datasources = conns)
fit_ref <- survival::coxph(survival::Surv(time, status) ~ age + sex_num,
                             data = pooled)
cat(sprintf("max|Δβ| = %.3e\n",
             max(abs(as.numeric(fit_ds$coefficients) -
                       as.numeric(coef(fit_ref))))))
datashield.logout(conns)
```

# Method 8 — Ordinal proportional-odds (joint Newton, K=2)

```{r ord-joint-k2}
set.seed(8L)
n <- 200L
pid <- sprintf("R%04d", seq_len(n))
x1 <- rnorm(n); x2 <- rnorm(n)
eta <- 0.5 * x1 - 0.3 * x2
cuts <- c(-1, 0.5)
prob_low  <- plogis(cuts[1L] - eta)
prob_med  <- plogis(cuts[2L] - eta) - prob_low
y <- factor(apply(cbind(prob_low,
                          plogis(cuts[2L] - eta) - prob_low,
                          1 - plogis(cuts[2L] - eta)),
                    1, function(p) sample(c("low","med","high"), 1, prob=p)),
             levels = c("low","med","high"), ordered = TRUE)
low_leq <- as.integer(as.integer(y) <= 1L)
med_leq <- as.integer(as.integer(y) <= 2L)
high_leq <- 1L
s1 <- data.frame(patient_id = pid, x1 = x1, x2 = x2)
s2 <- data.frame(patient_id = pid, y = y,
                  low_leq = low_leq, med_leq = med_leq, high_leq = high_leq)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
fit_ds <- ds.vertOrdinalJointNewton(
  y ~ x1 + x2, data = "DA",
  levels_ordered = c("low","med","high"),
  cumulative_template = "%s_leq",
  max_outer = 8L, tol = 1e-6,
  verbose = FALSE, datasources = conns)
fit_ref <- MASS::polr(y ~ x1 + x2, data = pooled,
                        method = "logistic", Hess = TRUE)
ds_b <- as.numeric(fit_ds$beta_po_joint %||% fit_ds$coefficients)
rf_b <- as.numeric(coef(fit_ref))
cat(sprintf("max|Δβ| = %.3e\n",
             if (length(ds_b) == length(rf_b)) max(abs(ds_b - rf_b)) else NA))
datashield.logout(conns)
```

# Method 9 — Multinomial joint Newton (K=2)

```{r mnl-joint-k2}
set.seed(9L)
n <- 200L
pid <- sprintf("R%04d", seq_len(n))
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.5)
eta_med <- -0.5 + 0.4 * x1 + 0.3 * x2
eta_high <- -0.2 + 0.6 * x1 - 0.2 * x2
denom <- 1 + exp(eta_med) + exp(eta_high)
P <- cbind(low = 1/denom, med = exp(eta_med)/denom,
            high = exp(eta_high)/denom)
y <- factor(apply(P, 1, function(p) sample(colnames(P), 1, prob = p)),
             levels = c("low","med","high"))
low_ind <- as.integer(y == "low")
med_ind <- as.integer(y == "med")
high_ind <- as.integer(y == "high")
s1 <- data.frame(patient_id = pid, x1 = x1, x2 = x2)
s2 <- data.frame(patient_id = pid, y = y,
                  low_ind = low_ind, med_ind = med_ind, high_ind = high_ind)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_ds <- ds.vertMultinomJointNewton(
  y ~ x1 + x2, data = "DA",
  levels = c("low","med","high"),
  indicator_template = "%s_ind",
  max_outer = 8L, tol = 1e-6,
  verbose = FALSE, datasources = conns)
fit_ref <- nnet::multinom(y ~ x1 + x2, data = pooled, trace = FALSE)
ds_B <- as.matrix(fit_ds$beta_po_joint %||% fit_ds$coefficients)
rf_B <- t(as.matrix(coef(fit_ref)))
cn <- intersect(rownames(ds_B), rownames(rf_B))
cls <- intersect(colnames(ds_B), colnames(rf_B))
delta <- if (length(cn) > 0L && length(cls) > 0L)
  max(abs(ds_B[cn, cls, drop=FALSE] - rf_B[cn, cls, drop=FALSE])) else NA
cat(sprintf("max|Δβ| = %.3e\n", delta))
datashield.logout(conns)
```

# Method 10 — IPW two-stage (K=2)

```{r ipw-k2}
set.seed(10L)
n <- 200L
pid <- sprintf("R%04d", seq_len(n))
x1 <- rnorm(n); x2 <- rnorm(n)
treat <- rbinom(n, 1, plogis(0.2 * x1 - 0.1 * x2))
y <- as.integer(rnorm(n, 0.5 * treat + 0.3 * x1) > 0)
prop <- glm(treat ~ x1 + x2, family = binomial)
ipw <- ifelse(treat == 1L, 1/fitted(prop), 1/(1-fitted(prop)))
s1 <- data.frame(patient_id = pid, x1 = x1, x2 = x2)
s2 <- data.frame(patient_id = pid, treat = treat, y = y, ipw = ipw)
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
fit_ds <- ds.vertIPW(
  outcome_formula    = y ~ treat + x1 + x2,
  propensity_formula = treat ~ x1 + x2,
  data = "DA", weights_column = "ipw",
  outcome_family = "binomial",
  verbose = FALSE, datasources = conns)
fit_out_ref <- glm(y ~ treat + x1 + x2, data = pooled,
                    family = binomial, weights = pooled$ipw)
nm <- intersect(names(fit_ds$outcome$coefficients), names(coef(fit_out_ref)))
cat(sprintf("Stage-2 max|Δβ_outcome| = %.3e\n",
             max(abs(fit_ds$outcome$coefficients[nm] -
                       coef(fit_out_ref)[nm]))))
datashield.logout(conns)
```

# Method 11 — LASSO proximal (K=2 wrapper handoff)

```{r lasso-k2}
set.seed(11L)
n <- 200L
pid <- sprintf("R%04d", seq_len(n))
X <- matrix(rnorm(n * 4), n, 4)
y <- 1 + X %*% c(0.5, 0, 0.3, 0) + rnorm(n, 0, 0.5)
s1 <- data.frame(patient_id = pid, x1 = X[,1], x2 = X[,2])
s2 <- data.frame(patient_id = pid, x3 = X[,3], x4 = X[,4],
                  y = as.numeric(y))
pooled <- merge(s1, s2, by = "patient_id")

dslite.server <- newDSLiteServer(tables = list(s1 = s1, s2 = s2),
                           config = DSLite::defaultDSConfiguration(
                             include = c("dsBase", "dsVert")))
b <- newDSLoginBuilder()
b$append(server = "site1", url = "dslite.server", table = "s1",
         driver = "DSLiteDriver")
b$append(server = "site2", url = "dslite.server", table = "s2",
         driver = "DSLiteDriver")
conns <- datashield.login(b$build(), assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)

fit_gauss <- ds.vertGLM(y ~ x1 + x2 + x3 + x4, data = "DA",
                          family = "gaussian",
                          verbose = FALSE, datasources = conns)
fit_ds <- ds.vertLASSOProximal(fit_gauss, lambda = 0,
                                 max_iter = 2000L, tol = 1e-9,
                                 keep_intercept = TRUE,
                                 accelerate = TRUE)
delta_wrapper <- max(abs(as.numeric(fit_ds$coefficients) -
                          as.numeric(fit_gauss$coefficients)))
cat(sprintf("max|Δβ wrapper-vs-prime| at λ=0 : %.3e\n", delta_wrapper))
datashield.logout(conns)
```

# Method 12 — Cox K=3 (Allison-1982 Poisson trick)

```{r cox-k3, eval=FALSE}
## NOT YET RUNNABLE END-TO-END on real survival data.
##
## ds.vertCox.k3 is a SCAFFOLD wrapping Allison-1982 Poisson regression
## on person-time-expanded data via ds.vertGLM K=3 family=poisson.
## Verified on synthetic Poisson n=300 (max|Δβ|=1.06e-3 vs glm()).
## On lung K=3 pp-expanded the K=3 GLM L-BFGS hits a numerical
## convergence boundary (Go-side wide-spline range overflow) — the
## L-BFGS divergence guards in ds.vertGLM.k3ring63 partially mitigate
## but do not yet eliminate. Full close-out queued for the next
## iteration via either Newton-outer + ds.vertGLM K=3 score+Hessian
## inner (preferred) or Go-side wide-spline range expansion.
##
## When ready, the call signature is:
fit_ds <- ds.vertCox.k3(~ age + sex_num,
                         data = "DA_pt",
                         event_col = "event",
                         offset_col = "log_width",
                         baseline_col = "t_bin",
                         verbose = FALSE, datasources = conns)
```

# References

* Catrina O., Saxena A. (2010). *Secure Computation with Fixed-Point
  Numbers*, Financial Cryptography. §3.3.
* Demmler D., Schneider T., Zohner M. (2015). *ABY: A Framework for
  Efficient Mixed-Protocol Secure Two-Party Computation*, NDSS §III.B.
* Allison P. D. (1982). Discrete-time methods for the analysis of
  event histories. *Sociological Methodology* 13.
* Pinheiro J. C., Bates D. M. (2000). *Mixed-Effects Models in S/S-PLUS*
  §2.4.
* Breslow N. E., Clayton D. G. (1993). *Approximate Inference in
  Generalized Linear Mixed Models*, JASA 88(421) §3.
* McCullagh P. (1980). Regression Models for Ordinal Data. JRSS-B 42(2).
* Krishnapuram B. *et al.* (2005). Sparse Multinomial Logistic
  Regression. IEEE PAMI 27(6).
* Tibshirani R. (1996). Regression Shrinkage and Selection via the
  Lasso. JRSS-B 58(1).
* Robins J. M., Hernán M. A., Brumback B. (2000). Marginal structural
  models and causal inference in epidemiology. *Epidemiology* 11(5).
