---
title: "Beaver preprocessing profile validation: correlation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Beaver preprocessing profile validation: correlation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "),
                     character(1)),
    stringsAsFactors = FALSE)
}
```

## Scope

This validation exercises the dealer-free `iknp` Beaver preprocessing backend
on the vertically partitioned correlation workflow. `iknp` is the strengthened
OT-extension profile in which the two computation parties derive their triple
shares interactively, so neither party ever holds an unsplit triple.

The route is compared against the same centralized `stats::cor` reference.

## Executed validation

```{r execution}
K <- 2L
tolerance <- 1e-3

pooled <- build_pima(60L)
tables <- list(
  s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
  s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin"), drop = FALSE]
)
variables <- list(
  s1 = c("age", "bmi", "ped"),
  s2 = c("npreg", "glu", "bp", "skin")
)
ref <- stats::cor(pooled[, unlist(variables), drop = FALSE])

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

run_profile <- function(mode) {
  options(dsvert.beaver_preprocessing = mode)
  result <- elapsed({
    conns <- connect_dslite(tables)
    tryCatch({
      psi_align(conns)
      dsVertClient::ds.vert.cor(
        "DA", variables = variables,
        verbose = validation_demo_verbose(), datasources = conns)
    }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
  })
  observed <- max(abs(result$value$correlation[colnames(ref), colnames(ref)] -
                        ref))
  row_result(
    "beaver_profile_correlation", paste("Correlation", mode), K,
    paste0("ds.vert.cor / ", mode),
    "MASS::Pima.tr fixture", "stats::cor",
    "correlation_max_abs_delta", observed, tolerance, "strict-practical",
    "Releases the low-dimensional correlation matrix only.",
    result$runtime_s,
    paste("beaver_preprocessing =", mode))
}

rows <- run_profile("iknp")

display_validation(rows)

assert_validation(rows)
```

## Verdict

Rendering fails if the IKNP route leaves the accepted numerical envelope or if
the route is marked disclosive.
