---
title: "Validation Study: Centralized vs. Federated Vertical Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation Study: Centralized vs. Federated Vertical Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

col_palette <- colorRampPalette(c("#2166AC", "#67A9CF", "#D1E5F0",
                                   "#FDDBC7", "#EF8A62", "#B2182B"))(100)
diff_palette <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
col_local  <- "#4393C3"
col_dsvert <- "#D6604D"
```

## Introduction

This vignette presents a comparative validation of the dsVert federated
vertical analysis pipeline against centralized R computations. We exercise the
full protocol stack --- ECDH-PSI record alignment, MHE-CKKS correlation,
spectral PCA, and encrypted-label GLMs --- on real data vertically partitioned
across three DataSHIELD servers.

The study is organized in three parts:

1. **Federated vertical analysis** --- PSI alignment followed by correlation,
   PCA, and GLMs computed via dsVert, where each server holds a different
   subset of variables and no raw data leaves any server.
2. **Centralized analysis** --- ground truth computed with standard R functions
   on the matched cohort.
3. **Comparative assessment** --- side-by-side numerical comparison with error
   analysis.

### Dataset

We use a modified version of the Pima Indian women diabetes training set
(`Pima.tr`, $n = 200$) from the MASS R package (Venables & Ripley, 2002). It
contains clinical measurements for 200 women of Pima heritage aged $\geq 21$
years, collected by the US National Institute of Diabetes and Digestive and
Kidney Diseases near Phoenix, Arizona (Smith et al., 1988). Seven numeric
variables are available: `npreg` (number of pregnancies), `glu` (plasma
glucose), `bp` (diastolic blood pressure), `skin` (triceps skin fold
thickness), `bmi` (body mass index), `ped` (diabetes pedigree function), and
`age` (years).

### Vertical partitioning scheme

The data are vertically split across three servers, each representing a
different clinical domain. Crucially, **the patient populations are not
identical**: each server holds a different number of patients, with only a
partial overlap. This simulates a realistic scenario where institutions have
data on overlapping but non-identical patient cohorts.

| Server  | Patients | Domain                        | Variables                         |
|---------|----------|-------------------------------|-----------------------------------|
| server1 | 180      | Anthropometric / familial     | `patient_id`, `age`, `bmi`, `ped` |
| server2 | 175      | Glucose and cardiovascular    | `patient_id`, `glu`, `bp`, `skin` |
| server3 | 170      | Reproductive / outcome        | `patient_id`, `npreg`, `diabetes` |

Row order is independently shuffled per server. The ECDH-PSI protocol must
identify the common cohort (155 patients present in all three servers) without
revealing which IDs are shared.

---

## Connect to DataSHIELD and align records

We connect to three independent DataSHIELD sessions, each hosting one vertical
partition on separate Opal instances.

```{r connect, results = "hide", message = FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)

options(opal.verifyssl = FALSE)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "server1", url = "https://localhost:8443",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server2", url = "https://localhost:8444",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")
builder$append(server = "server3", url = "https://localhost:8445",
               user = "administrator", password = "admin123",
               driver = "OpalDriver", table = "pima_project.pima_data")

conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

### PSI Record Alignment

Each server only knows its own patient identifiers. The ECDH-PSI protocol
finds the common cohort using elliptic curve Diffie-Hellman blind relay,
without revealing which IDs are shared or which patients are unique to each
server.

```{r psi, message = FALSE}
psi_result <- ds.psiAlign(
  data_name   = "D",
  id_col      = "patient_id",
  newobj      = "D_aligned",
  datasources = conns
)
```

The PSI protocol identified **`r psi_result$n_intersect`** patients present
in all three servers. The matched cohort is now row-aligned across servers:
each row index corresponds to the same patient, enabling cross-server
computations.

---

## Centralized analysis (ground truth)

We load the matched ground truth dataset and compute all analyses with
standard R functions. These serve as the reference baseline.

```{r ground-truth}
demo_dir <- normalizePath(file.path(
  system.file(package = "dsVertClient"), "..", "..", "demo"
), mustWork = FALSE)
if (!dir.exists(demo_dir))
  demo_dir <- normalizePath("../../demo", mustWork = FALSE)
if (!dir.exists(demo_dir))
  demo_dir <- Sys.getenv("DSVERT_DEMO_DIR",
    unset = "/Users/david/Documents/GitHub/dsvert-paper/demo")

full <- merge(merge(
  read.csv(file.path(demo_dir, "pima_server1.csv")),
  read.csv(file.path(demo_dir, "pima_server2.csv")), by = "patient_id"),
  read.csv(file.path(demo_dir, "pima_server3.csv")), by = "patient_id")
# Keep only the matched cohort
full <- full[order(full$patient_id), ]
num_vars <- c("age", "bmi", "ped", "glu", "bp", "skin", "npreg")
cat("Ground truth dataset:", nrow(full), "rows x", ncol(full), "columns\n")
```

### Pearson correlation

```{r local-cor}
local_cor <- cor(full[, num_vars])
```

```{r local-cor-print, echo = FALSE}
print(round(local_cor, 4))
```

```{r local-cor-heatmap, echo = FALSE, fig.height = 5.5}
heatmap(local_cor, symm = TRUE, col = col_palette, margins = c(8, 8),
        Rowv = NA, Colv = NA,
        main = "Centralized Pearson Correlation")
```

### PCA (spectral decomposition)

```{r local-pca}
local_pca <- eigen(local_cor)
local_eigenvalues <- local_pca$values
local_loadings    <- local_pca$vectors
rownames(local_loadings) <- num_vars
colnames(local_loadings) <- paste0("PC", seq_along(num_vars))
local_ve <- local_eigenvalues / sum(local_eigenvalues) * 100
```

```{r local-pca-table, echo = FALSE}
pca_df <- data.frame(
  Component  = paste0("PC", 1:7),
  Eigenvalue = round(local_eigenvalues, 4),
  Variance   = paste0(round(local_ve, 1), "%"),
  Cumulative = paste0(round(cumsum(local_ve), 1), "%")
)
knitr::kable(pca_df, align = c("l", "r", "r", "r"),
             caption = "Centralized PCA: eigenvalues and variance explained")
```

```{r local-scree, echo = FALSE, fig.height = 4}
barplot(local_ve, names.arg = paste0("PC", 1:7),
        ylab = "Variance explained (%)", col = col_local,
        main = "Centralized PCA: Scree Plot", border = NA, las = 1)
```

### GLM --- Gaussian (y = `glu`)

```{r local-gauss}
local_gauss <- glm(glu ~ age + bmi + ped + bp + skin + npreg + diabetes,
                   data = full, family = gaussian)
```

```{r local-gauss-print, echo = FALSE}
summary(local_gauss)
```

### GLM --- Binomial (y = `diabetes`)

```{r local-binom}
local_binom <- glm(diabetes ~ age + bmi + ped + glu + bp + skin + npreg,
                   data = full, family = binomial)
```

```{r local-binom-print, echo = FALSE}
summary(local_binom)
```

### GLM --- Poisson (y = `npreg`)

```{r local-pois}
local_pois <- glm(npreg ~ age + bmi + ped + glu + bp + skin + diabetes,
                  data = full, family = poisson)
```

```{r local-pois-print, echo = FALSE}
summary(local_pois)
```

---

## Federated vertical analysis (dsVert)

We now repeat every analysis using the dsVert federated protocol. Each server
only has access to its own subset of variables. Cross-server quantities (such
as correlation matrix off-diagonal blocks, and gradients involving variables
from multiple servers) are computed under multiparty homomorphic encryption
(MHE-CKKS) with threshold decryption, so individual-level data never leaves
any server.

### Correlation (MHE-CKKS)

The `ds.vertCor` function computes the full $7 \times 7$ Pearson correlation
matrix. Within-server blocks are computed locally; cross-server blocks use
CKKS encrypted dot products with threshold decryption and server-side fusion.

```{r dsvert-cor, message = FALSE}
variables <- list(
  server1 = c("age", "bmi", "ped"),
  server2 = c("glu", "bp", "skin"),
  server3 = c("npreg")
)

cor_result <- ds.vertCor(
  data_name   = "D_aligned",
  variables   = variables,
  log_n       = 12,
  log_scale   = 40,
  datasources = conns
)

dsvert_cor <- cor_result$correlation
```

```{r dsvert-cor-print, echo = FALSE}
print(round(dsvert_cor, 4))
```

```{r dsvert-cor-heatmap, echo = FALSE, fig.height = 5.5}
heatmap(dsvert_cor, symm = TRUE, col = col_palette, margins = c(8, 8),
        Rowv = NA, Colv = NA,
        main = "dsVert Federated Pearson Correlation (MHE-CKKS)")
```

### PCA

PCA is computed by eigendecomposition of the federated correlation matrix.

```{r dsvert-pca}
dsvert_pca <- eigen(dsvert_cor)
dsvert_eigenvalues <- dsvert_pca$values
dsvert_loadings    <- dsvert_pca$vectors
rownames(dsvert_loadings) <- num_vars
colnames(dsvert_loadings) <- paste0("PC", seq_along(num_vars))
dsvert_ve <- dsvert_eigenvalues / sum(dsvert_eigenvalues) * 100
```

```{r dsvert-pca-table, echo = FALSE}
pca_df2 <- data.frame(
  Component  = paste0("PC", 1:7),
  Eigenvalue = round(dsvert_eigenvalues, 4),
  Variance   = paste0(round(dsvert_ve, 1), "%"),
  Cumulative = paste0(round(cumsum(dsvert_ve), 1), "%")
)
knitr::kable(pca_df2, align = c("l", "r", "r", "r"),
             caption = "dsVert PCA: eigenvalues and variance explained")
```

```{r dsvert-scree, echo = FALSE, fig.height = 4}
barplot(dsvert_ve, names.arg = paste0("PC", 1:7),
        ylab = "Variance explained (%)", col = col_dsvert,
        main = "dsVert PCA: Scree Plot", border = NA, las = 1)
```

### GLM --- Gaussian (y = `glu`)

The dsVert GLM uses encrypted-label block coordinate descent (BCD-IRLS).
The label variable (`glu`) resides on server2; the remaining predictors are
distributed across all three servers. Gradients are computed under MHE-CKKS
and decrypted via share-wrapped threshold decryption with server-side fusion.

```{r dsvert-gauss, message = FALSE}
glm_gauss <- ds.vertGLM(
  data_name = "D_aligned", y_var = "glu",
  x_vars    = list(server1 = c("age", "bmi", "ped"),
                   server2 = c("bp", "skin"),
                   server3 = c("npreg", "diabetes")),
  y_server  = "server2", family = "gaussian",
  max_iter = 100, tol = 1e-4, lambda = 1e-4,
  log_n = 12, log_scale = 40,
  datasources = conns
)
```

```{r dsvert-gauss-print, echo = FALSE}
summary(glm_gauss)
```

### GLM --- Binomial (y = `diabetes`)

Logistic regression with the binary outcome on server3. The IRLS working
weights are computed on the label server and routed to non-label servers via
transport encryption.

```{r dsvert-binom, message = FALSE}
glm_binom <- ds.vertGLM(
  data_name = "D_aligned", y_var = "diabetes",
  x_vars    = list(server1 = c("age", "bmi", "ped"),
                   server2 = c("glu", "bp", "skin"),
                   server3 = c("npreg")),
  y_server  = "server3", family = "binomial",
  max_iter = 100, tol = 1e-4, lambda = 1e-4,
  log_n = 12, log_scale = 40,
  datasources = conns
)
```

```{r dsvert-binom-print, echo = FALSE}
summary(glm_binom)
```

### GLM --- Poisson (y = `npreg`)

Count regression for the number of pregnancies. The Poisson log-link compresses
the coefficient scale, which tends to aid convergence of the encrypted BCD
procedure.

```{r dsvert-pois, message = FALSE}
glm_pois <- ds.vertGLM(
  data_name = "D_aligned", y_var = "npreg",
  x_vars    = list(server1 = c("age", "bmi", "ped"),
                   server2 = c("glu", "bp", "skin"),
                   server3 = c("diabetes")),
  y_server  = "server3", family = "poisson",
  max_iter = 100, tol = 1e-4, lambda = 1e-4,
  log_n = 12, log_scale = 40,
  datasources = conns
)
```

```{r dsvert-pois-print, echo = FALSE}
summary(glm_pois)
```

---

## Comparative assessment

We now compare the centralized and federated results quantitatively.

### Correlation: element-wise differences

```{r cmp-cor-diff}
cor_diff <- dsvert_cor - local_cor
```

```{r cmp-cor-diff-print, echo = FALSE}
print(round(cor_diff, 7))
```

```{r cmp-cor-heatmap, echo = FALSE, fig.height = 5.5}
max_abs <- max(abs(cor_diff))
image(seq_along(num_vars), seq_along(num_vars), cor_diff,
      col = diff_palette, zlim = c(-max_abs, max_abs),
      xlab = "", ylab = "", axes = FALSE,
      main = paste0("Correlation Difference (dsVert - Centralized)\nmax |diff| = ",
                     format(max_abs, digits = 3, scientific = TRUE)))
axis(1, at = seq_along(num_vars), labels = num_vars, las = 2)
axis(2, at = seq_along(num_vars), labels = num_vars, las = 1)
for (i in seq_along(num_vars)) for (j in seq_along(num_vars))
  text(i, j, format(cor_diff[i, j], digits = 2, scientific = TRUE), cex = 0.6)
```

All element-wise differences are of order $10^{-6}$, well below any
substantively meaningful threshold.

### PCA: eigenvalue comparison

```{r cmp-pca}
eig_diff <- dsvert_eigenvalues - local_eigenvalues
```

```{r cmp-pca-table, echo = FALSE}
eig_cmp <- data.frame(
  Component  = paste0("PC", 1:7),
  Local      = round(local_eigenvalues, 6),
  dsVert     = round(dsvert_eigenvalues, 6),
  Difference = format(eig_diff, digits = 3, scientific = TRUE)
)
knitr::kable(eig_cmp, align = c("l", "r", "r", "r"),
             caption = "Eigenvalue comparison (centralized vs. dsVert)")
```

```{r cmp-pca-plot, echo = FALSE, fig.height = 4.5}
x <- barplot(rbind(local_eigenvalues, dsvert_eigenvalues),
             beside = TRUE, names.arg = paste0("PC", 1:7),
             col = c(col_local, col_dsvert), border = NA, las = 1,
             ylab = "Eigenvalue",
             main = "PCA Eigenvalue Comparison")
legend("topright", legend = c("Centralized", "dsVert"),
       fill = c(col_local, col_dsvert), border = NA,
       bty = "o", bg = "white", box.col = "black")
```

The eigenvalues are virtually identical; differences are of order $10^{-5}$,
propagated directly from the correlation matrix precision.

### GLM: Gaussian coefficient comparison (y = `glu`)

```{r cmp-gauss-setup}
gauss_local  <- coef(local_gauss)
gauss_dsvert <- coef(glm_gauss)
gauss_diff   <- gauss_dsvert - gauss_local
```

```{r cmp-gauss-table, echo = FALSE}
knitr::kable(
  data.frame(Term = names(gauss_local),
             Local = round(gauss_local, 4),
             dsVert = round(gauss_dsvert, 4),
             Difference = format(gauss_diff, digits = 3, scientific = TRUE)),
  row.names = FALSE, align = c("l", "r", "r", "r"),
  caption = "Gaussian GLM coefficient comparison (y = glu)")
```

```{r cmp-gauss-plot, echo = FALSE, fig.width = 8, fig.height = 5}
idx <- seq_along(gauss_local)
par(mar = c(5, 4, 3, 8))
plot(idx - 0.1, gauss_local, pch = 16, col = col_local, cex = 1.5,
     xlab = "", ylab = "Coefficient", main = "Gaussian GLM (y = glu)",
     xaxt = "n", xlim = c(0.5, length(idx) + 0.5),
     ylim = range(c(gauss_local, gauss_dsvert)) * 1.1)
points(idx + 0.1, gauss_dsvert, pch = 17, col = col_dsvert, cex = 1.5)
segments(idx - 0.1, gauss_local, idx + 0.1, gauss_dsvert,
         col = "grey60", lty = 2)
axis(1, at = idx, labels = names(gauss_local), las = 1)
abline(h = 0, col = "grey80")
legend(par("usr")[2] + 0.4, par("usr")[4],
       legend = c("Centralized", "dsVert"),
       pch = c(16, 17), col = c(col_local, col_dsvert),
       bty = "o", bg = "white", box.col = "black", xpd = TRUE)
```

Maximum absolute coefficient difference:
`r format(max(abs(gauss_diff)), digits = 3, scientific = TRUE)`.

### GLM: Binomial coefficient comparison (y = `diabetes`)

```{r cmp-binom-setup}
binom_local  <- coef(local_binom)
binom_dsvert <- coef(glm_binom)
binom_diff   <- binom_dsvert - binom_local
```

```{r cmp-binom-table, echo = FALSE}
knitr::kable(
  data.frame(Term = names(binom_local),
             Local = round(binom_local, 4),
             dsVert = round(binom_dsvert, 4),
             Difference = format(binom_diff, digits = 3, scientific = TRUE)),
  row.names = FALSE, align = c("l", "r", "r", "r"),
  caption = "Binomial GLM coefficient comparison (y = diabetes)")
```

```{r cmp-binom-plot, echo = FALSE, fig.width = 8, fig.height = 5}
idx <- seq_along(binom_local)
par(mar = c(5, 4, 3, 8))
plot(idx - 0.1, binom_local, pch = 16, col = col_local, cex = 1.5,
     xlab = "", ylab = "Coefficient", main = "Binomial GLM (y = diabetes)",
     xaxt = "n", xlim = c(0.5, length(idx) + 0.5),
     ylim = range(c(binom_local, binom_dsvert)) * 1.1)
points(idx + 0.1, binom_dsvert, pch = 17, col = col_dsvert, cex = 1.5)
segments(idx - 0.1, binom_local, idx + 0.1, binom_dsvert,
         col = "grey60", lty = 2)
axis(1, at = idx, labels = names(binom_local), las = 1)
abline(h = 0, col = "grey80")
legend(par("usr")[2] + 0.4, par("usr")[4],
       legend = c("Centralized", "dsVert"),
       pch = c(16, 17), col = c(col_local, col_dsvert),
       bty = "o", bg = "white", box.col = "black", xpd = TRUE)
```

Maximum absolute coefficient difference:
`r format(max(abs(binom_diff)), digits = 3, scientific = TRUE)`.

### GLM: Poisson coefficient comparison (y = `npreg`)

```{r cmp-pois-setup}
pois_local  <- coef(local_pois)
pois_dsvert <- coef(glm_pois)
pois_diff   <- pois_dsvert - pois_local
```

```{r cmp-pois-table, echo = FALSE}
knitr::kable(
  data.frame(Term = names(pois_local),
             Local = round(pois_local, 4),
             dsVert = round(pois_dsvert, 4),
             Difference = format(pois_diff, digits = 3, scientific = TRUE)),
  row.names = FALSE, align = c("l", "r", "r", "r"),
  caption = "Poisson GLM coefficient comparison (y = npreg)")
```

```{r cmp-pois-plot, echo = FALSE, fig.width = 8, fig.height = 5}
idx <- seq_along(pois_local)
par(mar = c(5, 4, 3, 8))
plot(idx - 0.1, pois_local, pch = 16, col = col_local, cex = 1.5,
     xlab = "", ylab = "Coefficient", main = "Poisson GLM (y = npreg)",
     xaxt = "n", xlim = c(0.5, length(idx) + 0.5),
     ylim = range(c(pois_local, pois_dsvert)) * 1.1)
points(idx + 0.1, pois_dsvert, pch = 17, col = col_dsvert, cex = 1.5)
segments(idx - 0.1, pois_local, idx + 0.1, pois_dsvert,
         col = "grey60", lty = 2)
axis(1, at = idx, labels = names(pois_local), las = 1)
abline(h = 0, col = "grey80")
legend(par("usr")[2] + 0.4, par("usr")[4],
       legend = c("Centralized", "dsVert"),
       pch = c(16, 17), col = c(col_local, col_dsvert),
       bty = "o", bg = "white", box.col = "black", xpd = TRUE)
```

Maximum absolute coefficient difference:
`r format(max(abs(pois_diff)), digits = 3, scientific = TRUE)`.

### Summary

```{r cmp-summary, echo = FALSE}
knitr::kable(
  data.frame(
    Protocol = c("PSI Alignment", "Correlation", "PCA",
                 "GLM Gaussian", "GLM Binomial", "GLM Poisson"),
    Metric = c("Matched cohort",
               "Max |r| error",
               "Max eigenvalue error",
               "Max |coef| error",
               "Max |coef| error",
               "Max |coef| error"),
    Value = c(paste0(psi_result$n_intersect, " / 200"),
              format(max(abs(cor_diff)), digits = 3, scientific = TRUE),
              format(max(abs(eig_diff)), digits = 3, scientific = TRUE),
              format(max(abs(gauss_diff)), digits = 3, scientific = TRUE),
              format(max(abs(binom_diff)), digits = 3, scientific = TRUE),
              format(max(abs(pois_diff)), digits = 3, scientific = TRUE))
  ),
  align = c("l", "l", "r"),
  caption = "Validation summary (dsVert vs. centralized R)"
)
```

### Error analysis

The three error regimes observed correspond to different sources of
approximation in the dsVert pipeline:

**Correlation and PCA** ($\sim 10^{-6}$). These errors arise from CKKS
approximate arithmetic used in the MHE threshold protocol. The encoding
precision is governed by `log_scale = 40`, providing approximately $2^{40}
\approx 10^{12}$ precision units. After encoding, homomorphic multiplication,
inner-sum rotations, and threshold decryption, the effective precision degrades
to $\sim 10^{-6}$, consistent with CKKS error propagation theory.

**GLM coefficients** ($10^{-4}$ to $10^{-3}$). GLM errors are slightly larger
because they accumulate over multiple BCD-IRLS iterations, each involving an
encrypted gradient computation. Additionally, the ridge regularization parameter
($\lambda = 10^{-4}$) introduces a small systematic bias relative to the
unpenalized centralized `glm()`. The binomial and Poisson families show smaller
errors than Gaussian because their link functions (logit, log) compress the
coefficient scale.

In all cases, the errors are far below any practically meaningful threshold for
statistical inference.

```{r disconnect, include = FALSE}
DSI::datashield.logout(conns)
```
