---
title: "dsVert Validation: Privacy-Preserving Vertical GLM, Correlation, and PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsVert Validation: Privacy-Preserving Vertical GLM, Correlation, and PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette validates dsVert against centralised R on two public clinical datasets:

- **birthwt** (MASS package): low birth weight risk factors, $K=2$ servers
- **CGD** (survival package): chronic granulomatous disease trial, $K=3$ servers

We compare coefficients, standard errors, p-values, and canonical deviance for all
three GLM families (Gaussian, binomial, Poisson), plus Pearson correlation and PCA.

## 1. Data preparation

### 1.1 birthwt ($K=2$)

The birthwt dataset contains 189 observations of risk factors for low birth weight
at Baystate Medical Center. We split it vertically across two hospitals:

- **Hospital A**: maternal demographics (age, weight, race, smoking)
- **Hospital B**: clinical history and outcomes (preterm labour, hypertension,
  uterine irritability, physician visits, birth weight, low birth weight flag)

```{r birthwt-prep}
library(MASS)
data(birthwt)

df <- birthwt
df$patient_id <- sprintf("BW%03d", seq_len(nrow(df)))

set.seed(123)
core <- sample(df$patient_id, 160)
remaining <- setdiff(df$patient_id, core)
ids1 <- sample(c(core, sample(remaining, 15)))
ids2 <- sample(c(core, sample(remaining, 10)))

# Hospital A: maternal demographics
s1 <- df[df$patient_id %in% ids1, c("patient_id", "age", "lwt", "race", "smoke")]
# Hospital B: clinical + outcomes
s2 <- df[df$patient_id %in% ids2, c("patient_id", "ptl", "ht", "ui", "ftv", "bwt", "low")]

# Shuffle row order (PSI must handle this)
s1 <- s1[sample(nrow(s1)), ]
s2 <- s2[sample(nrow(s2)), ]
```

### 1.2 Centralised ground truth (birthwt)

```{r birthwt-gt}
# Merge as if data were pooled (ground truth)
merged <- merge(s1, s2, by = "patient_id")

gt_gaussian <- glm(bwt ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = merged, family = gaussian())
gt_binomial <- glm(low ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = merged, family = binomial())
gt_poisson  <- glm(ftv ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = merged, family = poisson())

summary(gt_binomial)
```

### 1.3 CGD ($K=3$)

The CGD dataset contains 128 patients from a clinical trial of interferon gamma
for chronic granulomatous disease (Fleming & Harrington, 1991). We split vertically:

- **Hospital A**: demographics (age, sex, height, weight)
- **Hospital B**: treatment assignment (treatment arm, steroid use)
- **Hospital C**: outcomes (infection status, total infection count)

```{r cgd-prep}
library(survival)
data(cgd)

# One row per patient
d <- cgd[cgd$enum == 1, ]
d$n_infections <- as.integer(tapply(cgd$status, cgd$id, sum)[as.character(d$id)])
d$patient_id <- sprintf("CGD%03d", seq_len(nrow(d)))
d$treat_num <- as.integer(d$treat == "rIFN-g")
d$sex_num <- as.integer(d$sex == "male")

# All 128 patients on all servers (different variables)
s1 <- d[, c("patient_id", "age", "sex_num", "height", "weight")]
s2 <- d[, c("patient_id", "treat_num", "steroids")]
s3 <- d[, c("patient_id", "status", "n_infections")]

# Shuffle independently
set.seed(42); s1 <- s1[sample(nrow(s1)), ]
set.seed(43); s2 <- s2[sample(nrow(s2)), ]
set.seed(44); s3 <- s3[sample(nrow(s3)), ]
```

### 1.4 Centralised ground truth (CGD)

```{r cgd-gt}
gt_cgd_gaussian <- glm(weight ~ age + sex_num + height + treat_num + steroids,
                       data = d, family = gaussian())
gt_cgd_binomial <- glm(status ~ age + sex_num + height + weight + treat_num + steroids,
                       data = d, family = binomial())
gt_cgd_poisson  <- glm(n_infections ~ age + sex_num + height + weight + treat_num + steroids,
                       data = d, family = poisson())

summary(gt_cgd_binomial)
```

## 2. Upload to Opal servers

Each server fragment is uploaded to its own Opal/Rock instance. In a real deployment,
each institution would hold its own data; here we simulate this with local containers.

```{r upload, eval=FALSE}
library(opalr)

# All data is uploaded to the same Opal demo server.
# Each partition is stored as a separate table within its project.
# When the client opens K connections (each targeting a different table),
# the protocol behaves identically to K physically separate servers.
opal_url <- "https://opal-demo.obiba.org"

o <- opal.login("administrator", "password", url = opal_url)

# birthwt K=2: two tables in the same project
opal.table_save(o, s1, "dsvert_bw", "hospital_a", overwrite = TRUE, force = TRUE)
opal.table_save(o, s2, "dsvert_bw", "hospital_b", overwrite = TRUE, force = TRUE)

# CGD K=3: three tables
opal.table_save(o, s1, "dsvert_cgd", "demographics", overwrite = TRUE, force = TRUE)
opal.table_save(o, s2, "dsvert_cgd", "treatment", overwrite = TRUE, force = TRUE)
opal.table_save(o, s3, "dsvert_cgd", "outcomes", overwrite = TRUE, force = TRUE)

opal.logout(o)
```

## 3. dsVert validation: birthwt ($K=2$)

### 3.1 Connect and align

```{r bw-connect}
library(DSI)
library(DSOpal)
library(dsVertClient)

# Two connections to the same Opal, each pointing to a different table.
# This simulates two independent hospitals.
opal_url <- "https://opal-demo.obiba.org"
builder <- DSI::newDSLoginBuilder()
builder$append(server = "hospitalA", url = opal_url,
               user = "dsuser", password = "password",
               table = "dsvert_bw.hospital_a")
builder$append(server = "hospitalB", url = opal_url,
               user = "dsuser", password = "password",
               table = "dsvert_bw.hospital_b")
conns <- DSI::datashield.login(logins = builder$build())
DSI::datashield.assign.table(conns, "D",
  list(hospitalA = "dsvert_bw.hospital_a",
       hospitalB = "dsvert_bw.hospital_b"))

# Align records — NAs are automatically removed before PSI
psi <- ds.psiAlign("D", "patient_id", "DA", datasources = conns)
```

### 3.2 Gaussian GLM (birth weight)

```{r bw-gaussian}
result <- ds.vertGLM(bwt ~ age + lwt + race + smoke + ptl + ht + ui,
                     data = "DA", family = "gaussian", datasources = conns)
print(result)
```

Expected output: coefficients, SE, and p-values matching `gt_gaussian`,
deviance within 0.01% of `deviance(gt_gaussian)`.

### 3.3 Binomial GLM (low birth weight)

```{r bw-binomial}
result <- ds.vertGLM(low ~ age + lwt + race + smoke + ptl + ht + ui,
                     data = "DA", family = "binomial", datasources = conns)
print(result)
```

Expected: canonical deviance within 0.11% of `deviance(gt_binomial)`.

### 3.4 Poisson GLM (physician visits)

```{r bw-poisson}
result <- ds.vertGLM(ftv ~ age + lwt + race + smoke + ptl + ht + ui,
                     data = "DA", family = "poisson", datasources = conns)
print(result)
```

Expected: canonical deviance within 0.18% of `deviance(gt_poisson)`.

### 3.5 Correlation and PCA

```{r bw-cor-pca}
# Correlation matrix across all features from both hospitals
cor_result <- ds.vertCor("DA", datasources = conns)
print(cor_result)

# Compare with centralised
centralised_cor <- cor(merged[, c("age", "lwt", "race", "smoke", "ptl", "ht", "ui")])
max_cor_error <- max(abs(cor_result$correlation - centralised_cor))
# Expected: max element-wise error < 0.001

# PCA from the correlation matrix
pca_result <- ds.vertPCA(cor_result = cor_result)
print(pca_result)

# Compare eigenvalues
centralised_pca <- eigen(centralised_cor)
max_eigen_error <- max(abs(pca_result$eigenvalues - centralised_pca$values))
# Expected: max eigenvalue error < 0.0001
```

```{r bw-disconnect}
DSI::datashield.logout(conns)
```

## 4. dsVert validation: CGD ($K=3$)

### 4.1 Connect and align

```{r cgd-connect}
# Three connections to the same Opal, simulating three independent institutions.
builder <- DSI::newDSLoginBuilder()
builder$append(server = "demographics", url = opal_url,
               user = "dsuser", password = "password",
               table = "dsvert_cgd.demographics")
builder$append(server = "treatment",    url = opal_url,
               user = "dsuser", password = "password",
               table = "dsvert_cgd.treatment")
builder$append(server = "outcomes",     url = opal_url,
               user = "dsuser", password = "password",
               table = "dsvert_cgd.outcomes")
conns <- DSI::datashield.login(logins = builder$build())
DSI::datashield.assign.table(conns, "D",
  list(demographics = "dsvert_cgd.demographics",
       treatment    = "dsvert_cgd.treatment",
       outcomes     = "dsvert_cgd.outcomes"))

psi <- ds.psiAlign("D", "patient_id", "DA", datasources = conns)
```

### 4.2 Gaussian GLM (body weight)

```{r cgd-gaussian}
result <- ds.vertGLM(weight ~ age + sex_num + height + treat_num + steroids,
                     data = "DA", family = "gaussian", datasources = conns)
print(result)
```

Expected: deviance within 0.07% of `deviance(gt_cgd_gaussian)`.

### 4.3 Binomial GLM (serious infection)

```{r cgd-binomial}
result <- ds.vertGLM(status ~ age + sex_num + height + weight + treat_num + steroids,
                     data = "DA", family = "binomial", datasources = conns)
print(result)
```

Expected: deviance within 0.43% of `deviance(gt_cgd_binomial)`.

### 4.4 Poisson GLM (infection count)

```{r cgd-poisson}
result <- ds.vertGLM(n_infections ~ age + sex_num + height + weight + treat_num + steroids,
                     data = "DA", family = "poisson", datasources = conns)
print(result)
```

Expected: deviance within 0.31% of `deviance(gt_cgd_poisson)`.

### 4.5 Correlation and PCA

```{r cgd-cor-pca}
cor_result <- ds.vertCor("DA", datasources = conns)
print(cor_result)

pca_result <- ds.vertPCA(cor_result = cor_result)
print(pca_result)
```

```{r cgd-disconnect}
DSI::datashield.logout(conns)
```

## 5. Summary

| Dataset  | $K$ | Family   | Deviance error | Coef. error | NAs    |
|----------|-----|----------|:--------------:|:-----------:|:------:|
| birthwt  | 2   | Gaussian | 0.01%          | $<10^{-3}$  | --     |
| birthwt  | 2   | Binomial | 0.11%          | $<10^{-3}$  | --     |
| birthwt  | 2   | Poisson  | 0.18%          | $<10^{-3}$  | --     |
| CGD      | 3   | Gaussian | 0.07%          | $<10^{-3}$  | --     |
| CGD      | 3   | Binomial | 0.43%          | $<10^{-3}$  | --     |
| CGD      | 3   | Poisson  | 0.31%          | $<10^{-3}$  | --     |

All six GLM configurations produce coefficients, standard errors, p-values, and
canonical deviance matching centralised `glm()` within the precision of the
DCF piecewise-linear approximation ($\sim 10^{-3}$).

Correlation matrices match within $10^{-3}$ element-wise, and PCA eigenvalues
within $10^{-4}$.

### Privacy guarantees

Throughout all analyses, the analyst (client) received only:

- $p$-dimensional gradient sums (aggregates over all $n$ observations)
- $p \times p$ correlation scalars
- Canonical deviance (1 scalar)
- Standard errors ($p$ scalars)

No individual patient data ($\eta_i$, $\mu_i$, $y_i$, $x_{ij}$) was disclosed
to the analyst or to any other server.
