---
title: "dsVert Validation: Federated Vertical vs Centralised"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsVert Validation: Federated Vertical vs Centralised}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(DSI)
library(DSOpal)
library(dsVertClient)
library(MASS)
library(survival)
```

dsVert runs generalised linear models, correlation, and PCA across hospitals
that each hold **different variables** for the same patients — without any
hospital sharing its raw data. This vignette runs a full analysis on a public
demo server and compares every result against standard centralised R.

## The demo server

All data lives on [opal-demo.obiba.org](https://opal-demo.obiba.org), the
public DataSHIELD demo maintained by OBiBa. The datasets are stored in the
`vert_demo` project. We open multiple connections to the same Opal — each
pointing to a different table — which behaves identically to connecting to
physically separate hospitals.

## Dataset 1: birthwt (K=2 hospitals)

The `birthwt` dataset from the MASS package contains 189 observations of risk
factors for low birth weight. It has been vertically partitioned into two
hospitals:

- **Hospital A** holds maternal demographics: age, weight, race, smoking status
- **Hospital B** holds clinical history and outcomes: preterm labour, hypertension,
  uterine irritability, physician visits, birth weight, low-birth-weight flag

### What each hospital sees

```{r bw-connect, message=FALSE}
opal_url <- "https://opal-demo.obiba.org"

builder <- DSI::newDSLoginBuilder()
builder$append(server = "hospitalA", url = opal_url,
               user = "administrator", password = "password",
               table = "vert_demo.bw_hospital_a")
builder$append(server = "hospitalB", url = opal_url,
               user = "administrator", password = "password",
               table = "vert_demo.bw_hospital_b")
bw_conns <- DSI::datashield.login(logins = builder$build())
DSI::datashield.assign.table(bw_conns, "D",
  list(hospitalA = "vert_demo.bw_hospital_a",
       hospitalB = "vert_demo.bw_hospital_b"))
```

Each hospital holds different columns. We can verify this without seeing the
actual patient data:

```{r bw-columns}
DSI::datashield.aggregate(bw_conns, call("dsvertColNamesDS", data_name = "D"))
```

### Align patients across hospitals

Hospitals have overlapping but not identical patient registries. Private Set
Intersection finds the common patients without revealing who is unique to each
site:

```{r bw-psi}
bw_psi <- ds.psiAlign("D", "patient_id", "DA", verbose = FALSE, datasources = bw_conns)
bw_psi$n_common
```

### Building the centralised equivalent

To verify dsVert's results, we first reconstruct the exact same data locally.
The `birthwt` dataset is publicly available in the MASS package. We partition
it using the same procedure and seed that was used to create the server data:

```{r bw-centralised-data}
data(birthwt)
bw <- birthwt
bw$patient_id <- paste0("BW", formatC(seq_len(nrow(bw)), width = 3, flag = "0"))

set.seed(123)
core <- sample(bw$patient_id, 160)
remaining <- setdiff(bw$patient_id, core)
ids_a <- sample(c(core, sample(remaining, 15)))
ids_b <- sample(c(core, sample(remaining, 10)))

# These two fragments are what each hospital holds:
hospital_a <- bw[bw$patient_id %in% ids_a, c("patient_id", "age", "lwt", "race", "smoke")]
hospital_b <- bw[bw$patient_id %in% ids_b, c("patient_id", "ptl", "ht", "ui", "ftv", "bwt", "low")]
```

Hospital A has `r nrow(hospital_a)` patients with demographics:

```{r bw-head-a}
head(hospital_a)
```

Hospital B has `r nrow(hospital_b)` patients with clinical data and outcomes:

```{r bw-head-b}
head(hospital_b)
```

In a centralised scenario, we could simply merge on patient ID — producing the
full dataset that no single hospital can see:

```{r bw-merge}
bw_merged <- merge(hospital_a, hospital_b, by = "patient_id")
head(bw_merged)
```

The merge yields `r nrow(bw_merged)` patients — the same intersection that the
privacy-preserving PSI found above (`r bw_psi$n_common`).

### Centralised GLM (ground truth)

```{r bw-centralised-binom}
gt_bw_binom <- glm(low ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = bw_merged, family = binomial())
summary(gt_bw_binom)
```

### Federated vertical GLM (dsVert)

Now the same model, but computed across the two hospitals without merging the
data. The analyst writes a standard R formula — dsVert detects that `low` is on
Hospital B and the predictors are split across both:

```{r bw-binomial}
dv_bw_binom <- ds.vertGLM(low ~ age + lwt + race + smoke + ptl + ht + ui,
                           data = "DA", family = "binomial",
                           verbose = FALSE, datasources = bw_conns)
dv_bw_binom
```

### Compare coefficients

```{r bw-binom-compare, fig.width=7, fig.height=5}
nms <- names(dv_bw_binom$coefficients)
comp <- data.frame(
  variable    = nms,
  centralised = coef(gt_bw_binom)[nms],
  dsVert      = dv_bw_binom$coefficients[nms],
  row.names   = NULL
)
comp$difference <- comp$dsVert - comp$centralised

# Visual comparison
par(mar = c(4, 8, 2, 2))
idx <- seq_len(nrow(comp))
plot(comp$centralised, idx, pch = 16, col = "black", xlim = range(c(comp$centralised, comp$dsVert)),
     yaxt = "n", ylab = "", xlab = "Coefficient value",
     main = "Binomial GLM: centralised vs dsVert (birthwt)")
points(comp$dsVert, idx, pch = 17, col = "red")
axis(2, at = idx, labels = comp$variable, las = 1)
segments(comp$centralised, idx, comp$dsVert, idx, col = "grey70", lty = 2)
legend("topright", legend = c("Centralised", "dsVert"), pch = c(16, 17), col = c("black", "red"))
```

```{r bw-binom-table}
comp
```

### All three families

```{r bw-all-families}
dv_bw_gauss <- ds.vertGLM(bwt ~ age + lwt + race + smoke + ptl + ht + ui,
                           data = "DA", family = "gaussian",
                           verbose = FALSE, datasources = bw_conns)

dv_bw_pois  <- ds.vertGLM(ftv ~ age + lwt + race + smoke + ptl + ht + ui,
                           data = "DA", family = "poisson",
                           verbose = FALSE, datasources = bw_conns)

gt_bw_gauss <- glm(bwt ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = bw_merged, family = gaussian())
gt_bw_pois  <- glm(ftv ~ age + lwt + race + smoke + ptl + ht + ui,
                   data = bw_merged, family = poisson())
```

### Deviance comparison

```{r bw-deviance}
bw_dev <- data.frame(
  family      = c("Gaussian", "Binomial", "Poisson"),
  centralised = round(c(deviance(gt_bw_gauss), deviance(gt_bw_binom), deviance(gt_bw_pois)), 2),
  dsVert      = round(c(dv_bw_gauss$deviance, dv_bw_binom$deviance, dv_bw_pois$deviance), 2)
)
bw_dev$error_pct <- round(100 * abs(bw_dev$dsVert - bw_dev$centralised) / bw_dev$centralised, 2)
bw_dev
```

### Correlation and PCA

```{r bw-cor}
bw_vars <- c("age", "lwt", "race", "smoke", "ptl", "ht", "ui")
dv_bw_cor <- ds.vertCor("DA", bw_vars, verbose = FALSE, datasources = bw_conns)
gt_bw_cor <- cor(bw_merged[, bw_vars])

round(dv_bw_cor$correlation, 3)
```

```{r bw-cor-error}
max(abs(dv_bw_cor$correlation - gt_bw_cor))
```

```{r bw-pca, fig.width=6, fig.height=4}
dv_bw_pca <- ds.vertPCA(cor_result = dv_bw_cor)
gt_bw_pca <- eigen(gt_bw_cor)

barplot(rbind(gt_bw_pca$values, dv_bw_pca$eigenvalues), beside = TRUE,
        names.arg = paste0("PC", seq_along(gt_bw_pca$values)),
        col = c("black", "red"), main = "PCA eigenvalues: centralised vs dsVert",
        ylab = "Eigenvalue")
legend("topright", legend = c("Centralised", "dsVert"), fill = c("black", "red"))
```

```{r bw-pca-table}
data.frame(
  component   = paste0("PC", seq_along(gt_bw_pca$values)),
  centralised = round(gt_bw_pca$values, 4),
  dsVert      = round(dv_bw_pca$eigenvalues, 4),
  difference  = round(abs(gt_bw_pca$values - dv_bw_pca$eigenvalues), 6)
)
```

```{r bw-disconnect}
DSI::datashield.logout(bw_conns)
```

## Dataset 2: CGD (K=3 hospitals)

The CGD dataset from the survival package contains 128 patients from a clinical
trial of interferon gamma for chronic granulomatous disease. It is vertically
partitioned across three institutions:

- **Demographics** holds age, sex, height, weight
- **Treatment centre** holds treatment arm and steroid use
- **Outcomes registry** holds infection status and total infection count

### Connect and explore

```{r cgd-connect, message=FALSE}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "demographics", url = opal_url,
               user = "administrator", password = "password",
               table = "vert_demo.cgd_demographics")
builder$append(server = "treatment", url = opal_url,
               user = "administrator", password = "password",
               table = "vert_demo.cgd_treatment")
builder$append(server = "outcomes", url = opal_url,
               user = "administrator", password = "password",
               table = "vert_demo.cgd_outcomes")
cgd_conns <- DSI::datashield.login(logins = builder$build())
DSI::datashield.assign.table(cgd_conns, "D",
  list(demographics = "vert_demo.cgd_demographics",
       treatment    = "vert_demo.cgd_treatment",
       outcomes     = "vert_demo.cgd_outcomes"))
```

```{r cgd-columns}
DSI::datashield.aggregate(cgd_conns, call("dsvertColNamesDS", data_name = "D"))
```

### Align patients

```{r cgd-psi}
cgd_psi <- ds.psiAlign("D", "patient_id", "DA", verbose = FALSE, datasources = cgd_conns)
cgd_psi$n_common
```

### Building the centralised equivalent

The CGD dataset is in the survival package. We prepare it identically to what
was uploaded to the servers:

```{r cgd-centralised}
data(cgd)
cgd_one <- cgd[cgd$enum == 1, ]  # one row per patient
cgd_one$n_infections <- as.integer(tapply(cgd$status, cgd$id, sum)[as.character(cgd_one$id)])
cgd_one$patient_id <- paste0("CGD", formatC(seq_len(nrow(cgd_one)), width = 3, flag = "0"))
cgd_one$treat_num <- as.integer(cgd_one$treat == "rIFN-g")
cgd_one$sex_num <- as.integer(cgd_one$sex == "male")
```

What each institution holds:

```{r cgd-fragments}
# Demographics (server 1)
head(cgd_one[, c("patient_id", "age", "sex_num", "height", "weight")])

# Treatment assignment (server 2)
head(cgd_one[, c("patient_id", "treat_num", "steroids")])

# Outcomes (server 3)
head(cgd_one[, c("patient_id", "status", "n_infections")])
```

In the centralised case, we have all columns together (`r nrow(cgd_one)` patients
= same as the PSI intersection of `r cgd_psi$n_common`):

```{r cgd-head-merged}
head(cgd_one[, c("patient_id", "age", "sex_num", "height", "weight",
                  "treat_num", "steroids", "status", "n_infections")])
```

### Centralised GLMs

```{r cgd-centralised-glm}
gt_cgd_gauss <- glm(weight ~ age + sex_num + height + treat_num + steroids,
                    data = cgd_one, family = gaussian())
gt_cgd_binom <- glm(status ~ age + sex_num + height + weight + treat_num + steroids,
                    data = cgd_one, family = binomial())
gt_cgd_pois  <- glm(n_infections ~ age + sex_num + height + weight + treat_num + steroids,
                    data = cgd_one, family = poisson())
```

### Federated vertical GLMs (dsVert)

```{r cgd-glm}
dv_cgd_gauss <- ds.vertGLM(weight ~ age + sex_num + height + treat_num + steroids,
                            data = "DA", family = "gaussian",
                            verbose = FALSE, datasources = cgd_conns)

dv_cgd_binom <- ds.vertGLM(status ~ age + sex_num + height + weight + treat_num + steroids,
                            data = "DA", family = "binomial",
                            verbose = FALSE, datasources = cgd_conns)

dv_cgd_pois  <- ds.vertGLM(n_infections ~ age + sex_num + height + weight + treat_num + steroids,
                            data = "DA", family = "poisson",
                            verbose = FALSE, datasources = cgd_conns)
```

### Compare: binomial (predicting serious infection)

```{r cgd-binom-compare, fig.width=7, fig.height=5}
nms <- names(dv_cgd_binom$coefficients)
comp <- data.frame(
  variable    = nms,
  centralised = coef(gt_cgd_binom)[nms],
  dsVert      = dv_cgd_binom$coefficients[nms],
  row.names   = NULL
)
comp$difference <- comp$dsVert - comp$centralised

par(mar = c(4, 10, 2, 2))
idx <- seq_len(nrow(comp))
plot(comp$centralised, idx, pch = 16, col = "black", xlim = range(c(comp$centralised, comp$dsVert)),
     yaxt = "n", ylab = "", xlab = "Coefficient value",
     main = "Binomial GLM: centralised vs dsVert (CGD, K=3)")
points(comp$dsVert, idx, pch = 17, col = "red")
axis(2, at = idx, labels = comp$variable, las = 1)
segments(comp$centralised, idx, comp$dsVert, idx, col = "grey70", lty = 2)
legend("topright", legend = c("Centralised", "dsVert"), pch = c(16, 17), col = c("black", "red"))
```

```{r cgd-binom-table}
comp
```

### Deviance: all families

```{r cgd-deviance}
cgd_dev <- data.frame(
  family      = c("Gaussian", "Binomial", "Poisson"),
  centralised = round(c(deviance(gt_cgd_gauss), deviance(gt_cgd_binom), deviance(gt_cgd_pois)), 2),
  dsVert      = round(c(dv_cgd_gauss$deviance, dv_cgd_binom$deviance, dv_cgd_pois$deviance), 2)
)
cgd_dev$error_pct <- round(100 * abs(cgd_dev$dsVert - cgd_dev$centralised) / cgd_dev$centralised, 2)
cgd_dev
```

### Correlation and PCA

```{r cgd-cor}
cgd_vars <- c("age", "sex_num", "height", "weight", "treat_num", "steroids")
dv_cgd_cor <- ds.vertCor("DA", cgd_vars, verbose = FALSE, datasources = cgd_conns)
gt_cgd_cor <- cor(cgd_one[, cgd_vars])

max(abs(dv_cgd_cor$correlation - gt_cgd_cor))
```

```{r cgd-pca, fig.width=6, fig.height=4}
dv_cgd_pca <- ds.vertPCA(cor_result = dv_cgd_cor)
gt_cgd_pca <- eigen(gt_cgd_cor)

barplot(rbind(gt_cgd_pca$values, dv_cgd_pca$eigenvalues), beside = TRUE,
        names.arg = paste0("PC", seq_along(gt_cgd_pca$values)),
        col = c("black", "red"), main = "PCA eigenvalues: centralised vs dsVert (K=3)",
        ylab = "Eigenvalue")
legend("topright", legend = c("Centralised", "dsVert"), fill = c("black", "red"))
```

```{r cgd-disconnect}
DSI::datashield.logout(cgd_conns)
```

## Overall summary

```{r summary, fig.width=7, fig.height=4}
all_dev <- rbind(
  transform(bw_dev, dataset = "birthwt", K = 2),
  transform(cgd_dev, dataset = "CGD", K = 3)
)
all_dev
```

```{r summary-plot, fig.width=7, fig.height=5}
par(mar = c(5, 4, 3, 2))
labels <- paste0(all_dev$dataset, "\n", all_dev$family)
barplot(rbind(all_dev$centralised, all_dev$dsVert), beside = TRUE,
        names.arg = labels, col = c("grey30", "firebrick"),
        main = "Canonical deviance: centralised vs dsVert",
        ylab = "Deviance", las = 1, cex.names = 0.8)
legend("topright", legend = c("Centralised glm()", "dsVert (federated)"),
       fill = c("grey30", "firebrick"))
```

All six GLM configurations match centralised `glm()` within 0.2% deviance error.
Correlation matrices match within 6 x 10^-6 element-wise, and PCA eigenvalues
within 5 x 10^-6.

**Privacy**: throughout this entire analysis, the analyst received only
p-dimensional aggregate sums. No individual patient data was disclosed.
