---
title: "Linear mixed model validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Linear mixed model validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertLMM()`, `ds.vertLMM.k3()`


The random-intercept LMM is \(y_{ij}=x_{ij}^T\beta+b_i+\epsilon_{ij}\),
with \(b_i\sim N(0,\sigma_b^2)\) and
\(\epsilon_{ij}\sim N(0,\sigma^2)\). The estimator profiles variance
components and solves GLS normal equations. K=2 uses the closed-form route;
K>=3 uses share-domain residual REML and exact cluster-mean GLS transforms.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- nlme::lme(y ~ x1 + x2 + x3,
                     random = ~ 1 | cluster_id,
                     data = pooled, method = "REML")

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "cluster_id", "y")]
)

```

```{r dslite-execution, eval=FALSE}

fit_lmm_k2 <- dsVertClient::ds.vertLMM(
  y ~ x1 + x2 + x3, data = "DA", cluster_col = "cluster_id",
  datasources = conns, verbose = FALSE)
fit_lmm_k3 <- dsVertClient::ds.vertLMM.k3(
  y ~ x1 + x2 + x3, data = "DA", cluster_col = "cluster_id",
  datasources = conns, verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_lmm.R
```

## Disclosure review


Cluster membership is method-level metadata and is sent only through encrypted
server-to-server channels. Original cluster labels are not returned, small
clusters fail closed, and patient-level residuals or BLUPs are not returned.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("lmm")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

