---
title: "Ordinal validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertOrdinal()`, `ds.vertOrdinalJointNewton()`


The proportional-odds model is
\(P(Y\le k|x)=F(\theta_k-x^T\beta)\), with ordered thresholds
\(\theta_1<\cdots<\theta_{K-1}\). The product route evaluates the joint
score for beta and thresholds with class masks, cumulative probabilities,
reciprocals and finite-difference Hessian probes held in Ring127 shares.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- MASS::polr(y_ord ~ x1 + x2 + x3,
                      data = pooled, method = "logistic", Hess = TRUE)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "low_leq", "mid_leq", "y_ord")]
)

```

```{r dslite-execution, eval=FALSE}

fit_ord <- dsVertClient::ds.vertOrdinal(
  y_ord ~ x1 + x2 + x3,
  data = "DA",
  levels_ordered = c("low", "mid", "high"),
  cumulative_template = "%s_leq",
  max_outer = 2L,
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_ordinal_joint.R --K 2 && Rscript scripts/validate_method_ordinal_joint.R --K 3
```

## Disclosure review


Class masks, probabilities, weights and residual-like terms remain encrypted
shares. The client receives guarded class counts and aggregate
gradient/parameter summaries only. The warm cumulative-binomial route is not a
user-facing final estimator.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("ordinal")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

