---
title: "Cox PH validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cox PH validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertCox()`, `ds.vertCoxProfileNonDisclosive()`, `ds.vertCoxDiscreteNonDisclosive()`


The Cox PH model has hazard \(h(t|x)=h_0(t)\exp(x^T\beta)\). With Breslow
ties, the profile partial-likelihood score is built from event-time risk-set
sums \(\sum_{i:t_i \ge t_j}\exp(x_i^T\beta)x_i\). The product route hides
risk-set and event masks as Ring127 shares and opens only slope
score/Hessian aggregates.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- survival::coxph(
  survival::Surv(time, event) ~ x1 + x2 + x3,
  data = pooled, ties = "breslow")

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "time", "event")]
)

```

```{r dslite-execution, eval=FALSE}

fit_cox <- dsVertClient::ds.vertCox(
  survival::Surv(time, event) ~ x1 + x2 + x3,
  data = "DA",
  max_iter = 5L,
  max_event_times = 50L,
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_cox.R --target cox_profile
```

## Disclosure review


Legacy rank/permutation and person-time Cox routes are not offered. Event
times, event indicators, event ranks, risk-set membership, baseline dummies,
and per-person period rows are not returned. Debug traces and bin summaries
are gated diagnostics only.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("cox")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

