---
title: "Ordinal validation (K>=3)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal validation (K>=3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_ordinal), collapse = "\n"))
```

## Scope

Functions: `ds.vert.ordinal()`.

This vignette validates the `K>=3` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

The supported route is joint proportional-odds Newton. Historical cumulative-binomial approximations are only warm starts.

## Mathematical target

P(Y<=k|X)=sigmoid(theta_k-X beta). Validation compares cumulative probabilities with MASS::polr().

## Disclosure review

Cumulative probabilities, residuals, and row scores stay in shares; the client receives thresholds, slopes, and scalar optimizer diagnostics.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 3L

validation_require("MASS")
old_fd <- getOption("dsvert.ord_strict_fd_max_dim", NULL)
options(dsvert.ord_strict_fd_max_dim = 0L)

pooled <- build_ordinal()
tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "x1", "x2"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x3", "y_ord",
                       "low_leq", "med_leq", "high_leq"), drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2"), drop = FALSE],
       s3 = pooled[, c("patient_id", "x3", "y_ord",
                       "low_leq", "med_leq", "high_leq"), drop = FALSE])
}

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref <- MASS::polr(y_ord ~ x1 + x2 + x3, data = pooled, Hess = TRUE)
ref_cum <- sapply(ref$zeta, function(th) {
  stats::plogis(th - as.numeric(as.matrix(pooled[, names(coef(ref))]) %*%
                                  coef(ref)))
})

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    dsVertClient::ds.vert.ordinal(
      y_ord ~ x1 + x2 + x3, data = "DA",
      levels_ordered = c("low", "med", "high"),
      cumulative_template = "%s_leq",
      verbose = validation_demo_verbose(), datasources = conns)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

fit <- result$value
ds_cum <- ordinal_cumprob(fit, pooled)
colnames(ds_cum) <- colnames(ref_cum)
observed <- max(abs(ds_cum - ref_cum))

rows <- row_result(
  "ordinal", "Ordinal", K, "ds.vert.ordinal",
  "balanced synthetic 3-level ordinal fixture",
  "MASS::polr cumulative probabilities",
  "cumulative_probability_max_abs_delta", observed, 0.001,
  "strict-practical",
  "Class probabilities/residuals stay Ring127 shares; no row probabilities are returned.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
options(dsvert.ord_strict_fd_max_dim = old_fd)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
