---
title: "Correlation validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Correlation validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertCor()`


The Pearson correlation between variables \(x_j\) and \(x_k\) is
\(\mathrm{cov}(x_j,x_k)/(s_j s_k)\). Local moments handle within-server
pairs. Cross-server pairs use Beaver cross-products to obtain the aggregate
\(\sum_i x_{ij}x_{ik}\) without revealing either column to the other site.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

X <- MASS::Pima.tr[seq_len(120), c("age", "bmi", "ped", "glu", "bp")]
stats::cor(X)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = data.frame(patient_id = sprintf("P%03d", seq_len(nrow(X))),
                  X[c("age", "bmi", "ped")]),
  s2 = data.frame(patient_id = sprintf("P%03d", seq_len(nrow(X))),
                  X[c("glu", "bp")])
)

```

```{r dslite-execution, eval=FALSE}

cor_fit <- dsVertClient::ds.vertCor(
  "DA",
  variables = list(s1 = c("age", "bmi", "ped"),
                   s2 = c("glu", "bp")),
  datasources = conns)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_descriptive.R
```

## Disclosure review


The full correlation matrix is a second-order aggregate and can become
reconstructive when \(p\) is large relative to \(n\). The product route enforces
minimum-n and p/n guards and blocks high-dimensional diagnostic fixtures by
default.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("correlation")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

