---
title: "PSI alignment validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PSI alignment validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.psiAlign()`, `ds.isPsiAligned()`, `ds.getIdentityPks()`


ECDH-PSI uses the commutativity of elliptic-curve scalar multiplication. For
an identifier hash point \(H(id)\), the reference server computes
\(\alpha H(id)\), the target server computes \(\beta H(id)\), and both sides
can compare \(\alpha\beta H(id)\) without revealing \(id\) or either scalar.
The multi-server intersection keeps only rows present at every site.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

ids <- list(
  s1 = sprintf("P%03d", 1:80),
  s2 = sprintf("P%03d", 11:90),
  s3 = sprintf("P%03d", c(11:80, 101:105))
)
Reduce(intersect, ids)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = data.frame(patient_id = ids$s1, x1 = rnorm(length(ids$s1))),
  s2 = data.frame(patient_id = ids$s2, x2 = rnorm(length(ids$s2))),
  s3 = data.frame(patient_id = ids$s3, y = rnorm(length(ids$s3)))
)

```

```{r dslite-execution, eval=FALSE}

server <- DSLite::newDSLiteServer(
  tables = tables,
  config = DSLite::defaultDSConfiguration(include = c("dsBase", "dsVert")))
builder <- DSI::newDSLoginBuilder()
for (nm in names(tables)) {
  builder$append(server = nm, url = "server", table = nm,
                 driver = "DSLiteDriver")
}
conns <- DSI::datashield.login(builder$build(), assign = TRUE,
                               symbol = "D", opts = list(server = server))
psi <- dsVertClient::ds.psiAlign("D", "patient_id", "DA",
                                 datasources = conns, verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_psi.R
```

## Disclosure review


The client sees aggregate counts and opaque encrypted blobs. Raw identifiers,
matched row-index vectors and per-row alignment positions are not returned.
Pinned peers use signed transport keys so a server can reject untrusted peers
before encrypted PSI payloads are accepted.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("psi")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

