---
title: "Descriptive statistics validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Descriptive statistics validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.desc()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

Each server computes guarded scalar summaries for its own variables. Histogram quantiles are guarded by disclosure thresholds, but this vignette validates the exact mean/sd surface used in the matrix.

## Mathematical target

For each variable x, the checked target is max(|mean_DS - mean_local|, |sd_DS - sd_local|).

## Disclosure review

Only scalar summaries and guarded histogram information are returned; no patient-level values or order statistics are disclosed.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

pooled <- build_pima(60L)
tables <- if (K == 2L) {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
    s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin"),
                drop = FALSE])
} else {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi"), drop = FALSE],
    s2 = pooled[, c("patient_id", "ped", "skin"), drop = FALSE],
    s3 = pooled[, c("patient_id", "npreg", "glu", "bp"), drop = FALSE])
}
vars <- lapply(tables, function(x) setdiff(names(x), "patient_id"))
all_vars <- unlist(vars, use.names = FALSE)

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref_mean <- vapply(pooled[all_vars], mean, numeric(1), na.rm = TRUE)
ref_sd <- vapply(pooled[all_vars], stats::sd, numeric(1), na.rm = TRUE)

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    dsVertClient::ds.vert.desc(
      "DA", variables = vars, n_buckets = 40L,
      verbose = validation_demo_verbose(), datasources = conns)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

desc <- result$value
ds_mean <- stats::setNames(desc$mean, desc$variable)
ds_sd <- stats::setNames(desc$sd, desc$variable)
observed <- max(max_named_delta(ds_mean, ref_mean),
                max_named_delta(ds_sd, ref_sd))

rows <- row_result(
  "descriptive", "Descriptive statistics", K, "ds.vert.desc",
  "MASS::Pima.tr fixture", "central mean/sd",
  "max(mean_sd_abs_delta)", observed, 1e-8, "strict-precise",
  "Returns guarded scalar summaries and histogram-based quantiles, not rows.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
