---
title: "GEE validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GEE validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertGEE()`


GEE solves \(\sum_i D_i^T V_i^{-1}(y_i-\mu_i)=0\) and reports the sandwich
variance \(B^{-1}MB^{-1}\). `ds.vertGEE()` supports independence,
exchangeable and guarded AR1 working correlations for Gaussian, binomial and
Poisson outcomes. Cluster and adjacent-pair scores are computed as aggregate
share-domain sufficient statistics.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- geepack::geeglm(
  y ~ x1 + x2 + x3, id = cluster_id, waves = visit,
  data = pooled, family = binomial(), corstr = "ar1")

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "cluster_id", "visit", "y")]
)

```

```{r dslite-execution, eval=FALSE}

fit_gee <- dsVertClient::ds.vertGEE(
  y ~ x1 + x2 + x3,
  data = "DA",
  family = "binomial",
  id_col = "cluster_id",
  order_col = "visit",
  corstr = "ar1",
  binomial_sigmoid_intervals = 150L,
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_gee.R
```

## Disclosure review


Residuals, fitted means, probabilities, Pearson residuals, row scores, visit
labels and adjacent-pair vectors are not returned. AR1 order metadata is
encrypted server-to-server and only guarded low-dimensional sufficient
statistics reach the client.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("gee")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

