---
title: "GLMM validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GLMM validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_balanced_glmm), collapse = "\n"))
```

## Scope

Functions: `ds.vert.glmm()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

This vignette shows both available product routes: PQL as the lower-cost approximation and Laplace as the more accurate route against lme4::glmer.

## Mathematical target

PQL alternates binomial working responses with aggregate mixed-model equations. Laplace approximates the random-intercept marginal likelihood and checks both fixed effects and scalar random-effect variance.

## Disclosure review

Both routes return fixed effects and scalar variance/quality diagnostics only. Per-patient probabilities, row scores, BLUPs, cluster labels, and cluster vectors are not returned.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

validation_require(c("MASS", "nlme", "lme4"))
pooled <- build_balanced_glmm(seed = 1L, ncl = 10L, m = 5L, b_sd = 0.6)
tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2", "cluster", "y"),
                   drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2"), drop = FALSE],
       s3 = pooled[, c("patient_id", "cluster", "y"), drop = FALSE])
}

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

pooled_ref <- pooled
pooled_ref$cluster <- factor(pooled_ref$cluster)
ref_pql <- suppressWarnings(MASS::glmmPQL(
  y ~ x1 + x2, random = ~1 | cluster, family = binomial(),
  data = pooled_ref, verbose = validation_demo_verbose()))
ref_pql_beta <- nlme::fixef(ref_pql)
ref_laplace <- suppressWarnings(lme4::glmer(
  y ~ x1 + x2 + (1 | cluster), data = pooled_ref, family = binomial(),
  control = lme4::glmerControl(
    optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    run_pql <- elapsed(fit_pql <- dsVertClient::ds.vert.glmm(
      y ~ x1 + x2, data = "DA", cluster_col = "cluster",
      method = "pql", compute_se = FALSE,
      verbose = validation_demo_verbose(), datasources = conns))
    run_laplace <- elapsed(fit_laplace <- dsVertClient::ds.vert.glmm(
      y ~ x1 + x2, data = "DA", cluster_col = "cluster",
      method = "laplace", max_outer = 1L, mode_max_iter = 1L,
      prime_iter = 5L, compute_se = FALSE,
      verbose = validation_demo_verbose(), datasources = conns))
    list(fit_pql = fit_pql, fit_laplace = fit_laplace,
         runtime_pql = run_pql$runtime_s,
         runtime_laplace = run_laplace$runtime_s)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

fit_pql <- result$value$fit_pql
fit_laplace <- result$value$fit_laplace
fixed_delta <- max_named_delta(fit_pql$coefficients, ref_pql_beta)
pql_ran <- isTRUE(fit_pql$iterations >= 1L) &&
  is.data.frame(fit_pql$trace) && nrow(fit_pql$trace) >= 1L &&
  is.finite(fit_pql$sigma_b2) &&
  identical(fit_pql$quality$status, "ok")
observed_pql <- max(fixed_delta, if (pql_ran) 0 else Inf)

fixed_laplace <- max_named_delta(fit_laplace$coefficients,
                                 lme4::fixef(ref_laplace))
sigma_laplace <- abs(fit_laplace$sigma_b2 -
                       as.numeric(lme4::VarCorr(ref_laplace)$cluster)[1L])
safe_return <- identical(fit_laplace$quality$status, "ok") &&
  isFALSE(fit_laplace$disclosure$patient_level_returned) &&
  isFALSE(fit_laplace$disclosure$random_effects_returned) &&
  isFALSE(fit_laplace$disclosure$cluster_vectors_returned)
observed_laplace <- max(fixed_laplace, sigma_laplace,
                        if (safe_return) 0 else Inf)

rows <- rbind(
  row_result(
    "glmm", "GLMM PQL", K, "ds.vert.glmm(method='pql')",
    "synthetic mixed binomial random-intercept fixture",
    "MASS::glmmPQL",
    "fixed_effect_max_abs_delta_and_pql_quality", observed_pql, 0.005,
    "strict-pql",
    "PQL route returns fixed effects and scalar variance diagnostics only.",
    result$value$runtime_pql),
  row_result(
    "glmm", "GLMM Laplace", K, "ds.vert.glmm(method='laplace')",
    "synthetic mixed binomial random-intercept fixture",
    "lme4::glmer",
    "max_fixed_effect_or_sigma_b2_abs_delta", observed_laplace, 0.06,
    "laplace-practical",
    paste("Returns fixed effects, scalar variance, and scalar quality only;",
          "no BLUPs, cluster labels, cluster vectors, or row scores."),
    result$value$runtime_laplace)
)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
