---
title: "Generalization validation"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Generalization validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
force_run <- TRUE
```

## What is validated

This vignette checks whether the public `ds.vert.*` frontdoor functions
generalize beyond the single method-specific fixture. It executes several
synthetic data-generating scenarios, splits each dataset vertically into
K=2 and K>=3 DSLite servers, runs the public default dsVert route, and
compares the result with a centralized R reference.

The default grid covers `glm`, `negative_binomial`, `cox`, `lmm`, `gee`,
`ipw`, and `mi` over two seeds. The heavier joint routes
`multinomial`, `ordinal`, and `glmm` can be included by setting
`DSVERT_GENERALIZATION_INCLUDE_HEAVY=true`. Specific methods and seeds
can be selected with comma-separated `DSVERT_GENERALIZATION_METHODS` and
`DSVERT_GENERALIZATION_SEEDS`.

## Mathematical target

Each row uses the same target as the method-specific validation:
regression coefficients are compared by max absolute coefficient delta,
survival models are compared to `coxph(..., ties = "breslow")`, mixed
models to their centralized random-intercept reference, probability
models by predicted probability delta, and MI by Rubin-pooled
coefficients against a central mean-imputation reference for the compact
check.

## Fixture and reference

The executable chunk below calls `run_generalization_validation()` from
`vignettes/validation_helpers.R` with `force = TRUE`. The helper creates
each fixture in R, partitions it vertically, opens an in-memory DSLite
server, aligns records with PSI, runs the public `ds.vert.*` function,
computes the centralized reference, and compares both outputs. No RDS or
prebuilt table is required or read during execution; cache files under
`vignettes/validation-cache/` are written by this execution path only.

## Disclosure review

The grid exercises only the accepted public paths. It does not call
legacy routes that reveal observation-level or patient-level aggregates.
Returned objects are model-level summaries, guarded low-dimensional
aggregate summaries, or scalar diagnostics. For MI scenarios with missing
values, PSI uses `na.action = "none"` so missing cells remain server-side
for imputation rather than being silently dropped before analysis.

The fixture keeps `datashield.privacyLevel = 5` and is sized so standard
disclosure guards remain active. Only `dsvert.require_trusted_peers` is
disabled for DSLite because there is no real Opal/Rock deployment in
this local validation context.

## Executed evidence

```{r evidence}
rows <- run_generalization_validation(force = force_run, assert = TRUE,
  trace = TRUE)
display_generalization(rows)
```

```{r assertion, include=FALSE}
assert_generalization(rows)
```

## Verdict

The vignette fails during rendering if any public default route leaves
the accepted numerical envelope or is marked with an unacceptable quality
or disclosure status.
