---
title: "Multiple imputation validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multiple imputation validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertMI()`


For \(m=1,\ldots,M\), missing values are imputed server-side, the analysis GLM
returns \(\hat\beta_m\) and \(\hat\Sigma_m\), and Rubin's rules compute
\(\bar\beta=M^{-1}\sum_m\hat\beta_m\) and
\(T=W+(1+M^{-1})B\), where \(W\) is the mean within-imputation covariance and
\(B\) is the between-imputation covariance.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fits <- lapply(seq_len(3), function(m) {
  imputed <- same_server_imputer(pooled, seed = 100 + m)
  glm(y ~ x1 + x2 + x3, data = imputed)
})
rubin_pool(fits)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1")],
  s2 = pooled[c("patient_id", "x2", "x3", "y")]
)

```

```{r dslite-execution, eval=FALSE}

fit_mi <- dsVertClient::ds.vertMI(
  y ~ x1 + x2 + x3,
  data = "DA",
  impute_columns = "x1",
  m = 3L,
  family = "gaussian",
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_mi.R
```

## Disclosure review


Imputed values never leave the server that owns the column. The client receives
only per-imputation model-scale coefficient/covariance summaries and Rubin
pooled summaries. Exact extrema in imputation diagnostics are suppressed.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("mi")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

