---
title: "LASSO validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{LASSO validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.lasso_proximal()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

The proximal LASSO path consumes the distributed GLM fit. The validation uses lambda = 0, where the penalized solution must equal the prime GLM estimate.

## Mathematical target

At lambda = 0, the proximal quadratic argmin is beta_hat, so the checked target is max |beta_lasso0 - beta_glm|.

## Disclosure review

The proximal step uses only model-level beta/Hessian information and returns coefficients; it does not request row-level data.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

pooled <- build_pima(60L)
tables <- if (K == 2L) {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
    s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin",
                    "diabetes"), drop = FALSE])
} else {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi"), drop = FALSE],
    s2 = pooled[, c("patient_id", "ped", "skin"), drop = FALSE],
    s3 = pooled[, c("patient_id", "npreg", "glu", "bp",
                    "diabetes"), drop = FALSE])
}
fm <- diabetes ~ age + bmi + ped + glu + bp + skin + npreg

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    fit <- dsVertClient::ds.vert.glm(
      fm, data = "DA", family = "gaussian", max_iter = 30L,
      verbose = validation_demo_verbose(), datasources = conns)
    prox0 <- dsVertClient::ds.vert.lasso_proximal(
      fit, lambda = 0, max_iter = 1000L, tol = 1e-8)
    list(fit = fit, prox0 = prox0)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

observed <- max_named_delta(result$value$prox0$coefficients,
                            result$value$fit$coefficients)

rows <- row_result(
  "lasso", "LASSO", K, "ds.vert.lasso_proximal(lambda=0)",
  "MASS::Pima.tr fixture", "OLS limit of ds.vert.glm",
  "lambda0_coef_abs_delta", observed, 1e-8, "strict-precise",
  "Consumes only model-level GLM aggregates; no extra patient-level values are returned.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
