---
title: "Multiple imputation validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multiple imputation validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_mi), collapse = "\n"))
```

## Scope

Functions: `ds.vert.mi()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

Missing covariates are imputed server-side for each round, then GLM fits are pooled client-side.

## Mathematical target

The compact reference uses central mean imputation, and the distributed route is checked by pooled coefficient delta.

## Disclosure review

Imputed columns stay server-side. The client receives pooled beta/covariance summaries, not imputed patient-level values.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

pooled <- build_mi()
tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2", "x3", "y"), drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2"), drop = FALSE],
       s3 = pooled[, c("patient_id", "x3", "y"), drop = FALSE])
}

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref_data <- pooled
ref_data$x2[is.na(ref_data$x2)] <- mean(ref_data$x2, na.rm = TRUE)
ref <- coef(stats::lm(y ~ x1 + x2 + x3, data = ref_data))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns, na.action = "none")
    dsVertClient::ds.vert.mi(
      y ~ x1 + x2 + x3, data = "DA", impute_columns = "x2",
      family = "gaussian",
      verbose = validation_demo_verbose(), datasources = conns, seed = 12L)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

fit <- result$value
observed <- max_named_delta(fit$coefficients, ref)

rows <- row_result(
  "mi", "Multiple imputation", K, "ds.vert.mi",
  "synthetic missing-covariate fixture",
  "central mean-imputation reference",
  "pooled_coef_abs_delta", observed, 0.02, "strict-practical",
  "Imputed columns stay server-side; client pools beta/covariance only.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
