---
title: "Negative binomial validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Negative binomial validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_nb), collapse = "\n"))
```

## Scope

Functions: `ds.vert.nb()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

This vignette validates the accurate non-disclosive full-regression route against MASS::glm.nb().

## Mathematical target

NB2 uses Var(Y|X)=mu+mu^2/theta and log(mu)=X beta. The validated route evaluates beta/theta score pieces in shares and returns model-level coefficients with a scalar dispersion estimate.

## Disclosure review

The route keeps eta, mu, reciprocal, score, and Fisher pieces in shares and returns beta plus scalar theta.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

validation_require("MASS")
pooled <- build_nb(60L)
tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "x1", "x2"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x3", "y"), drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2"), drop = FALSE],
       s3 = pooled[, c("patient_id", "x3", "y"), drop = FALSE])
}

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref_accurate <- coef(MASS::glm.nb(y ~ x1 + x2 + x3, data = pooled))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    run_accurate <- elapsed(accurate <- dsVertClient::ds.vert.nb(
      y ~ x1 + x2 + x3, data = "DA", method = "accurate",
      max_iter = 25L,
      verbose = validation_demo_verbose(), datasources = conns))
    list(accurate = accurate,
         runtime_accurate = run_accurate$runtime_s)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

accurate_delta <- max_named_delta(result$value$accurate$coefficients,
                                  ref_accurate)

rows <- row_result(
  "negative_binomial", "Negative binomial accurate", K,
  "ds.vert.nb(method='accurate')",
  "synthetic NB fixture", "MASS::glm.nb",
  "coef_max_abs_delta", accurate_delta, 0.001,
  "strict-practical",
  "Full-regression beta/theta score components remain aggregate.",
  result$value$runtime_accurate)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
