---
title: "Cox PH validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cox PH validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_cox), collapse = "\n"))
```

## Scope

Functions: `ds.vert.cox()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

The product Cox route is the non-disclosive Breslow profile route. The discrete-time route is a different estimand, not the fast approximation to this profile target, so this vignette validates the route used in the product matrix.

## Mathematical target

The Cox PH score is sum_i delta_i [x_i - E_beta{x | t_i in risk set}]. Validation compares beta to survival::coxph(ties='breslow').

## Disclosure review

Risk-set/event-time score terms remain shared. The client receives coefficients and scalar convergence diagnostics, not event-time vectors or patient risk-set membership.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

validation_require("survival")
pooled <- build_cox(60L)
tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "x1", "x2"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x3", "time", "event"), drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "x1"), drop = FALSE],
       s2 = pooled[, c("patient_id", "x2"), drop = FALSE],
       s3 = pooled[, c("patient_id", "x3", "time", "event"), drop = FALSE])
}

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref <- coef(survival::coxph(
  survival::Surv(time, event) ~ x1 + x2 + x3,
  data = pooled, ties = "breslow"))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    dsVertClient::ds.vert.cox(
      survival::Surv(time, event) ~ x1 + x2 + x3, data = "DA",
      method = "profile",
      verbose = validation_demo_verbose(), datasources = conns)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

fit <- result$value
observed <- max_named_delta(fit$coefficients, ref)

rows <- row_result(
  "cox", "Cox PH", K, "ds.vert.cox(method='profile')",
  "synthetic discretised survival fixture",
  "survival::coxph(ties='breslow')",
  "coef_max_abs_delta", observed, 1e-3, "strict-practical",
  "Event-time score terms remain shared; client receives beta and scalar diagnostics.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
