---
title: "GLM inference validation (K>=3)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GLM inference validation (K>=3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.confint(), ds.vert.wald(), ds.vert.contrast(), ds.vert.lr()`.

This vignette validates the `K>=3` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

Inference helpers post-process the model-level beta/covariance/deviance already returned by ds.vert.glm().

## Mathematical target

Wald and contrast tests use K beta and K V K'. LR tests use the deviance difference with a chi-square reference.

## Disclosure review

No extra server aggregate is opened beyond the fitted GLM objects.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 3L

pooled <- build_pima(60L)
tables <- if (K == 2L) {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
    s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin",
                    "diabetes"), drop = FALSE])
} else {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi"), drop = FALSE],
    s2 = pooled[, c("patient_id", "ped", "skin"), drop = FALSE],
    s3 = pooled[, c("patient_id", "npreg", "glu", "bp",
                    "diabetes"), drop = FALSE])
}
fm <- diabetes ~ age + bmi + ped + glu + bp + skin + npreg
red <- diabetes ~ age + bmi + ped + glu + bp + skin

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    full <- dsVertClient::ds.vert.glm(
      fm, data = "DA", family = "gaussian", max_iter = 30L,
      compute_se = TRUE, compute_deviance = TRUE,
      verbose = validation_demo_verbose(), datasources = conns)
    reduced <- dsVertClient::ds.vert.glm(
      red, data = "DA", family = "gaussian", max_iter = 30L,
      compute_se = TRUE, compute_deviance = TRUE,
      verbose = validation_demo_verbose(), datasources = conns)
    ci <- dsVertClient::ds.vert.confint(full)
    wald <- dsVertClient::ds.vert.wald(full, "age")
    Kmat <- matrix(0, nrow = 1, ncol = length(full$coefficients),
                   dimnames = list("age", names(full$coefficients)))
    Kmat[1, "age"] <- 1
    contrast <- dsVertClient::ds.vert.contrast(full, Kmat)
    lr <- dsVertClient::ds.vert.lr(reduced, full)
    list(full = full, ci = ci, wald = wald, contrast = contrast, lr = lr)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

full <- result$value$full
se <- full$std_errors["age"]
est <- full$coefficients["age"]
z <- est / se
manual_p <- 2 * stats::pnorm(-abs(z))
lr_p_ref <- stats::pchisq(result$value$lr$statistic,
                          df = result$value$lr$df,
                          lower.tail = FALSE)
observed <- max(abs(result$value$wald$p_value - manual_p),
                abs(result$value$contrast$estimate - est),
                abs(result$value$ci["age", "estimate"] - est),
                abs(result$value$lr$p_value - lr_p_ref))

rows <- row_result(
  "inference", "Inference helpers", K,
  "ds.vert.confint / ds.vert.wald / ds.vert.contrast / ds.vert.lr",
  "MASS::Pima.tr fixture", "manual algebra on ds.vert.glm output",
  "algebra_max_abs_delta", observed, 1e-10, "strict-precise",
  "Post-processes released model-level beta/covariance/deviance only.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
