---
title: "LMM validation"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{LMM validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = FALSE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
force_run <- identical(tolower(Sys.getenv("DSVERT_VALIDATION_FORCE", "false")), "true")
```

## What is validated

Functions: `ds.vertLMM(), ds.vertLMM.k3()`.

Random-intercept LMM uses guarded cluster metadata and MPC cross-products to fit fixed effects and variance components.

## Mathematical target

The model is y = X beta + b_cluster + epsilon with b ~ N(0, sigma_b^2). The validation compares fixed effects to lme4::lmer().

## Fixture and reference

Fixture: Synthetic balanced random-intercept fixture, with cluster sizes above privacy thresholds.

Centralized reference: lme4::lmer(... + (1 | cluster)) on the pooled fixture.

The executable chunk below calls `run_validation()` from
`vignettes/validation_helpers.R`. That helper constructs the fixture,
opens a DSLite server, performs PSI alignment, runs the dsVertClient
product route for K=2 and K=3, computes the centralized reference, and
compares both results. No RDS or result table outside this package is
required; if a local `vignettes/validation-cache/` file exists it is a
cache produced by this same execution path.

## Disclosure review

Cluster membership is used internally at the accepted LMM tier. Per-cluster residuals and BLUP vectors are not returned.

The fixture keeps `datashield.privacyLevel = 5` and is sized so the
standard disclosure guards remain active. Only `dsvert.require_trusted_peers`
is disabled for DSLite because there is no real Opal/Rock deployment in
this local validation context.

## Executed evidence

```{r evidence}
rows <- run_validation("lmm", force = force_run)
display_validation(rows)
```

```{r assertion, include=FALSE}
assert_validation(rows)
```

## Verdict

The vignette fails during rendering if either K=2 or K>=3 leaves the
accepted numerical envelope or is marked as disclosive.
