---
title: "Correlation validation"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Correlation validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = FALSE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
force_run <- identical(tolower(Sys.getenv("DSVERT_VALIDATION_FORCE", "false")), "true")
```

## What is validated

Functions: `ds.vertCor()`.

The method releases a low-dimensional correlation matrix from server-local moments and MPC cross-products.

## Mathematical target

For variables x_j and x_l, r_jl = cov(x_j, x_l) / (sd(x_j) sd(x_l)). The validation compares the full matrix to stats::cor().

## Fixture and reference

Fixture: MASS::Pima.tr numeric fixture split vertically.

Centralized reference: stats::cor() on the pooled fixture.

The executable chunk below calls `run_validation()` from
`vignettes/validation_helpers.R`. That helper constructs the fixture,
opens a DSLite server, performs PSI alignment, runs the dsVertClient
product route for K=2 and K=3, computes the centralized reference, and
compares both results. No RDS or result table outside this package is
required; if a local `vignettes/validation-cache/` file exists it is a
cache produced by this same execution path.

## Disclosure review

The disclosure surface is the guarded p by p correlation matrix, the same aggregate tier used by downstream PCA.

The fixture keeps `datashield.privacyLevel = 5` and is sized so the
standard disclosure guards remain active. Only `dsvert.require_trusted_peers`
is disabled for DSLite because there is no real Opal/Rock deployment in
this local validation context.

## Executed evidence

```{r evidence}
rows <- run_validation("correlation", force = force_run)
display_validation(rows)
```

```{r assertion, include=FALSE}
assert_validation(rows)
```

## Verdict

The vignette fails during rendering if either K=2 or K>=3 leaves the
accepted numerical envelope or is marked as disclosive.
