---
title: "IPW validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{IPW validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertIPW()`


IPW fits a propensity model \(e(x)=P(T=1|X=x)\), then an outcome model with
weights \(w_i=T_i/e(x_i)+(1-T_i)/(1-e(x_i))\). The product route validates the
two-stage propensity plus weighted GLM workflow while keeping weights and
sqrt-weights server-side or in additive shares.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

prop_ref <- glm(treat ~ x1 + x2, data = pooled, family = binomial())
pooled$ipw <- with(pooled, ifelse(treat == 1, 1 / fitted(prop_ref),
                                  1 / (1 - fitted(prop_ref))))
out_ref <- glm(y ~ treat + x1 + x2, data = pooled,
               family = binomial(), weights = ipw)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "treat", "ipw", "y")]
)

```

```{r dslite-execution, eval=FALSE}

fit_ipw <- dsVertClient::ds.vertIPW(
  outcome_formula = y ~ treat + x1 + x2,
  propensity_formula = treat ~ x1 + x2,
  data = "DA",
  weights_column = "ipw",
  outcome_family = "binomial",
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_ipw.R
```

## Disclosure review


Patient weights are not transported plaintext to DCF peers. Weighted
gradients/deviance use Beaver products of weight shares and residual shares.
The widened secure sigmoid fixed the previous stress accuracy gap without
changing disclosure.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("ipw")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

