---
title: "GEE validation (K>=3)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GEE validation (K>=3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.gee()`.

This vignette validates the `K>=3` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

The GEE route computes protected estimating-equation aggregates and robust sandwich-level summaries for clustered data.

## Mathematical target

For independence working correlation, sum_i D_i' V_i^{-1}(y_i-mu_i)=0. Validation compares beta to geepack::geeglm().

## Disclosure review

The client receives regression and sandwich-level aggregates, not row scores, fitted values, or residual vectors.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 3L

validation_require("geepack")
pooled <- build_pima(60L)
pooled$cluster <- as.integer((seq_len(nrow(pooled)) - 1L) %/% 10L) + 1L
tables <- if (K == 2L) {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
    s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin",
                    "diabetes", "cluster"), drop = FALSE])
} else {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi"), drop = FALSE],
    s2 = pooled[, c("patient_id", "ped", "skin"), drop = FALSE],
    s3 = pooled[, c("patient_id", "npreg", "glu", "bp",
                    "diabetes", "cluster"), drop = FALSE])
}
fm <- diabetes ~ age + bmi + ped + glu + bp + skin + npreg

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

ref <- coef(geepack::geeglm(
  fm, data = pooled, id = cluster, family = gaussian(),
  corstr = "independence"))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    dsVertClient::ds.vert.gee(
      fm, data = "DA", family = "gaussian", id_col = "cluster",
      corstr = "independence", max_iter = 30L,
      verbose = validation_demo_verbose(), datasources = conns)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

fit <- result$value
observed <- max_named_delta(fit$coefficients, ref)

rows <- row_result(
  "gee", "GEE", K, "ds.vert.gee(corstr='independence')",
  "MASS::Pima.tr clustered fixture", "geepack::geeglm",
  "coef_max_abs_delta", observed, 0.01, "strict-practical",
  "Returns regression and sandwich-level aggregates, not row scores.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
