---
title: "Negative binomial validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Negative binomial validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertNB()`, `ds.vertNBMoMTheta()`, `ds.vertNBFullRegTheta()`


For NB2 log-link regression,
\(\mathrm{Var}(Y_i|X_i)=\mu_i+\mu_i^2/\theta\). The full-regression theta
score contains terms such as \(\sum_i \log(\mu_i+\theta)\) and
\(\sum_i (y_i+\theta)/(\mu_i+\theta)\). The product route evaluates those
terms in Ring127 shares and alternates theta Newton updates with aggregate
Fisher beta updates.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- MASS::glm.nb(y ~ x1 + x2 + x3, data = pooled)
coef(fit_ref)
fit_ref$theta

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "y")]
)

```

```{r dslite-execution, eval=FALSE}

fit_nb <- dsVertClient::ds.vertNBFullRegTheta(
  y ~ x1 + x2 + x3,
  data = "DA",
  variant = "full_reg_nd",
  theta_max_iter = 8L,
  beta_max_iter = 2L,
  compute_covariance = TRUE,
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_nb.R --variant full_reg_nd
```

## Disclosure review


The removed `variant='full_reg'` route disclosed non-label eta in plaintext.
The `full_reg_nd` route keeps eta, mu, logarithms, reciprocals, score
residuals, Fisher weights, and weighted covariates in Ring127 additive shares.
Only scalar theta sums, aggregate beta scores, and aggregate Fisher matrices
are reconstructed.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("negative_binomial")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

