---
title: "LASSO validation"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{LASSO validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
force_run <- TRUE
```

## What is validated

Functions: `ds.vert.lasso_proximal()`.

The proximal LASSO solver consumes the distributed GLM fit and optimizes the penalized quadratic objective client-side.

## Mathematical target

The checked identity is the lambda = 0 limit: argmin 1/2 (beta - beta_hat)^T H (beta - beta_hat) equals beta_hat.

## Fixture and reference

Fixture: MASS::Pima.tr Gaussian GLM fixture.

Centralized reference: The prime ds.vert.glm coefficients at lambda = 0.

The executable chunk below calls `run_validation()` from
`vignettes/validation_helpers.R` with `force = TRUE`. That helper
constructs the fixture, opens a DSLite server, performs PSI alignment,
runs the dsVertClient product route for K=2 and K=3, computes the
centralized reference, and compares both results. No RDS or result table
outside this package is required or read during vignette execution; any
local `vignettes/validation-cache/` file is overwritten by this same
execution path.

## Disclosure review

The proximal step uses only model-level beta/Hessian information and returns coefficients; it does not request row-level data.

The fixture keeps `datashield.privacyLevel = 5` and is sized so the
standard disclosure guards remain active. Only `dsvert.require_trusted_peers`
is disabled for DSLite because there is no real Opal/Rock deployment in
this local validation context.

## Executed evidence

```{r evidence}
rows <- run_validation("lasso", force = force_run, trace = TRUE)
display_validation(rows)
```

```{r assertion, include=FALSE}
assert_validation(rows)
```

## Verdict

The vignette fails during rendering if either K=2 or K>=3 leaves the
accepted numerical envelope or is marked as disclosive.
