---
title: "PCA validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PCA validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.pca()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

PCA is computed from the federated correlation matrix, not from individual-level scores.

## Mathematical target

The target is R = V Lambda V^T. Validation compares eigenvalues and sign-aligned loadings with eigen(cor(X)).

## Disclosure review

Only eigenvalues and loadings derived from the aggregate correlation matrix are returned; no individual PCA scores are released.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

pooled <- build_pima(60L)
tables <- if (K == 2L) {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi", "ped"), drop = FALSE],
    s2 = pooled[, c("patient_id", "npreg", "glu", "bp", "skin"),
                drop = FALSE])
} else {
  list(
    s1 = pooled[, c("patient_id", "age", "bmi"), drop = FALSE],
    s2 = pooled[, c("patient_id", "ped", "skin"), drop = FALSE],
    s3 = pooled[, c("patient_id", "npreg", "glu", "bp"), drop = FALSE])
}
vars <- lapply(tables, function(x) setdiff(names(x), "patient_id"))

knitr::kable(utils::head(pooled))
knitr::kable(partition_summary(tables))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    cor_ds <- dsVertClient::ds.vert.cor(
      "DA", variables = vars,
      verbose = validation_demo_verbose(), datasources = conns)
    pca_ds <- dsVertClient::ds.vert.pca(
      cor_result = cor_ds,
      verbose = validation_demo_verbose(), datasources = conns)
    list(cor_ds = cor_ds, pca_ds = pca_ds)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

cor_ds <- result$value$cor_ds
pca_ds <- result$value$pca_ds
cor_ref <- stats::cor(pooled[cor_ds$var_names])
eig_ref <- eigen(cor_ref, symmetric = TRUE)
load_ref <- eig_ref$vectors[, seq_len(ncol(pca_ds$loadings)), drop = FALSE]
rownames(load_ref) <- rownames(pca_ds$loadings)
load_ds <- pca_ds$loadings
for (j in seq_len(ncol(load_ds))) {
  if (sum(load_ds[, j] * load_ref[, j]) < 0) load_ds[, j] <- -load_ds[, j]
}
observed <- max(max(abs(pca_ds$eigenvalues - eig_ref$values)),
                max(abs(load_ds - load_ref)))

rows <- row_result(
  "pca", "PCA", K, "ds.vert.pca",
  "MASS::Pima.tr fixture", "eigen(cor(X))",
  "max(eigen_loading_abs_delta)", observed, 1e-4,
  "strict-practical",
  "Reuses the correlation matrix; no score or loading per patient is returned.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
