---
title: "PSI alignment validation (K>=3)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PSI alignment validation (K>=3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Scope

Functions: `ds.vert.align()`.

This vignette validates the `K>=3` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

ECDH-PSI aligns vertically split tables on a shared identifier. Each server masks identifiers, the protocol computes the common set, and the aligned object keeps only matched rows in a common order.

## Mathematical target

The target is the set intersection I = intersection_k I_k. Validation checks matched counts and that deterministic id-derived columns align perfectly after PSI.

## Disclosure review

The analyst receives counts and status fields only. Patient IDs and matched row-index vectors are not returned; the legacy matched-index aggregate is checked as blocked.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 3L

contains_ids <- function(x) {
  if (is.character(x)) return(any(grepl("^P[0-9]{4}$", x)))
  if (is.list(x)) return(any(vapply(x, contains_ids, logical(1))))
  FALSE
}

make_table <- function(ids, server_index) {
  id_num <- as.integer(sub("^P", "", ids))
  ord <- sample(seq_along(ids))
  data.frame(patient_id = ids[ord], id_num = id_num[ord],
             value = id_num[ord] + server_index / 10)
}

set.seed(9100 + K)
id_sets <- if (K == 2L) {
  list(s1 = sprintf("P%04d", 1:80),
       s2 = sprintf("P%04d", 11:90))
} else {
  list(s1 = sprintf("P%04d", 1:85),
       s2 = sprintf("P%04d", 11:95),
       s3 = sprintf("P%04d", 6:80))
}
common_ids <- Reduce(intersect, id_sets)
tables <- Map(make_table, id_sets, seq_along(id_sets))

knitr::kable(partition_summary(tables))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi <- psi_align(conns)
    counts <- DSI::datashield.aggregate(
      conns, call(name = "getObsCountDS", data_name = "DA"))
    n_by_server <- vapply(counts, function(x) as.integer(x$n_obs), integer(1))
    variables <- stats::setNames(lapply(names(conns), function(.x) "id_num"),
                                 names(conns))
    cor_fit <- dsVertClient::ds.vert.cor(
      "DA", variables = variables,
      verbose = validation_demo_verbose(), datasources = conns)
    legacy_blocked <- tryCatch({
      DSI::datashield.aggregate(conns[1],
        call(name = "psiGetMatchedIndicesDS"))
      FALSE
    }, error = function(e) TRUE)
    list(psi = psi, n_by_server = n_by_server, cor_fit = cor_fit,
         legacy_blocked = legacy_blocked)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

offdiag <- result$value$cor_fit$correlation[
  upper.tri(result$value$cor_fit$correlation)]
observed <- max(abs(result$value$n_by_server - length(common_ids)),
                abs((result$value$psi$n_common %||%
                       result$value$psi[[1]]$n_matched) - length(common_ids)),
                max(abs(offdiag - 1)))
audit_ok <- !contains_ids(result$value$psi) &&
  isTRUE(result$value$legacy_blocked)

rows <- row_result(
  "psi", "PSI alignment", K, "ds.vert.align",
  "synthetic ID intersection", "deterministic set intersection",
  "max(count_delta, correlation_delta)", observed, 0,
  "strict-precise",
  "Returns match counts/status only; matched IDs and row indices are not returned.",
  result$runtime_s,
  if (audit_ok) "index reveal blocked" else "audit warning")

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
