---
title: "GLMM validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GLMM validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertGLMM()`


The supported GLMM route is binomial random-intercept PQL. The model is
\(\mathrm{logit}\{P(y_{ij}=1|b_i)\}=x_{ij}^T\beta+b_i\) with
\(b_i\sim N(0,\sigma_b^2)\). PQL iterates working responses and weights, but
the product implementation keeps these quantities in Ring127 shares and
returns only fixed effects, scalar variance components and diagnostics.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

fit_ref <- MASS::glmmPQL(
  y_bin ~ x1 + x2 + x3,
  random = ~ 1 | cluster_id,
  family = binomial(),
  data = pooled,
  verbose = FALSE)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "x1", "x2")],
  s2 = pooled[c("patient_id", "x3", "cluster_id", "y_bin")]
)

```

```{r dslite-execution, eval=FALSE}

fit_glmm <- dsVertClient::ds.vertGLMM(
  y_bin ~ x1 + x2 + x3,
  data = "DA",
  cluster_col = "cluster_id",
  ring = 127L,
  compute_se = FALSE,
  datasources = conns,
  verbose = FALSE)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_glmm.R
```

## Disclosure review


The removed method-selection path is not user-invokable. Working responses,
weights, probabilities, residuals and row scores remain in shares. The public
object is audited for no length-n and no cluster-length vectors.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("glmm")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

