---
title: "Contingency tests validation (K=2)"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Contingency tests validation (K=2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = TRUE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
validation_load_packages()
options(dsvert.beaver_preprocessing = "auto")
partition_summary <- function(tables) {
  data.frame(
    server = names(tables),
    n = vapply(tables, nrow, integer(1)),
    columns = vapply(tables, function(x) paste(names(x), collapse = ", "), character(1)),
    stringsAsFactors = FALSE)
}
```

## Dataset generator

The fixture is generated inside this vignette run. The helper body used
to create the pooled local data is printed here before the analysis.

```{r dataset-generator}
cat(paste(deparse(build_pima), collapse = "\n"))
```

## Scope

Functions: `ds.vert.chisq_cross()`.

This vignette validates the `K=2` modality. It creates the local
pooled fixture, partitions it vertically, starts an in-memory DSLite
server, aligns records with PSI, runs the distributed dsVert route, runs
the centralized R reference on the same pooled fixture, and computes the
numerical delta.

No external RDS, cache, or pre-loaded server table is used. Rendering the
vignette executes the workflow from dataset generation to assertion.

## Method

Cross-server contingency testing builds guarded cell counts across vertically separated categorical variables before applying the usual chi-square/Fisher calculations.

## Mathematical target

The validation compares the released observed table and X^2 statistic with table() and chisq.test() on the pooled fixture.

## Disclosure review

Exact counts are released only after small-cell checks pass. Sparse positive cells fail closed before release.

The DSLite validation uses `datashield.privacyLevel = 5`. Trusted-peer
pinning is disabled only because DSLite is an in-memory test backend, not
an Opal/Rock deployment. That does not weaken the method-level check of
what the product route returns to the analyst.

## Executed validation

```{r execution}
K <- 2L

pooled <- build_pima(80L)
pooled$diabetes <- factor(ifelse(pooled$diabetes == 1, "yes", "no"),
                          levels = c("no", "yes"))
pooled$age_grp <- factor(ifelse(pooled$age >= median(pooled$age),
                                "older", "younger"),
                         levels = c("younger", "older"))
pooled$bp_grp <- factor(ifelse(pooled$bp >= median(pooled$bp),
                               "high_bp", "low_bp"),
                        levels = c("low_bp", "high_bp"))

tables <- if (K == 2L) {
  list(s1 = pooled[, c("patient_id", "age_grp"), drop = FALSE],
       s2 = pooled[, c("patient_id", "bp_grp", "diabetes"), drop = FALSE])
} else {
  list(s1 = pooled[, c("patient_id", "age_grp"), drop = FALSE],
       s2 = pooled[, c("patient_id", "bp_grp"), drop = FALSE],
       s3 = pooled[, c("patient_id", "diabetes"), drop = FALSE])
}

knitr::kable(utils::head(pooled[, c("patient_id", "age_grp", "bp_grp",
                                    "diabetes")]))
knitr::kable(partition_summary(tables))

tab <- table(pooled$age_grp, pooled$diabetes)
chi_ref <- suppressWarnings(stats::chisq.test(tab, correct = FALSE))

result <- elapsed({
  conns <- connect_dslite(tables)
  tryCatch({
    psi_align(conns)
    dsVertClient::ds.vert.chisq_cross(
      "DA", "age_grp", "diabetes", correct = FALSE, fisher = TRUE,
      verbose = validation_demo_verbose(), datasources = conns)
  }, finally = try(DSI::datashield.logout(conns), silent = TRUE))
})

cross <- result$value
ref_mat <- unclass(tab)[rownames(cross$observed),
                        colnames(cross$observed), drop = FALSE]
observed <- max(max(abs(cross$observed - ref_mat)),
                abs(cross$chisq - unname(chi_ref$statistic)))

rows <- row_result(
  "contingency", "Contingency tests", K,
  "ds.vert.chisq / ds.vert.fisher / ds.vert.chisq_cross",
  "MASS::Pima.tr categorical fixture", "chisq.test / fisher.test",
  "max(count_or_chisq_delta)", observed, 0, "strict-precise",
  "Releases guarded table counts and test statistics; small cells fail closed.",
  result$runtime_s)

display_validation(rows)
assert_validation(rows)
```

## Verdict

Rendering fails if the distributed result leaves the accepted numerical
envelope or if the method row is marked disclosive.
