---
title: "Contingency table validation"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Contingency table validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## What is validated

Functions: `ds.vertChisq()`, `ds.vertFisher()`, `ds.vertChisqCross()`


For an observed table \(O_{ab}\), Pearson's statistic is
\(\sum_{ab}(O_{ab}-E_{ab})^2/E_{ab}\), where
\(E_{ab}=O_{a+}O_{+b}/n\). Same-server tables release guarded aggregate
counts. Cross-server tables construct one-hot indicators and compute
\(\sum_i 1\{A_i=a\}1\{B_i=b\}\) with Beaver dot products before the exact
table is opened.


## Centralized reference

The centralized reference is built from the same deterministic rows and
formula as the DSLite run. The long method harness stores the exact seed,
split and reference object in the cache listed below.

```{r central-reference, eval=FALSE}

pooled <- MASS::Pima.tr[seq_len(120), ]
pooled$age_group <- cut(pooled$age, breaks = c(0, 30, Inf))
pooled$glu_group <- cut(pooled$glu, breaks = c(0, 120, Inf))
tab <- table(pooled$age_group, pooled$glu_group)
stats::chisq.test(tab, correct = TRUE)
stats::fisher.test(tab)

```

## Vertical DSLite split

The validation split creates independent server tables with the same
`patient_id` key and then aligns them with `ds.psiAlign()`.

```{r vertical-split, eval=FALSE}

tables <- list(
  s1 = pooled[c("patient_id", "age_group")],
  s2 = pooled[c("patient_id", "glu_group")]
)

```

```{r dslite-execution, eval=FALSE}

same <- dsVertClient::ds.vertChisq(
  "DA", "age_group", "glu_group", server = "s2", datasources = conns)
exact <- dsVertClient::ds.vertFisher(
  "DA", "age_group", "glu_group", server = "s2", datasources = conns)
cross <- dsVertClient::ds.vertChisqCross(
  "DA", "age_group", "glu_group", fisher = TRUE, datasources = conns)

```

To reproduce the cache from the repository root:

```{r reproduce-cache, eval=FALSE}
Rscript scripts/validate_method_tables.R
```

## Disclosure review


Positive cells, row margins, and column margins below the configured privacy
threshold block the same-server release. In the cross-server route, DCF
threshold checks reveal only pass/fail before exact cell counts are released.
No one-hot patient vector is returned to the analyst.


## Executed evidence check

This chunk is evaluated when the vignette renders. It fails if either K mode is
not marked non-disclosive, is not `PASS`, or exceeds its accepted tolerance.

```{r evidence}
rows <- validation_rows("contingency")
assert_validation(rows)
display_validation(rows)
```

## Verdict

Both K=2 and K>=3 validation rows are inside their accepted numerical envelope
and use the current non-disclosive product route. Any legacy route mentioned in
the package is excluded from this evidence path.

