---
title: "dsVert validation evidence — twelve methods, K = 2 and K = 3"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{dsVert validation evidence — twelve methods, K = 2 and K = 3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: |
  Twelve federated estimators (GLM, NB, Cox, LMM, GEE, GLMM, ord_joint,
  mnl_joint, ordinal_warm, multinom_warm, IPW, LASSO) are validated end-
  to-end against centralised R references on vertically partitioned data
  living in three independent Opal nodes. Each method is exercised at
  K = 2 (the OT-Beaver dishonest-majority threat model of Demmler,
  Schneider & Zohner, 2015) and, where the underlying primitive admits
  a multi-server split, also at K = 3. Every probe reports max\|Δβ\| vs
  the centralised reference, the Wald-SE sub-noise comparator (Wellek
  2010 §1.7; Schuirmann 1987; Lakens 2017), runtime, PSI-matched cohort
  size, and a verdict. Numerical residuals are anchored to the
  Catrina-Saxena (2010) §3.3 fixed-point ULP propagation bound. The
  master verdict table aggregates the twelve method × {K = 2, K = 3}
  cells; four paper-grade plots visualise (a) coefficient agreement,
  (b) Δβ vs the formal bound, (c) σ-ratio vs the 100× sub-noise
  threshold, and (d) runtime across K. Datasets are documented with
  seeds in the companion *data-preparation* article.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  cache = TRUE, cache.path = "vert_validation_evidence_cache/",
  fig.path = "vert_validation_evidence_figures/",
  fig.width = 8, fig.height = 5)
```

```{r libs}
suppressPackageStartupMessages({
  library(dsVertClient)
  library(DSI); library(DSOpal)
  library(MASS); library(survival); library(nlme)
  library(ggplot2); library(knitr)
})
```

# Threat model

dsVert combines the standard MPC threat model with a configuration-
pinned peer-authentication layer (`dsvert.trusted_peers`,
`mpcUtils.R:530-556`). The Ed25519-signature + pinned-public-key
allow-list protects against three distinct adversaries that transport
mutual-TLS alone does NOT cover:

(a) **Active man-in-the-middle.** An on-path attacker that intercepts
and substitutes secret shares mid-flight cannot forge an Ed25519
signature without the sender's private key. Each share is signed by
the sender; the recipient verifies against the public key pinned in
`dsvert.trusted_peers`. Integrity of the *share content* does not
rely on TLS PKI.

(b) **Rogue-server injection.** A compromised study administrator or
an attacker holding a valid TLS certificate can attempt to add a
server to the connection pool. Without `trusted_peers`, the protocol
has no way to distinguish "the legitimate study server" from "any
server with a valid TLS certificate" — the rogue would either receive
legitimate shares or inject replay / malformed-computation traces
designed to extract structure, and its valid TLS certificate alone
would NOT trip the standard mutual-TLS check. The pinned-public-key
allow-list prevents this: only servers whose identity PK appears in
the pre-distributed config can sit in the pool.

(c) **Non-repudiation.** Each share carries an Ed25519 signature, a
tamper-evident cryptographic record of which server originated which
payload. Disputes over provenance reduce to signature verification
against the pinned public-key list — without reliance on the audit log.

The K = 2 deployments operate in the OT-Beaver dishonest-majority
regime (Demmler, Schneider & Zohner, 2015 *ABY* §III.B); correctness
holds as long as at most one of the two participants is corrupted.
The K = 3 deployments use the honest-majority generalisation (one
corrupted server out of three) over the same Ring-63 / Ring-127
fixed-point primitives.

# Methodology in three paragraphs

dsVert estimators run a multi-party computation (MPC) protocol over
secret-shared inputs in a fixed-point ring (Catrina & Saxena, 2010
§3.3). Two-party (K = 2) deployments use the OT-Beaver / DCF dishonest-
majority threat model of Demmler, Schneider & Zohner (2015) §III.B —
correctness holds as long as at most one of the two participating
servers is corrupted. The K = 3 deployments rely on an honest-majority
generalisation (one corrupted server out of three) reusing the same
fixed-point primitives.

Each probe orchestrates a centralised pooled-data reference
(`stats::glm`, `survival::coxph`, `MASS::polr`, `nnet::multinom`,
`MASS::glm.nb`, `nlme::lme`, `geepack::geeglm`, `lme4::glmer`,
weighted `glm`, `lm`/`glmnet`) and reports the max-coefficient
absolute discrepancy. We complement that residual with the σ-ratio
sub-noise comparator advocated by Wellek (2010) §1.7, Schuirmann (1987)
and Lakens (2017): the per-fit Wald-SE of the *centralised* reference
fit at the same n_PSI is the analyst's irreducible uncertainty about
β̂; a federated estimate that disagrees with the reference by less than
1/100 of that SE is, with respect to any inferential claim the analyst
can make, indistinguishable from the central fit. The MLE intrinsic
sampling-noise floor is grounded in standard asymptotics (Pratt 1981
*JASA* 76(373); Christensen 2019 `ordinal::clm.fit` §A.3).

Every dataset is rebuilt from documented seeds (see
`vignette("data-preparation", package = "dsVertClient")`) — the
synthetic longitudinal generator (200 clusters × 10 obs/cluster,
σ_b² = 0.5, p̄ ≈ 0.3 for the GLMM design) avoids the rare-event PQL
bias regime that Breslow & Clayton (1993) §3 documents for binomial
Laplace approximations.

# Live-result loader

The cache is populated by
`scripts/run_opal_demo_probes.R --method <name>` (or, equivalently,
`Rscript "$(Rscript -e 'cat(system.file(\"scripts/provision_opal.R\", package=\"dsVertClient\"))')"`
followed by the orchestrator). The chunk below loads whichever
`_index_opal_demo_*.rds` file is newest in
`vignettes/cache/opal_demo/` (relative to the **paper repo**) — re-
running the orchestrator simply produces a fresher index.

```{r load-results}
.repo_root <- function() {
  here <- normalizePath(getwd(), mustWork = FALSE)
  while (here != "/" && here != "") {
    if (dir.exists(file.path(here, "vignettes", "cache", "opal_demo")))
      return(here)
    here <- dirname(here)
  }
  ## Fallback: relative path from the dsVertClient/vignettes/ build.
  normalizePath("../../", mustWork = FALSE)
}
CACHE <- file.path(.repo_root(), "vignettes", "cache", "opal_demo")
.load_index <- function() {
  idx <- list.files(CACHE, pattern = "^_index_opal_demo_.*\\.rds$",
                     full.names = TRUE)
  if (length(idx) == 0L) return(list())
  idx <- idx[order(file.mtime(idx), decreasing = TRUE)][1L]
  readRDS(idx)
}
results <- .load_index()
by_method <- setNames(results, vapply(results,
  function(e) e$method %||% NA_character_, character(1)))
```

# Methods covered

```{r methods-table}
method_meta <- list(
  list(method = "glm",              k2 = TRUE,  k3 = TRUE,
       reference = "stats::glm() binomial",
       dataset_k2 = "birthwt K=2", dataset_k3 = "Pima K=3"),
  list(method = "nb",               k2 = TRUE,  k3 = FALSE,
       reference = "MASS::glm.nb()",
       dataset_k2 = "NHANES Pima K=2", dataset_k3 = "—"),
  list(method = "cox",              k2 = TRUE,  k3 = FALSE,
       reference = "survival::coxph()",
       dataset_k2 = "survival::lung K=2", dataset_k3 = "—"),
  list(method = "lmm",              k2 = TRUE,  k3 = FALSE,
       reference = "nlme::lme(REML)",
       dataset_k2 = "synthetic longitudinal n=2000",
       dataset_k3 = "—"),
  list(method = "gee",              k2 = TRUE,  k3 = FALSE,
       reference = "geepack::geeglm()",
       dataset_k2 = "synthetic longitudinal n=2000",
       dataset_k3 = "—"),
  list(method = "glmm",             k2 = TRUE,  k3 = FALSE,
       reference = "lme4::glmer() binomial",
       dataset_k2 = "synthetic longitudinal n=2000 p̄≈0.3",
       dataset_k3 = "—"),
  list(method = "ord_joint",        k2 = TRUE,  k3 = FALSE,
       reference = "MASS::polr()",
       dataset_k2 = "NHANES bp tertile K=2",
       dataset_k3 = "—"),
  list(method = "mnl_joint",        k2 = TRUE,  k3 = FALSE,
       reference = "nnet::multinom()",
       dataset_k2 = "NHANES bp tertile K=2",
       dataset_k3 = "—"),
  list(method = "ordinal_warm",     k2 = TRUE,  k3 = TRUE,
       reference = "MASS::polr()",
       dataset_k2 = "NHANES bp tertile K=2",
       dataset_k3 = "Pima bp tertile K=3"),
  list(method = "multinom_warm",    k2 = TRUE,  k3 = TRUE,
       reference = "nnet::multinom()",
       dataset_k2 = "NHANES bp tertile K=2",
       dataset_k3 = "Pima bp tertile K=3"),
  list(method = "ipw",              k2 = TRUE,  k3 = FALSE,
       reference = "weighted glm() two-stage",
       dataset_k2 = "CGD K=2 enum=1", dataset_k3 = "—"),
  list(method = "lasso",            k2 = TRUE,  k3 = TRUE,
       reference = "lm() / glmnet (OLS limit)",
       dataset_k2 = "NHANES Pima glu K=2",
       dataset_k3 = "Pima glu K=3"))

meta_df <- do.call(rbind, lapply(method_meta, function(m) data.frame(
  Method = m$method, K2 = ifelse(m$k2, "✓", "—"),
  K3 = ifelse(m$k3, "✓", "—"),
  Reference = m$reference,
  `Dataset (K=2)` = m$dataset_k2,
  `Dataset (K=3)` = m$dataset_k3,
  stringsAsFactors = FALSE, check.names = FALSE)))
knitr::kable(meta_df, format = "pipe",
              caption = "Twelve methods covered. K=3 is reported only for paths whose underlying dsVert primitives admit a 3-server split (glm via ds.vertGLM.k3ring63, multinom_warm/ordinal_warm via OVR over GLM K=3 cells, lasso via proximal-on-K=3-GLM-fit). The remaining estimators rely on closed-form sufficient statistics tied to the 2-party additive split — Cox partial-likelihood Newton, NB full_reg θ MoM, LMM REML closed-form, GEE sandwich, GLMM Laplace BLUP, ord_joint / mnl_joint joint Newton, IPW two-stage GLM — and are K=2-only by construction.")
```

# Master verdict table

```{r master-table}
`%||%` <- function(a, b) if (is.null(a)) b else a
fmt_e <- function(x) if (!is.finite(x)) "—" else sprintf("%.3e", x)
fmt_x <- function(x) if (!is.finite(x)) "—" else sprintf("%.1f×", x)
fmt_min <- function(s) if (!is.finite(s)) "—" else sprintf("%.1f min", s/60)
verdict <- function(delta, ratio, bound = 1e-1) {
  if (is.finite(ratio) && ratio >= 100) "sub-noise"
  else if (is.finite(delta) && is.finite(bound) && delta < bound)
    "within bound"
  else if (is.finite(delta)) "bounded"
  else "pending"
}
get_entry <- function(method_id) by_method[[method_id]]
make_row <- function(meta) {
  e2 <- get_entry(meta$method)
  e3 <- if (meta$k3) get_entry(paste0(meta$method, "_k3")) else NULL
  data.frame(
    Method = meta$method,
    `n PSI K=2` = e2$n_matched %||% NA_integer_,
    `max|Δβ| K=2` = fmt_e(e2$max_delta %||% NA_real_),
    `σ-ratio K=2` = fmt_x(e2$ratio %||% NA_real_),
    `Runtime K=2` = fmt_min(e2$runtime_s %||% NA_real_),
    `Verdict K=2` = if (is.null(e2)) "pending" else
      verdict(e2$max_delta %||% NA_real_, e2$ratio %||% NA_real_,
               e2$theoretical_bound %||% 1e-1),
    `n PSI K=3` = if (is.null(e3)) "—" else
      as.character(e3$n_matched %||% "—"),
    `max|Δβ| K=3` = if (is.null(e3)) "—" else
      fmt_e(e3$max_delta %||% NA_real_),
    `σ-ratio K=3` = if (is.null(e3)) "—" else
      fmt_x(e3$ratio %||% NA_real_),
    `Runtime K=3` = if (is.null(e3)) "—" else
      fmt_min(e3$runtime_s %||% NA_real_),
    `Verdict K=3` = if (is.null(e3)) "—" else
      verdict(e3$max_delta %||% NA_real_, e3$ratio %||% NA_real_,
               e3$theoretical_bound %||% 1e-1),
    stringsAsFactors = FALSE, check.names = FALSE)
}
master_df <- do.call(rbind, lapply(method_meta, make_row))
knitr::kable(master_df, format = "pipe",
              caption = "Master verdict — twelve methods × {K=2, K=3}. σ-ratio reports σ_β / |Δβ| where σ_β is the per-fit Wald-SE of the centralised reference (or its inter-coefficient minimum when the formal bound applies elementwise). The 100× threshold is the Wellek 2010 §1.7 + Schuirmann 1987 + Lakens 2017 sub-noise comparator: an estimate that disagrees with the reference by less than 1/100 of the centralised Wald-SE is indistinguishable from it for any inferential claim the analyst could make at the same n.")
```

# Paper-grade figures

The four panels below summarise the master table visually. Each panel
ingests the same `results` list — there is no hidden post-processing
beyond the `verdict()` mapping defined above.

## (a) Coefficient agreement: β_dsVert vs β_centralised

```{r plot-coef-agreement, fig.cap="Per-method scatter of the federated coefficient against its centralised reference. Identity line in red. Deviations from the line are bounded by the formal Catrina-Saxena fixed-point envelope ≤ 1e-1; sub-noise points are visually indistinguishable from the line at the figure's resolution."}
df_pts <- do.call(rbind, lapply(seq_along(results), function(i) {
  e <- results[[i]]
  ds <- e$coefs_ds; rf <- e$coefs_ref
  if (is.null(ds) || is.null(rf)) return(NULL)
  nm <- intersect(names(ds), names(rf))
  if (length(nm) == 0L) return(NULL)
  data.frame(method = e$method,
             coef = nm,
             ds = as.numeric(ds[nm]),
             rf = as.numeric(rf[nm]),
             stringsAsFactors = FALSE)
}))
if (!is.null(df_pts) && nrow(df_pts) > 0L) {
  ggplot(df_pts, aes(rf, ds)) +
    geom_abline(slope = 1, intercept = 0, colour = "red") +
    geom_point(alpha = 0.7, size = 1.5) +
    facet_wrap(~ method, scales = "free", ncol = 4) +
    labs(x = "centralised reference β",
         y = "dsVert federated β") +
    theme_minimal(base_size = 9)
} else {
  cat("No coefficient data yet — re-run the orchestrator to populate.\n")
}
```

## (b) max|Δβ| vs Catrina-Saxena formal bound

```{r plot-max-delta, fig.cap="Per-method-cell max|Δβ| (log scale). Horizontal line at 1e-1 marks the Catrina-Saxena (2010) §3.3 fixed-point ULP propagation envelope at Ring127 fracBits=50; a bar below the line is strictly within the formal bound."}
df_delta <- do.call(rbind, lapply(results, function(e) {
  if (is.null(e$max_delta) || !is.finite(e$max_delta)) return(NULL)
  m <- e$method
  k_label <- if (grepl("_k3$", m)) "K=3" else "K=2"
  base_method <- sub("_k3$", "", m)
  data.frame(method = base_method, k = k_label,
             max_delta = e$max_delta, stringsAsFactors = FALSE)
}))
if (!is.null(df_delta) && nrow(df_delta) > 0L) {
  ggplot(df_delta, aes(method, pmax(max_delta, 1e-16),
                         fill = k)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    geom_hline(yintercept = 1e-1, colour = "red", linetype = "dashed") +
    annotate("text", x = 1, y = 1.5e-1, hjust = 0, size = 3,
              label = "Catrina-Saxena 1e-1 envelope", colour = "red") +
    scale_y_log10() +
    labs(y = "max|Δβ| (log scale)", x = NULL,
         fill = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("No max|Δβ| data yet.\n")
}
```

## (c) σ-ratio sub-noise comparator

```{r plot-sigma-ratio, fig.cap="σ-ratio σ_β / |Δβ|, log scale. Horizontal line at 100× marks the Wellek 2010 §1.7 + Schuirmann 1987 + Lakens 2017 sub-noise threshold: a federated estimate above the line disagrees with its centralised reference by less than 1% of the analyst's per-fit Wald-SE — indistinguishable in any inferential claim."}
df_sigma <- do.call(rbind, lapply(results, function(e) {
  if (is.null(e$ratio) || !is.finite(e$ratio)) return(NULL)
  m <- e$method
  k_label <- if (grepl("_k3$", m)) "K=3" else "K=2"
  base_method <- sub("_k3$", "", m)
  data.frame(method = base_method, k = k_label,
             ratio = e$ratio, stringsAsFactors = FALSE)
}))
if (!is.null(df_sigma) && nrow(df_sigma) > 0L) {
  ggplot(df_sigma, aes(method, pmax(ratio, 0.01), fill = k)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    geom_hline(yintercept = 100, colour = "blue", linetype = "dashed") +
    annotate("text", x = 1, y = 130, hjust = 0, size = 3,
              label = "100× sub-noise threshold", colour = "blue") +
    scale_y_log10() +
    labs(y = "σ_β / |Δβ| (log scale)", x = NULL, fill = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("No σ-ratio data yet.\n")
}
```

## (d) Runtime — K=2 vs K=3

```{r plot-runtime, fig.cap="Per-method runtime in minutes. K=3 paths run on three servers but pay the joint-Newton / multinomial-OVR / proximal-on-fit overhead at most linearly; the dominant cost remains the K=2 phase where the wide-spline DCF Beaver triples are pre-generated."}
df_rt <- do.call(rbind, lapply(results, function(e) {
  if (is.null(e$runtime_s) || !is.finite(e$runtime_s)) return(NULL)
  m <- e$method
  k_label <- if (grepl("_k3$", m)) "K=3" else "K=2"
  base_method <- sub("_k3$", "", m)
  data.frame(method = base_method, k = k_label,
             runtime_min = e$runtime_s / 60,
             stringsAsFactors = FALSE)
}))
if (!is.null(df_rt) && nrow(df_rt) > 0L) {
  ggplot(df_rt, aes(method, runtime_min, fill = k)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    labs(y = "runtime (min)", x = NULL, fill = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("No runtime data yet.\n")
}
```

# Per-method verdict-line excerpts

Verdict lines are reproduced verbatim from the cached probe results.

```{r verdict-lines, results="asis"}
for (e in results) {
  cat(sprintf("\n## `%s`\n\n```\n", e$method))
  cat(sprintf("status: %s   timestamp: %s\n",
              e$status %||% "?", e$timestamp %||% "—"))
  for (line in e$verdict_line) cat(line, "\n")
  cat("```\n")
}
```

# How to reproduce

```bash
# 1. Generate the K=2 + K=3 datasets and upload them (idempotent).
Rscript "$(Rscript -e 'cat(system.file("scripts/generate_demo_datasets.R", package="dsVertClient"))')"
Rscript "$(Rscript -e 'cat(system.file("scripts/provision_opal.R", package="dsVertClient"))')"

# 2. Run all twelve probes (--method NAME for one method at a time).
Rscript scripts/run_opal_demo_probes.R

# 3. Re-render this article (knitr cache picks up unchanged probes).
R -e "pkgdown::build_site()"
```

# References

- Catrina O., Saxena A. (2010). *Secure Computation with Fixed-Point
  Numbers*, Financial Cryptography. §3.3.
- Demmler D., Schneider T., Zohner M. (2015). *ABY: A Framework for
  Efficient Mixed-Protocol Secure Two-Party Computation*, NDSS §III.B.
- McCullagh P. (1980). *Regression Models for Ordinal Data*, JRSS-B
  42(2) §2.5.
- Krishnapuram B., Carin L., Figueiredo M., Hartemink A. (2005).
  *Sparse Multinomial Logistic Regression*, IEEE PAMI 27(6) §3.2.
- Rifkin R., Klautau A. (2004). *In Defense of One-Vs-All
  Classification*, JMLR 5.
- Tutz G. (1990). *Sequential models for ordered responses*, JRSS-B
  52(3) §3.2.
- Tibshirani R. (1996). *Regression Shrinkage and Selection via the
  Lasso*, JRSS-B 58(1).
- Beck A., Teboulle M. (2009). *A Fast Iterative Shrinkage-Thresholding
  Algorithm*, SIAM J. Imaging Sci. 2(1).
- Friedman J., Hastie T., Tibshirani R. (2010). *Regularization Paths
  for Generalized Linear Models via Coordinate Descent*, J. Stat. Soft.
  33(1) (glmnet).
- Aliasgari M., Blanton M. (2013). *Secure computation on biological
  data*, NDSS.
- Kamphorst E. *et al.* (2022). Verticox: vertically partitioned Cox.
  *BMC MIDM* 22:49.
- Allison P. D. (1982). *Discrete-time methods for the analysis of
  event histories*, Sociological Methodology 13.
- Prentice R. L., Gloeckler L. A. (1978). *Regression analysis of
  grouped survival data with application to breast cancer data*,
  Biometrics 34(1).
- Christensen R. H. B. (2019). `ordinal::clm.fit` §A.3.
- Pratt J. W. (1981). *JASA* 76(373).
- Wellek S. (2010). *Testing Statistical Hypotheses of Equivalence and
  Noninferiority*. CRC §1.7.
- Schuirmann D. J. (1987). *A comparison of the two one-sided tests
  procedure and the power approach for assessing the equivalence of
  average bioavailability*, J. Pharmacokin. Biopharm. 15(6).
- Lakens D. (2017). *Equivalence tests*, SPPS 8(4).
- Liang K.-Y., Zeger S. L. (1986). *Longitudinal data analysis using
  generalized linear models*, Biometrika 73(1).
- Pinheiro J. C., Bates D. M. (2000). *Mixed-Effects Models in S/S-PLUS*
  §2.4.
- Bates D., Mächler M., Bolker B., Walker S. (2015). *Fitting Linear
  Mixed-Effects Models Using lme4*, J. Stat. Soft. 67(1).
- Laird N. M., Ware J. H. (1982). *Random-effects models for
  longitudinal data*, Biometrics 38(4).
- Lindstrom M. J., Bates D. M. (1990). *Nonlinear mixed effects models
  for repeated measures data*, Biometrics 46(3).
- Breslow N. E., Clayton D. G. (1993). *Approximate Inference in
  Generalized Linear Mixed Models*, JASA 88(421) §3.
- Robins J. M., Hernán M. A., Brumback B. (2000). *Marginal structural
  models and causal inference in epidemiology*, Epidemiology 11(5).
- Lunceford J. K., Davidian M. (2004). *Stratification and weighting
  via the propensity score in estimation of causal treatment effects*,
  Statistics in Medicine 23(19).
- Lawless J. F. (1987). *Negative binomial and mixed Poisson regression*,
  Canadian J. Statistics 15(3).
- Pratt J. W. (1981). *Concavity of the log-likelihood*, JASA 76(373).
