---
title: "dsVert validation evidence"
author: "David Sarrat Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{dsVert validation evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
  message = FALSE, collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
  file.path("..", "validation_helpers.R"),
  file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
force_run <- identical(tolower(Sys.getenv("DSVERT_VALIDATION_FORCE", "false")), "true")
```

## Scope

This is the executable validation index for the current dsVertClient
product surface. It covers every supported method in K=2 and K>=3 mode.
Each method page documents the statistical target, fixture, centralized
reference, DSLite execution path, and disclosure surface.

No external RDS files or private repository caches are required. On a clean
checkout, rendering this page builds all fixtures and runs the DSLite
validations. Local cache files under `vignettes/validation-cache/` are
generated by this same code path and can be deleted safely.

## Summary table

```{r summary}
all_rows <- run_all_validations(force = force_run)
assert_validation(all_rows)
knitr::kable(all_rows[, c("method_name", "k_mode", "function_route",
  "primary_metric", "observed", "tolerance", "tier", "status",
  "runtime_s")])
```

## Method pages

- [PSI alignment validation](validation_psi.html)
- [Descriptive statistics validation](validation_descriptive.html)
- [Contingency tests validation](validation_contingency.html)
- [Correlation validation](validation_correlation.html)
- [PCA validation](validation_pca.html)
- [GLM validation](validation_glm.html)
- [GLM inference validation](validation_inference.html)
- [LASSO validation](validation_lasso.html)
- [Negative binomial validation](validation_negative_binomial.html)
- [Cox PH validation](validation_cox.html)
- [LMM validation](validation_lmm.html)
- [GEE validation](validation_gee.html)
- [GLMM validation](validation_glmm.html)
- [IPW validation](validation_ipw.html)
- [Multiple imputation validation](validation_mi.html)
- [Multinomial validation](validation_multinomial.html)
- [Ordinal validation](validation_ordinal.html)

## Removed legacy routes

The package should not expose routes that were disclosive or materially
less accurate than an available product route. This check is executed
during rendering.

```{r legacy-removed}
legacy <- assert_legacy_routes_removed()
knitr::kable(legacy)
```
