---
title: "dsVert validation evidence"
author: "David Sarrat González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{dsVert validation evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>")
helper_candidates <- c("validation_helpers.R",
                       file.path("..", "validation_helpers.R"),
                       file.path("vignettes", "validation_helpers.R"))
source(helper_candidates[file.exists(helper_candidates)][1])
```

## Scope

This page is the validation index for the current dsVertClient product surface.
The historical vignettes under `vignettes/methods/` were discarded because they
mixed product methods with legacy or diagnostic routes. The pages linked here
document the current supported route for each method family, with K=2 and K>=3
evidence from deterministic DSLite harnesses.

Each method vignette contains:

- the statistical target and mathematical estimating equation;
- the centralized reference construction;
- the vertical DSLite split and product function call;
- a disclosure review describing what the analyst and peer servers can see;
- an executed evidence check against the cached DSLite result table.

## Summary table

```{r summary}
all_rows <- load_validation_summary()
assert_validation(all_rows)
knitr::kable(all_rows[, c("method_name", "k_mode", "function_route",
                          "primary_metric", "observed", "tolerance",
                          "tier", "status", "cache")])
```

## Method pages

- [PSI alignment](validation_psi.html)
- [Descriptive statistics](validation_descriptive.html)
- [Contingency tests](validation_contingency.html)
- [Correlation](validation_correlation.html)
- [PCA](validation_pca.html)
- [GLM](validation_glm.html)
- [GLM inference helpers](validation_inference.html)
- [LASSO](validation_lasso.html)
- [Negative binomial](validation_negative_binomial.html)
- [Cox PH](validation_cox.html)
- [LMM](validation_lmm.html)
- [GEE](validation_gee.html)
- [GLMM](validation_glmm.html)
- [IPW](validation_ipw.html)
- [Multiple imputation](validation_mi.html)
- [Multinomial](validation_multinomial.html)
- [Ordinal](validation_ordinal.html)

## Removed legacy routes

The package should not expose routes that are disclosive or materially less
accurate than an available product route. This check is executed during
rendering.

```{r legacy-removed}
library(dsVertClient)
legacy <- load_legacy_removed()
checks <- assert_legacy_routes_removed()
knitr::kable(merge(legacy, checks, by.x = "route", by.y = "route", all.x = TRUE))
```

## Interpretation

All rows in the current table are `PASS`, non-disclosive under the project
standard, and cover both K=2 and K>=3. Some methods are exact to their
centralized target; others are strict-practical because the remaining gap is a
documented secure approximation floor, such as histogram quantile bucketization,
secure sigmoid spline resolution, PQL versus GLMM ML, or penalized optimizer
depth.

