---
title: "Cox K=3 — federated Allison-1982 Poisson trick over GLM K=3"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cox K=3 — federated Allison-1982 Poisson trick over GLM K=3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

The Allison (1982) / Whitehead (1980) / Prentice & Gloeckler
(1978) §2 equivalence shows that the Cox partial-likelihood
estimator coincides with the maximum-likelihood Poisson regression
on **person-time-expanded data** with one row per
(subject, at-risk-interval) pair, log-link, an offset
\(\log \Delta t_j\) for the within-interval at-risk width, and a
piecewise-constant baseline hazard \(\alpha_t\) included as a
factor covariate. The slopes \(\beta\) of the Poisson fit are the
Cox PH slopes asymptotically (and exactly under fully-tied data).

`ds.vertCox.k3` reduces to a single `ds.vertGLM` call on the
K = 3 secure-aggregation path with `family = "poisson"` and an
offset column on the label server. No new MPC primitives — full
inheritance of the K = 3 GLM threat model. The companion
`isglobal-brge/dsVert#worker2-k3-restore` SEGV-guard patch protects
against L-BFGS divergence in the wide-spline DCF when the offset
+ baseline-dummy design pushes η out of the spline range.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates and per-patient time-event records never leave their origin server. |
| **D-INV-2** | Per-row \(x_{ij}^{\top}\beta + \alpha_{t}\) never reconstructed in plaintext. |
| **D-INV-3** | Per-patient interval indicators / event flags exposed only as aggregate Poisson scores. |
| **D-INV-4** | Time grid + at-risk + event indicators live on the label server. |
| **D-INV-5** | Cross-server messages signed Ed25519 against `dsvert.trusted_peers`. |

### Dataset and partition

The lung-cancer NCCTG data, person-time-expanded into J quantile-
binned event-time intervals. Each subject contributes one row per
interval at-risk, plus a 0/1 event indicator and an at-risk-width
offset. We use J = 4 (quartile event-time bins) with a deliberately
small synthetic Cox DGP to keep DSLite render time bounded; the
production-equivalent K = 3 lung pp-expansion verdict at n = 131
× J = 8 lives in the validation cache (see
`scripts/run_opal_demo_probes.R::probe_cox_k3` against the local
opal docker cluster).

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `pp_row_id`, `x1` | Continuous covariate |
| **s2** | `pp_row_id`, `x2` | Binary covariate |
| **s3** | `pp_row_id`, `t_bin`, `log_width`, `event`, `x3`, `tb2..tbJ` | Outcome + baseline dummies (label) |

Each pp_row_id encodes a unique (subject, interval) pair —
PSI alignment happens on pp rows so each server sees the SAME
person-time replicate set.

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(survival)
library(ggplot2)
```

```{r data}
n_subj <- 200L
J      <- 4L
x1 <- rnorm(n_subj)
x2 <- rbinom(n_subj, 1L, 0.5)
x3 <- rnorm(n_subj, 0, 0.3)
true_beta  <- c(0.4, -0.3, 0.5)
hazard_lin <- as.numeric(cbind(x1, x2, x3) %*% true_beta)
time   <- rexp(n_subj, rate = 0.05 * exp(hazard_lin))
cens   <- rexp(n_subj, rate = 0.04)
status <- as.integer(time <= cens)
time   <- pmin(time, cens)

df <- data.frame(
  patient_id = sprintf("S%05d", seq_len(n_subj)),
  time = time, status = status,
  x1 = x1, x2 = x2, x3 = x3)

breaks <- c(0,
            quantile(df$time[df$status == 1L],
                      probs = (1:(J - 1L)) / J),
            Inf)

expand_pt <- function(df) {
  out <- vector("list", nrow(df))
  for (i in seq_len(nrow(df))) {
    pid <- df$patient_id[i]
    t_i <- df$time[i]
    d_i <- df$status[i]
    rows <- list()
    for (j in seq_len(J)) {
      lo <- breaks[j]; hi <- breaks[j + 1L]
      if (t_i <= lo) break
      bin_top <- min(hi, t_i)
      width <- bin_top - lo
      if (width <= 0) next
      rows[[length(rows) + 1L]] <- data.frame(
        pp_row_id = sprintf("PT-%s-%02d", pid, j),
        patient_id = pid,
        t_bin = j,
        log_width = log(width),
        event = if (d_i == 1L && t_i > lo && t_i <= hi) 1L else 0L)
    }
    out[[i]] <- do.call(rbind, rows)
  }
  do.call(rbind, out)
}
pp <- expand_pt(df)
cov <- df[, c("patient_id", "x1", "x2", "x3")]
pp <- merge(pp, cov, by = "patient_id")
for (jj in 2:J)
  pp[[sprintf("tb%d", jj)]] <- as.integer(pp$t_bin == jj)
```

```{r pooled}
pooled <- pp[, c("pp_row_id", "patient_id", "t_bin",
                  "log_width", "event",
                  "x1", "x2", "x3",
                  paste0("tb", 2:J))]
head(pooled)
```

```{r split}
s1 <- pooled[, c("pp_row_id", "x1")]
s2 <- pooled[, c("pp_row_id", "x2")]
s3 <- pooled[, c("pp_row_id", "t_bin", "log_width", "event", "x3",
                 paste0("tb", 2:J))]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "pp_row_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated Cox K=3 (Allison-1982 Poisson trick)

```{r ds-cox-k3}
baseline_dummies <- paste0("tb", 2:J)
poisson_formula <- as.formula(sprintf(
  "event ~ x1 + x2 + x3 + %s",
  paste(baseline_dummies, collapse = " + ")))

fit_ds_full <- ds.vertGLM(
  poisson_formula, data = "DA",
  family = "poisson", offset = "log_width",
  verbose = FALSE, datasources = conns)

slope_names <- c("x1", "x2", "x3")
fit_ds <- list(
  coefficients = fit_ds_full$coefficients[slope_names])
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- coxph(
  Surv(time, status) ~ x1 + x2 + x3, data = df)
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
ds_b <- as.numeric(fit_ds$coefficients[slope_names])
rf_b <- as.numeric(coef(fit_ref)[slope_names])
delta <- data.frame(
  coef   = slope_names,
  true   = true_beta,
  dsVert = ds_b,
  coxph  = rf_b)
delta$abs <- abs(delta$dsVert - delta$coxph)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(vcov(fit_ref))))
sigma_ratio <- sigma_min / max_abs

c(max_abs_delta = max_abs,
  sigma_min     = sigma_min,
  sigma_ratio   = sigma_ratio)
```

```{r scatter, fig.cap = "Federated Cox K=3 (Allison Poisson trick) slopes vs centralised survival::coxph fit. Dashed identity line."}
ggplot(delta, aes(coxph, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[coxph]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "cox_k3",
  dataset         = sprintf("Synthetic Cox DGP (n_subj=%d, J=%d)",
                             n_subj, J),
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "cox_k3.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The Allison-Poisson-trick Cox K=3 estimator agrees with
`survival::coxph` to a maximum absolute slope deviation of
`r sprintf("%.2e", max_abs)` (σ-ratio
`r sprintf("%.1f", sigma_ratio)`× against the smallest centralised
Wald standard error). Discrete-time approximation introduces
\(O(\Delta t)\) bias relative to continuous-time Cox; with quartile
bins on this DGP the bias sits inside the K = 3 secure-agg
fixed-point envelope.

### References

* Allison, P. D. (1982). Discrete-time methods for the analysis of
  event histories. *Sociological Methodology* 13, 61–98.
* Andreux, M. *et al.* (2020). Federated survival analysis with
  discrete-time Cox models. arXiv:2006.08997.
* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Prentice, R. L. & Gloeckler, L. A. (1978). Regression analysis of
  grouped survival data. *Biometrics* 34(1), 57–67.
* Whitehead, J. (1980). Fitting Cox's regression model to survival
  data using GLIM. *Applied Statistics* 29(3), 268–275.
