---
title: "GEE — federated Liang–Zeger sandwich estimator (K = 3)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GEE — federated Liang–Zeger sandwich estimator (K = 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

Generalised estimating equations (GEE) extend GLM to clustered or
longitudinal outcomes by specifying a working correlation structure
\(R(\alpha)\) on within-cluster errors and solving the quasi-score

\[
U(\beta) \;=\; \sum_{i=1}^K
  D_i^{\top} V_i^{-1}\bigl(y_i - \mu_i(\beta)\bigr) \;=\; 0
\]

where \(D_i = \partial \mu_i / \partial \beta\),
\(V_i = A_i^{1/2} R(\alpha) A_i^{1/2}\), and \(A_i\) is the
diagonal of GLM-implied marginal variances. Liang & Zeger (1986)
showed that \(\hat\beta\) is consistent and asymptotically normal
even under mis-specified \(R(\alpha)\), with covariance estimated by
the sandwich \(\widehat{\operatorname{Cov}}(\hat\beta) =
B^{-1} M B^{-1}\) where \(B = \sum D_i^{\top} V_i^{-1} D_i\) and
\(M = \sum D_i^{\top} V_i^{-1} (y_i - \mu_i)(y_i - \mu_i)^{\top}
V_i^{-1} D_i\).

In the federated K = 3 setting the design partitions across servers;
the cluster id and outcome live on the label server. `ds.vertGEE`
runs an alternating fixed-effects IRLS / working-correlation moment
update and returns the sandwich-corrected coefficients consistent
with `geepack::geeglm`.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-row \(x_i^{\top}\beta\) is never reconstructed in plaintext. |
| **D-INV-3** | Per-patient residuals are never revealed; only aggregate \(D_i^{\top} V_i^{-1} (y_i - \mu_i)\) blocks are. |
| **D-INV-4** | Cluster identifiers are visible only on the label server. |
| **D-INV-5** | All cross-server messages signed Ed25519, peer pinned via `dsvert.trusted_peers`. |

### Dataset and partition

The Pothoff–Roy *Orthodont* dataset (Pinheiro & Bates, 2000 §1.4)
records four longitudinal dental measurements per child across ages
8, 10, 12, 14 — 27 children × 4 visits = 108 rows. The repeated
measures within child are clustered, motivating GEE.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `age` | Time covariate |
| **s2** | `patient_id`, `sex_num` | Sex covariate |
| **s3** | `patient_id`, `cluster_id`, `distance` | Cluster id + outcome |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(nlme)
library(geepack)
library(ggplot2)
```

```{r data}
data("Orthodont", package = "nlme")
od <- as.data.frame(Orthodont)
od$patient_id <- sprintf("R%03d", seq_len(nrow(od)))
od$cluster_id <- as.character(od$Subject)
od$sex_num    <- as.integer(od$Sex == "Male")
```

```{r pooled}
pooled <- od[, c("patient_id", "cluster_id",
                  "age", "sex_num", "distance")]
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "age")]
s2 <- pooled[, c("patient_id", "sex_num")]
s3 <- pooled[, c("patient_id", "cluster_id", "distance")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated GEE

```{r ds-gee}
fit_ds <- ds.vertGEE(
  formula = distance ~ age + sex_num,
  data = "DA", family = "gaussian",
  id_col = "cluster_id", corstr = "exchangeable",
  verbose = FALSE, datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- geepack::geeglm(
  distance ~ age + sex_num,
  id = cluster_id, data = pooled,
  family = gaussian, corstr = "exchangeable")
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
nm <- intersect(names(fit_ds$coefficients), names(coef(fit_ref)))
delta <- data.frame(
  coef    = nm,
  dsVert  = as.numeric(fit_ds$coefficients[nm]),
  geeglm  = as.numeric(coef(fit_ref)[nm]))
delta$abs <- abs(delta$dsVert - delta$geeglm)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(summary(fit_ref)$cov.scaled)))
sigma_ratio <- sigma_min / max_abs

c(max_abs_delta = max_abs,
  sigma_min     = sigma_min,
  sigma_ratio   = sigma_ratio)
```

```{r scatter, fig.cap = "Federated GEE coefficients vs centralised geepack::geeglm fit. Dashed identity line."}
ggplot(delta, aes(geeglm, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[geeglm]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "gee",
  dataset         = "Orthodont (Pothoff-Roy)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "gee.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The federated GEE estimator agrees with `geepack::geeglm` to a
maximum absolute slope deviation of `r sprintf("%.2e", max_abs)`,
giving a σ-ratio of `r sprintf("%.0f", sigma_ratio)`× against the
smallest sandwich-corrected standard error of the centralised fit.
The clustered-error working correlation \(R(\alpha)\) is recovered
within the Ring63 Beaver fixed-point envelope.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Hardin, J. W. & Hilbe, J. M. (2003). *Generalized Estimating
  Equations*. Chapman & Hall.
* Liang, K.-Y. & Zeger, S. L. (1986). Longitudinal data analysis
  using generalised linear models. *Biometrika* 73(1), 13–22.
* Pinheiro, J. C. & Bates, D. M. (2000). *Mixed-Effects Models in
  S/S-PLUS* §1.4.
