---
title: "Binomial GLMM -- aggregate PQL validation (K = 2 and K >= 3)"
author: "David Sarrat Gonzalez, Miron Banjac, Juan R. Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Binomial GLMM -- aggregate PQL validation (K = 2 and K >= 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(20260502L)
```

## 1. Context

A random-intercept binomial GLMM models a binary outcome
\(y_{ij} \in \{0,1\}\) clustered within group \(i\):

\[
\operatorname{logit}\{\Pr(y_{ij}=1 \mid b_i)\} =
x_{ij}^{\top}\beta + b_i,\qquad b_i \sim N(0,\sigma_b^2).
\]

`ds.vertGLMM(method = "pql_aggregate")` uses a PQL working-response
update and solves the random-intercept weighted LMM from aggregate
sufficient statistics. The route is deliberately Ring127-only because
the working statistics require repeated share-domain products and
reciprocals.

The validation below runs the same deterministic dataset as a central
`lme4::glmer` reference and as two DSLite vertical partitions:

| Mode | Split |
|:---|:---|
| K = 2 | `s1`: `x1`, `x2`; `s2`: `cluster_id`, `x3`, `y_bin` |
| K = 3 | `s1`: `x1`; `s2`: `x2`; `s3`: `cluster_id`, `x3`, `y_bin` |

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Fitted probabilities, working responses, PQL weights, residuals and row scores remain server-side or in Ring127 additive shares. |
| **D-INV-3** | The client receives low-dimensional coefficients, variance components, guarded cluster-level PQL sufficient statistics and cluster-level BLUPs, not patient-level vectors. |
| **D-INV-4** | Original cluster labels are not returned; clusters below `datashield.privacyLevel` fail closed. |
| **D-INV-5** | The returned-object audit flags any atomic vector of length \(n\); only cluster-length `b_hat` and `cluster_sizes` are accepted. |

## 2. Data Preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(lme4)
library(ggplot2)
```

```{r data}
build_pooled <- function() {
  set.seed(20260502L)
  n_cluster <- 10L
  cluster_size <- 5L
  n <- n_cluster * cluster_size
  cluster_id <- rep(seq_len(n_cluster), each = cluster_size)
  x1 <- stats::rnorm(n)
  x2 <- stats::rnorm(n)
  x3 <- stats::rnorm(n)
  b_i <- stats::rnorm(n_cluster, sd = 0.65)
  eta <- -0.35 + 0.55 * x1 - 0.45 * x2 + 0.35 * x3 + b_i[cluster_id]
  p <- stats::plogis(eta)
  y_bin <- stats::rbinom(n, 1L, p)
  data.frame(
    patient_id = sprintf("GLMM%04d", seq_len(n)),
    cluster_id = cluster_id,
    x1 = x1,
    x2 = x2,
    x3 = x3,
    y_bin = y_bin)
}

pooled <- build_pooled()
head(pooled)
```

```{r split-functions}
split_tables <- function(pooled, K) {
  if (K == 2L) {
    list(
      s1 = pooled[, c("patient_id", "x1", "x2")],
      s2 = pooled[, c("patient_id", "cluster_id", "x3", "y_bin")]
    )
  } else if (K == 3L) {
    list(
      s1 = pooled[, c("patient_id", "x1")],
      s2 = pooled[, c("patient_id", "x2")],
      s3 = pooled[, c("patient_id", "cluster_id", "x3", "y_bin")]
    )
  } else {
    stop("This vignette validates K = 2 and K = 3 only.")
  }
}

connect_dslite <- function(tables, symbol = "D") {
  server <- DSLite::newDSLiteServer(
    tables = tables,
    config = DSLite::defaultDSConfiguration(include = c("dsBase", "dsVert")))
  dslite.server <<- server
  builder <- DSI::newDSLoginBuilder()
  for (nm in names(tables)) {
    builder$append(server = nm, url = "dslite.server",
                   table = nm, driver = "DSLiteDriver")
  }
  DSI::datashield.login(builder$build(), assign = TRUE,
                         symbol = symbol, opts = list(server = server))
}
```

## 3. Central Reference

```{r reference}
formula <- y_bin ~ x1 + x2 + x3
fit_ref <- suppressWarnings(lme4::glmer(
  update(formula, . ~ . + (1 | cluster_id)),
  data = pooled,
  family = stats::binomial(),
  control = lme4::glmerControl(optimizer = "bobyqa",
                               optCtrl = list(maxfun = 2e5))))

ref_beta <- lme4::fixef(fit_ref)
ref_sigma_b2 <- as.numeric(lme4::VarCorr(fit_ref)$cluster_id)[1L]
ref_beta
ref_sigma_b2
```

## 4. Federated PQL Fits

```{r helpers}
compare_named <- function(a, b) {
  nm <- intersect(names(a), names(b))
  delta <- as.numeric(a[nm]) - as.numeric(b[nm])
  data.frame(
    coef = nm,
    dsVert = as.numeric(a[nm]),
    glmer = as.numeric(b[nm]),
    delta = delta,
    abs_delta = abs(delta),
    row.names = NULL)
}

audit_returned_object <- function(x, n, n_clusters) {
  bad <- character()
  cluster_level <- character()
  walk <- function(obj, path) {
    if (is.null(obj) || is.environment(obj) || is.function(obj) ||
        inherits(obj, "formula") || inherits(obj, "call")) {
      return(invisible(NULL))
    }
    if (is.atomic(obj)) {
      len <- length(obj)
      if (len == n) bad <<- c(bad, path)
      if (len == n_clusters && len != n) {
        cluster_level <<- c(cluster_level, path)
      }
      return(invisible(NULL))
    }
    if (is.matrix(obj) || is.data.frame(obj)) {
      d <- dim(obj)
      if (length(d) == 2L && any(d == n)) bad <<- c(bad, path)
      if (length(d) == 2L && any(d == n_clusters) && !any(d == n)) {
        cluster_level <<- c(cluster_level, path)
      }
    }
    if (is.list(obj)) {
      nm <- names(obj)
      if (is.null(nm)) nm <- as.character(seq_along(obj))
      for (i in seq_along(obj)) walk(obj[[i]], paste0(path, "$", nm[[i]]))
    }
    invisible(NULL)
  }
  walk(x, "fit_ds")
  list(pass = length(unique(bad)) == 0L,
       disallowed_length_n_paths = unique(bad),
       cluster_level_paths = unique(cluster_level),
       n = n,
       n_clusters = n_clusters)
}

fit_one <- function(K) {
  options(dsvert.require_trusted_peers = FALSE,
          datashield.privacyLevel = 5L,
          datashield.errors.print = TRUE)
  conns <- connect_dslite(split_tables(pooled, K))
  on.exit(try(DSI::datashield.logout(conns), silent = TRUE), add = TRUE)
  psi <- ds.psiAlign("D", "patient_id", "DA",
                      datasources = conns, verbose = FALSE)
  t0 <- proc.time()[3]
  fit_ds <- ds.vertGLMM(
    formula = formula,
    data = "DA",
    cluster_col = "cluster_id",
    max_outer = 1L,
    inner_iter = 2L,
    tol = 1e-2,
    method = "pql_aggregate",
    ring = 127L,
    compute_se = FALSE,
    verbose = FALSE,
    datasources = conns)
  runtime_s <- unname(proc.time()[3] - t0)
  cmp <- compare_named(fit_ds$coefficients, ref_beta)
  list(
    K = K,
    n = nrow(pooled),
    n_matched = psi[[1]]$n_matched,
    runtime_s = runtime_s,
    coefficients = cmp,
    max_abs_delta = max(cmp$abs_delta),
    sigma_b2_ds = fit_ds$sigma_b2,
    sigma_b2_ref = ref_sigma_b2,
    sigma_b2_abs_delta = abs(fit_ds$sigma_b2 - ref_sigma_b2),
    audit = audit_returned_object(fit_ds, nrow(pooled),
                                  length(unique(pooled$cluster_id))),
    diagnostics = list(iterations = fit_ds$iterations,
                       phi = fit_ds$phi,
                       tau = fit_ds$tau,
                       trace = fit_ds$trace))
}
```

```{r fit-k2}
res_k2 <- fit_one(2L)
res_k2[c("K", "n_matched", "runtime_s", "max_abs_delta",
         "sigma_b2_abs_delta")]
res_k2$audit
```

```{r fit-k3}
res_k3 <- fit_one(3L)
res_k3[c("K", "n_matched", "runtime_s", "max_abs_delta",
         "sigma_b2_abs_delta")]
res_k3$audit
```

## 5. Comparison And Verdict

```{r compare}
coef_table <- rbind(
  cbind(K = "2", res_k2$coefficients),
  cbind(K = "3", res_k3$coefficients)
)
coef_table

summary_table <- data.frame(
  K = c(2L, 3L),
  n = c(res_k2$n, res_k3$n),
  n_matched = c(res_k2$n_matched, res_k3$n_matched),
  max_abs_beta_delta = c(res_k2$max_abs_delta, res_k3$max_abs_delta),
  sigma_b2_ds = c(res_k2$sigma_b2_ds, res_k3$sigma_b2_ds),
  sigma_b2_ref = c(res_k2$sigma_b2_ref, res_k3$sigma_b2_ref),
  sigma_b2_abs_delta = c(res_k2$sigma_b2_abs_delta,
                         res_k3$sigma_b2_abs_delta),
  returned_object_audit = c(res_k2$audit$pass, res_k3$audit$pass),
  runtime_s = c(res_k2$runtime_s, res_k3$runtime_s)
)
summary_table
```

```{r scatter, fig.cap = "Federated GLMM fixed-effect coefficients vs centralised lme4::glmer fit. Dashed identity line."}
ggplot(coef_table, aes(glmer, dsVert, shape = K)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[glmer]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method = "glmm",
  method_variant = "pql_aggregate",
  dataset = "synthetic clustered binomial (G=10, m=5)",
  n = nrow(pooled),
  K = c(2L, 3L),
  summary = summary_table,
  per_coef = coef_table,
  diagnostics = list(K2 = res_k2$diagnostics, K3 = res_k3$diagnostics),
  audits = list(K2 = res_k2$audit, K3 = res_k3$audit),
  timestamp = Sys.time(),
  dsVert_version = utils::packageVersion("dsVertClient"))
results_dir <- "results"
if (!dir.exists(results_dir) && dir.exists("vignettes/methods/results")) {
  results_dir <- "vignettes/methods/results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "glmm.rds"))
```

The audited aggregate-PQL route matches the centralized `glmer`
singular fixture with maximum fixed-effect gaps of
`r sprintf("%.3e", max(summary_table$max_abs_beta_delta))` and
random-effect variance gaps of
`r sprintf("%.3e", max(summary_table$sigma_b2_abs_delta))` for both
K modes. The returned-object audit passes in K = 2 and K = 3; no
patient-length atomic vector is visible to the analyst.

### References

* Breslow, N. E. & Clayton, D. G. (1993). Approximate inference in
  generalized linear mixed models. *JASA* 88(421), 9-25.
* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* section 3.3.
* Pinheiro, J. C. & Bates, D. M. (2000). *Mixed-Effects Models in
  S/S-PLUS*, section 7.4.
