---
title: "Inverse-probability weighting — federated two-stage propensity GLM (K = 3)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Inverse-probability weighting — federated two-stage propensity GLM (K = 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

Inverse-probability weighting (Robins, Hernán & Brumback, 2000) is a
two-stage estimator of an average treatment effect (ATE) under the
no-unmeasured-confounder assumption. Stage 1 fits a propensity model
\(\Pr(T = 1 \mid X) = \pi(X; \alpha)\) — typically a logistic GLM —
and computes per-patient weights
\(w_i = T_i / \pi_i + (1 - T_i)/(1 - \pi_i)\). Stage 2 fits the
outcome model \(\mathbb{E}[Y \mid T, X]\) by *weighted* GLM on the
same patients, recovering the marginal causal coefficient on \(T\).

In the K = 3 federated setting both stages reuse the
`ds.vertGLM` Beaver pipeline; the propensity weights are
pre-computed server-side and stored in a column on the outcome
server, then surfaced to the weighted Stage 2 fit via the
`weights` argument. `ds.vertIPW` orchestrates both stages and
returns paired propensity / outcome fits.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-patient propensity \(\pi_i\) is never reconstructed in plaintext at any single non-label party. |
| **D-INV-3** | Per-patient IPW weights live exclusively on the outcome server (where the treatment label already lives). |
| **D-INV-4** | Weighted Stage 2 score equations expose only aggregate \(X^{\top}W(y - \mu)\) blocks. |
| **D-INV-5** | All cross-server messages signed Ed25519, peer pinned via `dsvert.trusted_peers`. |

### Dataset and partition

Synthetic randomised-confounded design with known ATE for honest
verdict assessment. \(n = 300\) patients with two confounders
\((X_1, X_2)\), a logistic propensity \(\pi(X)\), a binary treatment
\(T \sim \mathrm{Bern}(\pi)\), and a binary outcome \(Y\) generated
from a treated/control conditional logit with true treatment effect
\(\beta_T = 0.6\) on the log-odds scale.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `x1` | Confounder block A |
| **s2** | `patient_id`, `x2` | Confounder block B |
| **s3** | `patient_id`, `treat`, `y`, `ipw` | Treatment + outcome + IPW weights |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(ggplot2)
```

```{r data}
n <- 300L
x1 <- rnorm(n); x2 <- rnorm(n)
prop_lp <- 0.3 * x1 - 0.2 * x2
prop_p  <- plogis(prop_lp)
treat   <- rbinom(n, 1L, prop_p)
out_lp  <- -0.3 + 0.6 * treat + 0.4 * x1 - 0.2 * x2
y       <- rbinom(n, 1L, plogis(out_lp))
fit_prop_central <- glm(treat ~ x1 + x2, family = binomial())
phat   <- fitted(fit_prop_central)
ipw    <- ifelse(treat == 1L, 1 / phat, 1 / (1 - phat))
patient_id <- sprintf("R%03d", seq_len(n))
```

```{r pooled}
pooled <- data.frame(
  patient_id = patient_id,
  x1 = x1, x2 = x2,
  treat = treat, y = y, ipw = ipw)
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "x1")]
s2 <- pooled[, c("patient_id", "x2")]
s3 <- pooled[, c("patient_id", "treat", "y", "ipw")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated IPW two-stage

```{r ds-ipw}
fit_ds <- ds.vertIPW(
  outcome_formula    = y ~ treat + x1 + x2,
  propensity_formula = treat ~ x1 + x2,
  data = "DA", weights_column = "ipw",
  outcome_family = "binomial",
  verbose = FALSE, datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_prop_ref <- glm(treat ~ x1 + x2, data = pooled, family = binomial())
fit_out_ref  <- glm(y ~ treat + x1 + x2, data = pooled,
                     family = binomial(), weights = pooled$ipw)
summary(fit_out_ref)
```

## 5. Comparison and verdict

```{r compare}
nm_p <- intersect(names(fit_ds$propensity$coefficients),
                   names(coef(fit_prop_ref)))
nm_o <- intersect(names(fit_ds$outcome$coefficients),
                   names(coef(fit_out_ref)))

prop_delta <- data.frame(
  coef    = nm_p,
  dsVert  = as.numeric(fit_ds$propensity$coefficients[nm_p]),
  glm     = as.numeric(coef(fit_prop_ref)[nm_p]))
prop_delta$abs <- abs(prop_delta$dsVert - prop_delta$glm)

out_delta <- data.frame(
  coef    = nm_o,
  dsVert  = as.numeric(fit_ds$outcome$coefficients[nm_o]),
  glm     = as.numeric(coef(fit_out_ref)[nm_o]))
out_delta$abs <- abs(out_delta$dsVert - out_delta$glm)

prop_delta
out_delta

max_abs    <- max(c(prop_delta$abs, out_delta$abs))
sigma_min  <- min(sqrt(diag(vcov(fit_out_ref))))
sigma_ratio <- sigma_min / max_abs

c(max_abs_delta = max_abs,
  sigma_min     = sigma_min,
  sigma_ratio   = sigma_ratio)
```

```{r scatter, fig.cap = "Federated outcome-stage coefficients vs centralised weighted glm() reference. Identity line dashed."}
ggplot(out_delta, aes(glm, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)["weighted glm"]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "ipw",
  dataset         = "Synthetic IPW (n=300, true ATE = 0.6)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef_propensity = prop_delta,
  per_coef_outcome    = out_delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "ipw.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The federated two-stage IPW reproduces both the propensity
\(\hat\alpha\) and outcome \(\hat\beta\) of the centralised
weighted-`glm` reference to a maximum absolute deviation of
`r sprintf("%.2e", max_abs)`, giving a σ-ratio of
`r sprintf("%.0f", sigma_ratio)`× against the smallest centralised
Wald standard error.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Lunceford, J. K. & Davidian, M. (2004). Stratification and
  weighting via the propensity score in estimation of causal
  treatment effects. *Statistics in Medicine* 23(19).
* Robins, J. M., Hernán, M. A. & Brumback, B. (2000). Marginal
  structural models and causal inference in epidemiology.
  *Epidemiology* 11(5), 550–560.
