---
title: "LASSO — federated proximal-gradient handoff over the Gaussian GLM normal-equations (K = 3)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{LASSO — federated proximal-gradient handoff over the Gaussian GLM normal-equations (K = 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

The LASSO (Tibshirani, 1996) regularises a Gaussian regression by
adding an \(\ell_1\) penalty to the residual sum of squares:

\[
\hat\beta(\lambda) \;=\; \arg\min_{\beta}
  \frac{1}{2n}\,\bigl\lVert y - X\beta\bigr\rVert_2^2
  \;+\; \lambda \,\bigl\lVert\beta\bigr\rVert_1.
\]

Beck & Teboulle (2009) show the proximal-gradient (FISTA-accelerated)
fixed-point iteration converges linearly:

\[
\beta_{t+1} \;=\;
  \mathcal{S}_{\lambda / L}\!\bigl(
    \beta_t - \tfrac{1}{L}\,\nabla\!\bigl(\tfrac{1}{2n}\lVert y - X\beta\rVert_2^2\bigr)\bigr),
\]

where \(\mathcal{S}_t(z)\) is the soft-threshold operator and \(L\)
is an upper bound on the Hessian spectrum. The gradient depends on
the centralised normal-equations \(X^{\top}X / n\) and \(X^{\top}y/n\)
which `ds.vertGLM(family = "gaussian")` already exposes via
\(\$\text{covariance}\) and \(\$\text{coefficients}\).

`ds.vertLASSOProximal` therefore performs no additional MPC rounds
beyond the initial federated Gaussian fit: the proximal-gradient
iteration runs entirely on the client using only summary statistics
the federated fit already returned. At \(\lambda = 0\) the wrapper
reproduces the prime fit's coefficients exactly — a fast handoff
sanity check.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-patient \(x_i^{\top}\beta\) never reconstructed in plaintext. |
| **D-INV-3** | The proximal-gradient client iteration consumes only the federated normal-equations summary, not raw data. |
| **D-INV-4** | At the non-label servers, partial linear-predictor shares only. |
| **D-INV-5** | Cross-server messages signed Ed25519, peer pinned via `dsvert.trusted_peers`. |

### Dataset and partition

The Hitters dataset (James, Witten, Hastie & Tibshirani, *An
Introduction to Statistical Learning*, §6.6) — Major League
Baseball player career statistics with the response `Salary`
(log-transformed). Complete cases give \(n = 263\), \(p = 19\)
covariates. We retain a numeric subset for compactness.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `AtBat`, `Hits`, `HmRun`, `Runs`, `RBI`, `Walks` | Career batting stats |
| **s2** | `patient_id`, `Years`, `CAtBat`, `CHits`, `CHmRun`, `CRuns`, `CRBI`, `CWalks` | Career totals |
| **s3** | `patient_id`, `PutOuts`, `Assists`, `Errors`, `LogSalary` | Defence + outcome (label) |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(ggplot2)
```

```{r data}
if (requireNamespace("ISLR", quietly = TRUE)) {
  data("Hitters", package = "ISLR")
} else {
  ## Synthetic Gaussian DGP with the same dimensions if ISLR
  ## is unavailable on the host.
  Hitters <- data.frame(matrix(rnorm(263 * 16), 263, 16))
  names(Hitters) <- c("AtBat","Hits","HmRun","Runs","RBI","Walks",
                       "Years","CAtBat","CHits","CHmRun","CRuns",
                       "CRBI","CWalks","PutOuts","Assists","Errors")
  Hitters$Salary <- 1 + Hitters$Hits + 0.5 * Hitters$Runs +
                     0.3 * Hitters$Walks + rnorm(263, 0, 0.5)
}
hitters <- na.omit(Hitters)
hitters$LogSalary  <- log(hitters$Salary)
hitters$patient_id <- sprintf("R%03d", seq_len(nrow(hitters)))
```

```{r pooled}
keep <- c("patient_id",
          "AtBat","Hits","HmRun","Runs","RBI","Walks",
          "Years","CAtBat","CHits","CHmRun","CRuns","CRBI","CWalks",
          "PutOuts","Assists","Errors",
          "LogSalary")
pooled <- hitters[, keep]
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id",
                 "AtBat","Hits","HmRun","Runs","RBI","Walks")]
s2 <- pooled[, c("patient_id",
                 "Years","CAtBat","CHits","CHmRun","CRuns","CRBI","CWalks")]
s3 <- pooled[, c("patient_id",
                 "PutOuts","Assists","Errors","LogSalary")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated proximal-LASSO at OLS limit (\(\lambda = 0\))

```{r ds-lasso}
fit_gauss <- ds.vertGLM(
  formula = LogSalary ~ AtBat + Hits + HmRun + Runs + RBI + Walks +
                       Years + CAtBat + CHits + CHmRun + CRuns +
                       CRBI + CWalks + PutOuts + Assists + Errors,
  data = "DA", family = "gaussian",
  verbose = FALSE, datasources = conns)

fit_ds <- ds.vertLASSOProximal(
  fit_gauss, lambda = 0,
  max_iter = 2000L, tol = 1e-9,
  keep_intercept = TRUE,
  accelerate = TRUE)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- lm(
  LogSalary ~ AtBat + Hits + HmRun + Runs + RBI + Walks +
              Years + CAtBat + CHits + CHmRun + CRuns +
              CRBI + CWalks + PutOuts + Assists + Errors,
  data = pooled)
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
delta_wrapper <- max(abs(as.numeric(fit_ds$coefficients) -
                          as.numeric(fit_gauss$coefficients)))

ds_b <- fit_ds$coefficients
rf_b <- coef(fit_ref)
nm <- intersect(names(ds_b), names(rf_b))
delta <- data.frame(
  coef   = nm,
  dsVert = as.numeric(ds_b[nm]),
  lm     = as.numeric(rf_b[nm]))
delta$abs <- abs(delta$dsVert - delta$lm)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(vcov(fit_ref))))
sigma_ratio <- sigma_min / max_abs

c(delta_wrapper_vs_prime = delta_wrapper,
  max_abs_delta_vs_lm    = max_abs,
  sigma_min              = sigma_min,
  sigma_ratio            = sigma_ratio)
```

```{r scatter, fig.cap = "Federated proximal-LASSO coefficients (lambda = 0, OLS limit) vs centralised lm() fit. Dashed identity line."}
ggplot(delta, aes(lm, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[lm]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "lasso",
  dataset         = "Hitters (ISLR)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  delta_wrapper_vs_prime = delta_wrapper,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "lasso.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

At \(\lambda = 0\) the proximal-gradient handoff reproduces the
prime federated Gaussian fit to within
`r sprintf("%.2e", delta_wrapper)`. Against the centralised `lm()`
reference the maximum absolute deviation is
`r sprintf("%.2e", max_abs)`, giving a σ-ratio of
`r sprintf("%.1f", sigma_ratio)`×.

### References

* Beck, A. & Teboulle, M. (2009). A fast iterative shrinkage-
  thresholding algorithm. *SIAM J. Imaging Sci.* 2(1), 183–202.
* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Friedman, J., Hastie, T. & Tibshirani, R. (2010). Regularization
  paths for generalized linear models via coordinate descent.
  *J. Statistical Software* 33(1), 1–22.
* James, G., Witten, D., Hastie, T. & Tibshirani, R. (2013).
  *An Introduction to Statistical Learning* §6.6.
* Tibshirani, R. (1996). Regression shrinkage and selection via the
  Lasso. *JRSS-B* 58(1), 267–288.
