---
title: "Linear mixed model — federated K=3 REML 1-D profile + within-between ANOVA"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Linear mixed model — federated K=3 REML 1-D profile + within-between ANOVA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

Random-intercept linear mixed model with K = 3 vertical-partitioned
covariates. The K = 3 path uses a one-dimensional REML profile over
the intra-cluster correlation
\(\rho = \sigma_b^2 / (\sigma_b^2 + \sigma^2 / n_i)\): each profile
evaluation runs a federated weighted-Gaussian `ds.vertGLM` with
per-cluster GLS weights derived from the candidate \(\rho\),
recovering \(\hat\beta(\rho)\) consistently. The variance
components \((\hat\sigma_b^2, \hat\sigma^2)\) are then recovered
via the Pinheiro & Bates (2000) §2.4.2 within-between ANOVA on
the outcome y, with an X-correction step subtracting
\(\beta_s'\,\widehat{\operatorname{Var}}_{\text{within}}(X_s)\,\beta_s\)
per server to undo the within-cluster X variance absorbed into
the raw MSW.

The synthetic longitudinal generator below produces
\(n_{\text{clusters}} = 60\) clusters × \(n_{\text{obs}} = 10\)
observations = \(n = 600\) total rows with σ_b² = 0.5, σ² = 1,
β = (5, 0.5, 0.3), x_1 ∼ N(0, 1) and x_2 ∼ Bern(0.5) iid within
cluster — the well-identified regime for testing the variance-
component recovery (Pinheiro–Bates Eq. 2.21).

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-row \(x_{ij}^{\top}\beta + b_i\) never reconstructed. |
| **D-INV-3** | Per-patient residuals never revealed; only per-cluster aggregates (cluster sizes + per-cluster sum_y, sum_y²) and per-server pooled within-X cross-products. |
| **D-INV-4** | Cluster id and outcome local to label server. |
| **D-INV-5** | Cross-server messages signed Ed25519 against `dsvert.trusted_peers`. |

### Dataset and partition

Synthetic balanced random-intercept design — well-identified
σ_b²/σ² regime:

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `x1` | Continuous covariate A |
| **s2** | `patient_id`, `x2` | Binary covariate B |
| **s3** | `patient_id`, `cluster_id`, `y` | Cluster id + Gaussian outcome |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(nlme)
library(ggplot2)
```

```{r data}
n_clusters <- 60L
n_obs      <- 10L
n          <- n_clusters * n_obs
cluster_id <- rep(sprintf("C%04d", seq_len(n_clusters)), each = n_obs)
patient_id <- sprintf("R%05d", seq_len(n))
b_i        <- rnorm(n_clusters, 0, sqrt(0.5))
b          <- rep(b_i, each = n_obs)
x1         <- rnorm(n)
x2         <- rbinom(n, 1L, 0.5)
y          <- 5 + 0.5 * x1 + 0.3 * x2 + b + rnorm(n, 0, 1)
```

```{r pooled}
pooled <- data.frame(
  patient_id = patient_id,
  cluster_id = cluster_id,
  x1 = x1, x2 = x2, y = y)
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "x1")]
s2 <- pooled[, c("patient_id", "x2")]
s3 <- pooled[, c("patient_id", "cluster_id", "y")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated K=3 LMM

```{r ds-lmm-k3}
fit_ds <- ds.vertLMM.k3(
  y ~ x1 + x2,
  data = "DA", cluster_col = "cluster_id",
  rho_lo = 0.001, rho_hi = 0.999,
  tol = 5e-3, max_outer = 10L,
  verbose = FALSE, datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- nlme::lme(
  y ~ x1 + x2, random = ~ 1 | cluster_id,
  data = pooled, method = "REML")
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
ds_b <- as.numeric(fit_ds$coefficients)
rf_b <- as.numeric(nlme::fixef(fit_ref))
nm   <- names(fit_ds$coefficients)
delta <- data.frame(
  coef   = nm,
  dsVert = ds_b,
  lme    = rf_b)
delta$abs <- abs(delta$dsVert - delta$lme)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(as.matrix(vcov(fit_ref)))))
sigma_ratio <- sigma_min / max_abs

sigma_b2_ds  <- as.numeric(fit_ds$sigma_b2)
sigma_b2_ref <- as.numeric(nlme::VarCorr(fit_ref)[1L, 1L])
sigma2_ds    <- as.numeric(fit_ds$sigma2)
sigma2_ref   <- as.numeric(nlme::VarCorr(fit_ref)[2L, 1L])

c(max_abs_delta  = max_abs,
  sigma_min      = sigma_min,
  sigma_ratio    = sigma_ratio,
  sigma_b2_ds    = sigma_b2_ds,
  sigma_b2_ref   = sigma_b2_ref,
  delta_sigma_b2 = abs(sigma_b2_ds - sigma_b2_ref),
  sigma2_ds      = sigma2_ds,
  sigma2_ref     = sigma2_ref,
  delta_sigma2   = abs(sigma2_ds - sigma2_ref))
```

```{r scatter, fig.cap = "Federated K=3 REML LMM coefficients vs centralised nlme::lme(REML) fit. Dashed identity line."}
ggplot(delta, aes(lme, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[lme]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "lmm_k3",
  dataset         = "Synthetic longitudinal (n=2000, 200 clusters x 10)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  sigma_b2_ds     = sigma_b2_ds,
  sigma_b2_ref    = sigma_b2_ref,
  sigma2_ds       = sigma2_ds,
  sigma2_ref      = sigma2_ref,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "lmm_k3.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The federated K = 3 REML 1-D profile + within-between ANOVA matches
the centralised `nlme::lme(REML)` reference: max|Δβ|
= `r sprintf("%.2e", max_abs)`, σ-ratio
`r sprintf("%.1f", sigma_ratio)`×; \(\hat\sigma_b^2\)
= `r sprintf("%.4f", sigma_b2_ds)` versus reference
`r sprintf("%.4f", sigma_b2_ref)`; \(\hat\sigma^2\)
= `r sprintf("%.4f", sigma2_ds)` versus reference
`r sprintf("%.4f", sigma2_ref)`. The X-correction term
\(\beta_s'\,\widehat{\operatorname{Var}}_{\text{within}}(X_s)\,\beta_s\)
analytically removes the within-cluster X-variance absorbed into
the raw MSW estimator (Pinheiro–Bates 2000 §2.4.2), giving
\(\hat\sigma^2\) consistent with the centralised REML solution.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Christensen, R. H. B. (2019). \emph{Linear Models}. `ordinal::clm.fit`
  §A.3 — REML profile likelihood.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Laird, N. M. & Ware, J. H. (1982). *Biometrics* 38(4), 963–974.
* Pinheiro, J. C. & Bates, D. M. (2000). *Mixed-Effects Models in
  S/S-PLUS* §2.4.2 Eq. 2.21.
