---
title: "Linear mixed model — federated K=2 closed-form REML"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Linear mixed model — federated K=2 closed-form REML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

A random-intercept linear mixed model

\[
y_{ij} \;=\; x_{ij}^{\top}\beta + b_i + \varepsilon_{ij},\qquad
b_i \sim \mathcal{N}(0, \sigma_b^2),\
\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2),
\]

is fit by restricted maximum likelihood (REML; Patterson &
Thompson, 1971). The marginal covariance is
\(V_i(\theta) = \sigma^2 I + \sigma_b^2 \boldsymbol{1}\boldsymbol{1}^{\top}\)
with \(\theta = (\sigma_b^2, \sigma^2)\); the GLS estimator
\(\hat\beta(\theta) = (X^{\top} V^{-1} X)^{-1} X^{\top} V^{-1} y\)
plus the REML profile \(\ell_R(\theta)\) jointly identify the
solution.

`ds.vertLMM` runs the K = 2 closed-form path: each party computes
its block of the cluster-aware Gram matrix \(X_s^{\top} V^{-1}\!X_s\)
and the cross-block \(X_1^{\top} V^{-1} X_2\) via Beaver vector
multiplication; the cluster id and outcome live on the label
server. The closed-form REML equations are recovered exactly up
to the Ring127 fixed-point ULP envelope.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-row \(x_{ij}^{\top}\beta + b_i\) never reconstructed. |
| **D-INV-3** | Per-patient residuals never revealed; only Gram and cross-Gram blocks. |
| **D-INV-4** | Cluster ID and outcome on label server only. |
| **D-INV-5** | Cross-server messages signed Ed25519 against `dsvert.trusted_peers`. |

### Dataset and partition

The Pothoff–Roy *Orthodont* dataset (Pinheiro & Bates, 2000 §1.4) —
27 children × 4 visits = 108 longitudinal dental measurements.
The vertical K = 2 split puts age on `s1` and the (cluster,
outcome) on `s2`.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `age`, `sex_num` | Time + sex covariates |
| **s2** | `patient_id`, `cluster_id`, `distance` | Cluster id + outcome (label) |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(nlme)
library(ggplot2)
```

```{r data}
data("Orthodont", package = "nlme")
od <- as.data.frame(Orthodont)
od$patient_id <- sprintf("R%03d", seq_len(nrow(od)))
od$cluster_id <- as.character(od$Subject)
od$sex_num    <- as.integer(od$Sex == "Male")
```

```{r pooled}
pooled <- od[, c("patient_id", "cluster_id",
                  "age", "sex_num", "distance")]
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "age", "sex_num")]
s2 <- pooled[, c("patient_id", "cluster_id", "distance")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated K=2 LMM

```{r ds-lmm}
fit_ds <- ds.vertLMM(
  formula = distance ~ age + sex_num,
  data = "DA", cluster_col = "cluster_id",
  reml = TRUE, verbose = FALSE,
  datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- nlme::lme(
  distance ~ age + sex_num,
  random = ~ 1 | cluster_id,
  data = pooled, method = "REML")
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
ds_b <- as.numeric(fit_ds$coefficients %||% fit_ds$beta)
rf_b <- as.numeric(nlme::fixef(fit_ref))
nm <- if (length(ds_b) == length(rf_b)) names(nlme::fixef(fit_ref)) else
  intersect(names(fit_ds$coefficients), names(nlme::fixef(fit_ref)))
delta <- data.frame(
  coef   = nm,
  dsVert = ds_b,
  lme    = rf_b)
delta$abs <- abs(delta$dsVert - delta$lme)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(as.matrix(vcov(fit_ref)))))
sigma_ratio <- sigma_min / max_abs

c(max_abs_delta = max_abs,
  sigma_min     = sigma_min,
  sigma_ratio   = sigma_ratio)
```

```{r scatter, fig.cap = "Federated K=2 LMM coefficients vs centralised nlme::lme(REML) fit. Dashed identity line."}
ggplot(delta, aes(lme, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[lme]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "lmm_k2",
  dataset         = "Orthodont (Pothoff-Roy)",
  n               = nrow(pooled),
  K               = 2L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "lmm_k2.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The federated K = 2 closed-form REML estimator agrees with
`nlme::lme(method = "REML")` to a maximum absolute deviation of
`r sprintf("%.2e", max_abs)`, σ-ratio
`r sprintf("%.0f", sigma_ratio)`×.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Laird, N. M. & Ware, J. H. (1982). Random-effects models for
  longitudinal data. *Biometrics* 38(4), 963–974.
* Patterson, H. D. & Thompson, R. (1971). Recovery of inter-block
  information when block sizes are unequal. *Biometrika* 58(3).
* Pinheiro, J. C. & Bates, D. M. (2000). *Mixed-Effects Models in
  S/S-PLUS* §1.4 + §2.4.
