---
title: "Multinomial warm — federated OVR + softmax-anchor (K = 3)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multinomial warm — federated OVR + softmax-anchor (K = 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

The Rifkin–Klautau (2004) one-vs-rest (OVR) framework + softmax-
anchor decomposes a K-class multinomial logistic regression into
K − 1 independent binomial GLMs (per non-reference class
indicator), then anchors them through the softmax constraint
\(\sum_k \Pr(y = k \mid x) = 1\) to recover the multinomial slope
matrix. This is the "warm path" — slower than the joint Newton
path on a single iteration but compatible with the K = 3
secure-aggregation pipeline because each binomial cell is just a
`ds.vertGLM` family-binomial call.

`ds.vertMultinom` orchestrates the OVR cells across the K = 3
vertical-partition; the softmax-anchor reconciliation runs
client-side using only the per-cell `ds.glm` summaries.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-row \(x_i^{\top}\beta_k\) never reconstructed in plaintext. |
| **D-INV-3** | Per-patient one-hot class indicator lives only on the label server. |
| **D-INV-4** | Each OVR cell exchanges only aggregate score / Fisher scalars. |
| **D-INV-5** | Cross-server messages signed Ed25519 against `dsvert.trusted_peers`. |

### Dataset and partition

Same `iris` dataset and species target as `mnl_joint`, but with
the four covariates split across THREE servers.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `Sepal_Length`, `Sepal_Width` | Sepal block |
| **s2** | `patient_id`, `Petal_Length`, `Petal_Width` | Petal block |
| **s3** | `patient_id`, `species`, `setosa_ind`, `versicolor_ind`, `virginica_ind` | Outcome (label) |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(nnet)
library(ggplot2)
```

```{r data}
data("iris", package = "datasets")
ir <- iris
ir$patient_id    <- sprintf("R%03d", seq_len(nrow(ir)))
ir$species       <- factor(ir$Species,
                            levels = c("setosa", "versicolor",
                                        "virginica"))
ir$setosa_ind    <- as.integer(ir$species == "setosa")
ir$versicolor_ind <- as.integer(ir$species == "versicolor")
ir$virginica_ind <- as.integer(ir$species == "virginica")
ir$Sepal_Length  <- ir$Sepal.Length
ir$Sepal_Width   <- ir$Sepal.Width
ir$Petal_Length  <- ir$Petal.Length
ir$Petal_Width   <- ir$Petal.Width
```

```{r pooled}
pooled <- ir[, c("patient_id",
                 "Sepal_Length", "Sepal_Width",
                 "Petal_Length", "Petal_Width",
                 "species",
                 "setosa_ind", "versicolor_ind", "virginica_ind")]
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "Sepal_Length", "Sepal_Width")]
s2 <- pooled[, c("patient_id", "Petal_Length", "Petal_Width")]
s3 <- pooled[, c("patient_id", "species",
                  "setosa_ind", "versicolor_ind", "virginica_ind")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated K=3 multinomial warm

```{r ds-mnl-warm}
fit_ds <- ds.vertMultinom(
  species ~ Sepal_Length + Sepal_Width + Petal_Length + Petal_Width,
  data = "DA",
  classes = c("setosa", "versicolor", "virginica"),
  reference = "setosa",
  verbose = FALSE, datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- nnet::multinom(
  species ~ Sepal_Length + Sepal_Width + Petal_Length + Petal_Width,
  data = pooled, trace = FALSE)
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
ds_B <- as.matrix(fit_ds$coefficients %||% fit_ds$beta)
rf_B <- t(as.matrix(coef(fit_ref)))
cn   <- intersect(rownames(ds_B), rownames(rf_B))
cls  <- intersect(colnames(ds_B), colnames(rf_B))
delta <- data.frame(
  coef     = c(t(outer(cn, cls, paste, sep = ":"))),
  dsVert   = as.numeric(ds_B[cn, cls, drop = FALSE]),
  multinom = as.numeric(rf_B[cn, cls, drop = FALSE]))
delta$abs <- abs(delta$dsVert - delta$multinom)
delta

max_abs    <- max(delta$abs)
sigma_min  <- NA_real_
sigma_ratio <- NA_real_

c(max_abs_delta = max_abs)
```

```{r scatter, fig.cap = "Federated K=3 warm-path multinomial coefficients vs centralised nnet::multinom fit. Dashed identity line."}
ggplot(delta, aes(multinom, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[multinom]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "mnl_warm",
  dataset         = "iris (Fisher/Anderson)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "mnl_warm.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The K = 3 warm-path multinomial estimator agrees with
`nnet::multinom` to a maximum absolute deviation of
`r sprintf("%.2e", max_abs)` across the
\((K - 1) \times p\) coefficient matrix.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* Krishnapuram, B. *et al.* (2005). *IEEE PAMI* 27(6) §3.2.
* Rifkin, R. & Klautau, A. (2004). *JMLR* 5, 101–141.
