---
title: "Multinomial logistic -- strict joint Newton evidence (K = 2 and K >= 3)"
author: "David Sarrat Gonzalez, Miron Banjac, Juan R. Gonzalez"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multinomial logistic -- strict joint Newton evidence (K = 2 and K >= 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
```

## 1. Context

`ds.vertMultinomJointNewton` fits a true joint softmax model rather
than the older one-vs-rest warm approximation. The strict route uses
Ring127 share-domain softmax probabilities and residuals, with a
Bohning majorization step based on low-dimensional aggregate Gram
information. K >= 3 uses outcome and fusion DCF parties; non-DCF
feature servers send encrypted additive shares only.

The full DSLite validation is intentionally not re-run during vignette
rendering because each K mode takes about 14 minutes on the deterministic
n = 50 fixture. This vignette loads the checked DSLite evidence generated
by the method harness and records the exact commands needed to reproduce
the caches.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-patient softmax probabilities and residuals remain Ring127 additive shares. |
| **D-INV-3** | The client receives aggregate gradients, the aggregate Bohning Gram and fitted coefficients only. |
| **D-INV-4** | K >= 3 non-DCF servers do not receive labels, probabilities, residuals or fitted values. |
| **D-INV-5** | Returned-object audit rejects numeric vectors of patient length. |

## 2. Reproduction Commands

```{r commands, eval = FALSE}
# From the dsvert-paper repository root.
Rscript scripts/validate_method_multinom_joint.R \
  --K 2 --n 50 --max-outer 30 --tol 1e-4 --privacy-level 5

Rscript scripts/validate_method_multinom_joint.R \
  --K 3 --n 50 --max-outer 30 --tol 1e-4 --privacy-level 5
```

## 3. Cached DSLite Evidence

```{r libs}
library(ggplot2)
```

```{r load-results}
result <- readRDS("results/mnl_joint.rds")
result$summary
```

```{r audit}
result$audits
```

## 4. Accuracy

```{r accuracy}
summary_table <- result$summary
summary_table[, c("K", "n", "max_abs_prob_delta",
                  "warm_max_abs_prob_delta", "joint_improvement",
                  "max_abs_coef_delta", "runtime_s",
                  "returned_object_audit")]
```

```{r plot, fig.cap = "Joint softmax probability error versus warm OVR baseline."}
plot_df <- data.frame(
  K = factor(summary_table$K),
  strict = summary_table$max_abs_prob_delta,
  warm = summary_table$warm_max_abs_prob_delta
)
plot_df <- reshape(
  plot_df, varying = c("strict", "warm"),
  v.names = "max_abs_prob_delta",
  timevar = "route", times = c("strict_joint", "warm_ovr"),
  direction = "long")
ggplot(plot_df, aes(K, max_abs_prob_delta, fill = route)) +
  geom_col(position = "dodge") +
  scale_y_log10() +
  labs(x = "K mode", y = "max absolute probability delta") +
  theme_minimal(base_size = 11)
```

The strict joint route reaches maximum class-probability gaps of
`r sprintf("%.3e", max(summary_table$max_abs_prob_delta))` across
K = 2 and K >= 3, while the warm OVR route is around
`r sprintf("%.3e", max(summary_table$warm_max_abs_prob_delta))` on
the same fixture. Both returned-object audits pass.

### References

* Bohning, D. (1992). Multinomial logistic regression algorithm.
  *Annals of the Institute of Statistical Mathematics* 44(1).
* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* section 3.3.
* Nocedal, J. & Wright, S. J. (2006). *Numerical Optimization* section 3.5.
