---
title: "Ordinal warm — federated cumulative-binomial OVR (K = 3)"
author: "David Sarrat González, Miron Banjac, Juan R. González"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal warm — federated cumulative-binomial OVR (K = 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  warning = FALSE, message = FALSE,
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 5,
  fig.align = "center")
set.seed(2026L)
```

## 1. Context

The "warm" ordinal path (Rifkin & Klautau, 2004; McCullagh, 1980
§2.5) decomposes a K-class proportional-odds model into K − 1
independent cumulative-binomial fits — each row of the model
matrix predicts \(\Pr(y \le k \mid x)\) for one threshold k. With
K classes, the warm path runs K − 1 separate `ds.vertGLM`
binomial cells; the slope vector is recovered as the precision-
weighted average of the per-threshold slopes (proportional-odds
constraint).

Compared to the joint Newton path (`ord_joint`), the warm path
loses the closed-form cross-block Hessian \(H_{\beta\theta}\) but
gains compatibility with the K = 3 secure-aggregation pipeline:
each cumulative-binomial cell is just a `ds.vertGLM` binomial
call on a 3-server vertical split, no joint Newton block-diagonal
solver required.

### Non-disclosure invariants

| Invariant | Statement |
|---:|:---|
| **D-INV-1** | Per-patient covariates never leave their origin server. |
| **D-INV-2** | Per-row \(x_i^{\top}\beta\) never reconstructed in plaintext. |
| **D-INV-3** | Per-patient cumulative indicators \(\mathbf{1}\{y_i \le k\}\) live only on the label server. |
| **D-INV-4** | Each warm cell exchanges only aggregate score / Fisher scalars. |
| **D-INV-5** | Cross-server messages signed Ed25519 against `dsvert.trusted_peers`. |

### Dataset and partition

The MASS `birthwt` dataset, ordinal-tertile target as in
`ord_joint`, but the design partitions across THREE servers.

| Server | Columns | Role |
|:---|:---|:---|
| **s1** | `patient_id`, `age`, `lwt` | Maternal demographics |
| **s2** | `patient_id`, `smoke`, `ht`, `ui` | Risk indicators |
| **s3** | `patient_id`, `bwt_cls`, `low_leq`, `med_leq`, `high_leq` | Ordinal outcome + cumulative indicators (label) |

## 2. Data preparation

```{r libs}
library(dsVertClient)
library(DSLite)
library(DSI)
library(MASS)
library(ggplot2)
```

```{r data}
data("birthwt", package = "MASS")
bw <- birthwt
bw$patient_id <- sprintf("BW%03d", seq_len(nrow(bw)))
brks <- quantile(bw$bwt, c(0, 1/3, 2/3, 1), na.rm = TRUE)
bw$bwt_cls <- factor(
  cut(bw$bwt, breaks = brks, include.lowest = TRUE,
      labels = c("low", "med", "high")),
  levels = c("low", "med", "high"), ordered = TRUE)
bw$low_leq  <- as.integer(as.integer(bw$bwt_cls) <= 1L)
bw$med_leq  <- as.integer(as.integer(bw$bwt_cls) <= 2L)
bw$high_leq <- 1L
```

```{r pooled}
pooled <- bw[, c("patient_id",
                 "age", "lwt", "smoke", "ht", "ui",
                 "bwt_cls",
                 "low_leq", "med_leq", "high_leq")]
head(pooled)
```

```{r split}
s1 <- pooled[, c("patient_id", "age", "lwt")]
s2 <- pooled[, c("patient_id", "smoke", "ht", "ui")]
s3 <- pooled[, c("patient_id", "bwt_cls",
                  "low_leq", "med_leq", "high_leq")]
```

```{r head-s1}
head(s1)
```

```{r head-s2}
head(s2)
```

```{r head-s3}
head(s3)
```

```{r dslite}
dslite.server <- newDSLiteServer(
  tables = list(s1 = s1, s2 = s2, s3 = s3),
  config = DSLite::defaultDSConfiguration(
    include = c("dsBase", "dsVert")))
```

```{r login}
options(dsvert.require_trusted_peers = FALSE,
        datashield.privacyLevel = 0L)

builder <- newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "s1", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "s2", driver = "DSLiteDriver")
builder$append(server = "site3", url = "dslite.server",
               table = "s3", driver = "DSLiteDriver")
conns <- datashield.login(builder$build(),
                           assign = TRUE, symbol = "D",
                           opts = list(server = dslite.server))
```

```{r psi}
psi <- ds.psiAlign("D", "patient_id", "DA",
                    datasources = conns, verbose = FALSE)
psi[[1]]$n_matched
```

## 3. Federated K=3 ordinal warm

```{r ds-ord-warm}
fit_ds <- ds.vertOrdinal(
  bwt_cls ~ age + lwt + smoke + ht + ui,
  data = "DA",
  levels_ordered = c("low", "med", "high"),
  cumulative_template = "%s_leq",
  verbose = FALSE, datasources = conns)
fit_ds
```

## 4. Centralised reference

```{r ref}
fit_ref <- MASS::polr(
  bwt_cls ~ age + lwt + smoke + ht + ui,
  data = pooled, method = "logistic", Hess = TRUE)
summary(fit_ref)
```

## 5. Comparison and verdict

```{r compare}
ds_b <- as.numeric(fit_ds$coefficients %||% fit_ds$beta)
rf_b <- as.numeric(coef(fit_ref))
nm   <- names(coef(fit_ref))
delta <- data.frame(
  coef   = nm,
  dsVert = ds_b[seq_along(rf_b)],
  polr   = rf_b)
delta$abs <- abs(delta$dsVert - delta$polr)
delta

max_abs    <- max(delta$abs)
sigma_min  <- min(sqrt(diag(vcov(fit_ref))))
sigma_ratio <- sigma_min / max_abs

c(max_abs_delta = max_abs,
  sigma_min     = sigma_min,
  sigma_ratio   = sigma_ratio)
```

```{r scatter, fig.cap = "Federated K=3 warm-path ordinal slopes vs centralised MASS::polr fit. Dashed identity line."}
ggplot(delta, aes(polr, dsVert)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 2.5) +
  labs(x = expression(hat(beta)[polr]),
       y = expression(hat(beta)[dsVert])) +
  theme_minimal(base_size = 11)
```

```{r save-verdict, include = FALSE}
verdict <- list(
  method          = "ord_warm",
  dataset         = "MASS::birthwt (bwt-tertile ordinal)",
  n               = nrow(pooled),
  K               = 3L,
  max_abs_delta   = max_abs,
  sigma_min_wald  = sigma_min,
  sigma_ratio     = sigma_ratio,
  per_coef        = delta,
  timestamp       = Sys.time(),
  dsVert_version  = utils::packageVersion("dsVertClient"))
results_dir <- here::here("vignettes/methods/results")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
saveRDS(verdict, file.path(results_dir, "ord_warm.rds"))
```

```{r logout, include = FALSE}
datashield.logout(conns)
```

The K = 3 warm-path ordinal estimator agrees with `MASS::polr` to
a maximum absolute deviation of `r sprintf("%.2e", max_abs)`,
σ-ratio `r sprintf("%.1f", sigma_ratio)`×.

### References

* Catrina, O. & Saxena, A. (2010). *Financial Cryptography* §3.3.
* Demmler, D., Schneider, T. & Zohner, M. (2015). ABY. *NDSS* §III.B.
* McCullagh, P. (1980). *JRSS-B* 42(2), 109–142.
* Rifkin, R. & Klautau, A. (2004). In defense of one-vs-all
  classification. *JMLR* 5, 101–141.
